---
title: "A1 · Getting started — your first multisite simulation"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For researchers designing or evaluating a multisite trial who need a
  defensible synthetic dataset before real data exist. This vignette
  walks through one preset call, one `sim_multisite()` call, and the
  diagnostic readout that tells you whether the realized design hit
  its targets. You leave with a 50-site simulation tibble, a
  caterpillar and funnel plot, and a provenance string suitable for a
  grant appendix or manuscript caption.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A1 Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

Here is what you will produce in five minutes — a labeled view of 50
simulated school effects, ranked from most-shrunk on the left to
least-shrunk on the right, with each site's 95 percent interval drawn
around its observed estimate.

```{r teaser-prelude, echo = FALSE, fig.cap = "What you will produce in 5 minutes — every dot is a simulated school's effect estimate; ranked from most-shrunk to least.", fig.alt = "Caterpillar plot of 50 simulated site effects from preset_jebs_paper, with 95 percent intervals drawn around each observed estimate."}
.dat_teaser <- sim_multisite(preset_jebs_paper(), seed = 1L)
plot_effects(.dat_teaser, caption = FALSE) +
  ggplot2::labs(subtitle = "preset_jebs_paper() · seed = 1L · J = 50 sites")
```

Every dot above is a single simulated school. The vertical spread is
the heterogeneity the design encodes; the bar widths are the
sampling-error structure; the rank order is what your downstream
shrinkage estimator will see. The remainder of this vignette walks you
through producing the same plot from scratch — and reading the
diagnostics that justify trusting it.

## 1. Why you are here

Maya is writing a grant proposal for a multisite tutoring trial across
roughly 50 schools, and the reviewers want to see that her power
calculation holds up when between-site heterogeneity is around
$\sigma_\tau = 0.20$. She does not have data yet — she needs a
defensible synthetic dataset that mirrors the design she is proposing,
plus a one-line provenance string she can drop in the methods
appendix.

## 2. What you will leave with

- A 50-site simulated dataset with the canonical columns
  `site_index`, `z_j`, `tau_j`, `tau_j_hat`, `se_j`, `se2_j`, `n_j`.
- A diagnostic readout that compares realized values against the
  design's targets and flags PASS / WARN / FAIL.
- Two plots — a caterpillar and a funnel — that make the realized
  heterogeneity and precision pattern legible at a glance.
- A reproducible provenance string carrying a 16-character canonical
  hash, suitable for a grant appendix or manuscript caption.

## 3. Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("joonho112/multisiteDGP")
```

```{r load, eval = FALSE}
library(multisiteDGP)
library(ggplot2)
```

The setup chunk above already loaded both, with `set.seed(1L)`. Every
primary call below also passes `seed = 1L` so your local re-run
matches the rendered vignette exactly.

## 4. Your first simulation

Two calls. First, pick a *preset* — a small, defensible bundle of
design parameters tied to a specific scientific scenario. Maya picks
`preset_jebs_paper()` because it matches the JEBS-paper calibration
[@lee2025improving]: $J = 50$ sites, $\sigma_\tau = 0.20$,
mixture-shape latent effects, no induced precision–effect dependence,
site sizes drawn around $\bar{n}_j = 40$.

```{r design}
design <- preset_jebs_paper()
design
```

Read the printed object as four generative layers. Layer 1 (G-effects)
sets the latent shape and the between-site SD. Layer 2 (Margin)
controls site sizes and how their sampling variances are produced —
this preset uses the *site-size-driven* path (Paradigm A in the
package's underlying blueprint), where $\widehat{se}_j$ is generated
from $n_j$ rather than supplied directly. Layer 3 (Dependence) is set
to none. Layer 4 (Observation) adds sampling noise to the latent
effect. The design is immutable; to change a field, call
[`update_multisitedgp_design()`](../reference/update_multisitedgp_design.html)
or pass an override at the
[`multisitedgp_design()`](../reference/multisitedgp_design.html)
constructor.

Now simulate.

```{r simulate}
dat <- sim_multisite(design, seed = 1L)
dat
```

The print method front-loads what Maya cares about: the realized
$\sigma_\tau$ landed at 0.207 against a target of 0.200, marked PASS;
the realized rank correlation between effects and precisions came in
at $-0.193$ against a target of 0 (PASS within tolerance); and
operational feasibility is flagged WARN because the implied effective
sample size is around 13. The seven canonical columns then follow as
a tibble.

## 5. Read the output

The simulation tibble carries one row per site and seven columns:

- `site_index` — integer site identifier 1, 2, ..., J.
- `z_j` — standardized residual site effect, on the unit-variance
  scale (Layer 1).
- `tau_j` — latent site effect on the outcome scale:
  $\tau + \sigma_\tau z_j$.
- `tau_j_hat` — observed site estimate (latent effect plus sampling
  error).
- `se_j` — sampling standard error of the site estimate.
- `se2_j` — the corresponding sampling variance ($se_j^2$).
- `n_j` — generated site sample size.

The latent column `tau_j` is the *true* site effect Maya would recover
with infinite data. The observed column `tau_j_hat` is what an analyst
would actually see — it carries Layer 4 sampling noise on top of the
latent value, with magnitude `se_j`. Sites with smaller `n_j` carry
larger `se_j`, so noisy estimates and small samples line up the way a
real trial would produce them. The latent and the observed estimate
live side by side because that is the comparison every shrinkage
estimator is trying to make [@walters2024empirical].

A peek at the first few rows confirms the contract:

```{r columns-peek}
head(tibble::as_tibble(dat)[, c("site_index", "tau_j", "tau_j_hat", "se_j", "n_j")], 6)
```

Site 1 has a latent effect near $-0.116$ but the observed estimate is
$0.652$. That wide gap, which appears across many sites, is exactly
what makes shrinkage a non-trivial inferential problem. Site 3, with
$n_j = 61$, has a noticeably smaller standard error than site 2, with
$n_j = 12$ — the variation in `se_j` you see here is the
sampling-variance heterogeneity the diagnostic table will summarize as
$R$.

## 6. Read the diagnostics

`summary()` prints the full target-vs-realized diagnostic report.
This is the table you would screenshot for a methods appendix — it
answers whether the simulation actually delivered what the design
promised.

```{r diagnostics}
summary(dat)
```

What to look at, in order:

- **`sigma_tau` (Group A, Realized vs Intended)** — the realized
  between-site SD. Target 0.200, realized 0.207, status PASS. This is
  the headline number for any heterogeneity-driven power calculation.
- **`I` (informativeness, Group A)** — 0.250. The proportion of total
  variance that is signal rather than sampling noise; lower values
  mean the design is dominated by noise.
- **`R` (SE heterogeneity, Group A)** — 7.583. The ratio of largest
  to smallest sampling variance across sites.
- **`rank_corr residual` (Group B, Dependence)** — $-0.193$ against a
  target of 0, PASS. A non-zero realized value at $J = 50$ is sampling
  noise around the design target, not a violation of it.
- **`pearson_corr residual` (Group B)** — $-0.231$, FAIL against a
  target of 0. Rank-based diagnostics are the primary contract for the
  site-size-driven path; the Pearson flag is informational here, not a
  blocker. See [Diagnostics in practice](a3-diagnostics-in-practice.html)
  for the full rubric.
- **`feasibility_index` (Group D, Operational feasibility)** — 13.1,
  WARN. This is the effective per-site sample size after shrinkage;
  under 20 means partial pooling will move estimates substantially.
  Maya flags this as expected for the design and notes it in her power
  discussion.

The Group C entries (KS distance, Bhattacharyya, Q-Q residual) print
as N/A because the JEBS preset does not declare a target G shape to
compare against — these become live in
[Diagnostics in practice](a3-diagnostics-in-practice.html) and in
[Calibrating to real data](a5-calibrating-real-data.html).

## 7. Plot the results

Two plots make the realized design legible — the caterpillar reads
across sites, the funnel reads precision against effect.

```{r caterpillar, fig.cap = "Caterpillar plot. Black points are observed estimates with 95 percent CI bars; the solid horizontal line marks the grand mean. Sites are sorted by absolute estimate. Vertical spread of the dots is your realized between-site heterogeneity (sigma_tau hat = 0.207); bar-width variation across sites is the sampling-error heterogeneity (R = 7.6) set by site-size variation.", fig.alt = "Caterpillar plot of 50 simulated site effects from preset_jebs_paper, ordered left-to-right by absolute estimate, with 95 percent intervals around each observed estimate."}
plot_effects(dat, caption = FALSE) +
  ggplot2::labs(
    title    = "Site effects under preset_jebs_paper()",
    subtitle = "J = 50 sites · seed = 1L · sigma_tau hat = 0.207"
  )
```

Three things to check on the caterpillar:

1. **Vertical spread of the dots.** That is your realized site-level
   heterogeneity. Confirm it numerically against the `sigma_tau` line
   in the diagnostic table.
2. **Bar widths across sites.** Wide bars are small sites; narrow bars
   are large sites. The ratio of widest to narrowest is `R = 7.6` —
   meaningful but not pathological for site-size-driven designs.
3. **Rank order across the x-axis.** This is exactly what an empirical
   Bayes estimator will shrink, and the magnitude of shrinkage tracks
   how much each site's bar narrows in your downstream analysis.

```{r funnel, fig.cap = "Funnel plot. Standard error on the x-axis (smaller = more precise); observed estimate on the y-axis. Cloud width tapers as SE shrinks, the expected signature when SEs come from site sizes. Roughly symmetric scatter top-to-bottom confirms the rank-correlation target rho_S = 0 (PASS) — small-SE sites are not systematically higher- or lower-effect than large-SE sites.", fig.alt = "Funnel plot of 50 simulated site effects against their standard errors."}
plot_funnel(dat, caption = FALSE) +
  ggplot2::labs(
    title    = "Precision vs. effect: funnel view",
    subtitle = "Symmetric scatter confirms target rho_S = 0"
  )
```

What to read off the funnel: a *biased* design would tilt the cloud —
small-SE sites would skew toward one side of zero. This funnel is
roughly symmetric, the visual signature of the PASS verdict on
`rank_corr residual`. If Maya later modifies the design to induce
precision–effect dependence (Layer 3), the cloud will tilt visibly,
and the diagnostic flag will flip.

## 8. Adapt for downstream analysis

Three adapter functions hand the simulation off to specific downstream
packages without renaming columns by hand:
[`as_metafor()`](../reference/as_metafor.html),
[`as_baggr()`](../reference/as_baggr.html), and
[`as_multisitepower()`](../reference/as_multisitepower.html).

```{r adapter}
mf <- as_metafor(dat)
head(mf, 4)
```

After `as_metafor()` the columns are `yi` (effect estimate), `vi`
(sampling variance), and `sei` (sampling SE) — the convention
`metafor::rma()` expects. Sibling adapters return the column
conventions of the `baggr` and `multisitepower` packages. The recipe
gallery at [Cookbook](a8-cookbook.html) walks each adapter end-to-end
with a fitted model; [Adapters and downstream packages](m6-adapters-downstream.html)
gives the formal rename contract.

## 9. Provenance

A defensible simulation needs a one-line fingerprint Maya can paste
into the methods appendix. `provenance_string()` returns it.

```{r provenance}
provenance_string(dat)
```

The string carries the package version, the paradigm, the seed, the
canonical hash of the simulation output, the design hash, the hash
algorithm, the R version, and any active hooks. The canonical hash
itself can be pulled directly:

```{r canonical-hash}
canonical_hash(dat)
```

If a coauthor reruns `sim_multisite(preset_jebs_paper(), seed = 1L)`
on the same package version, that 16-character hash is the contract:
identical hash means byte-identical simulation. A different hash flags
a drift in the design, the seed, or the package — exactly what a
methods reviewer wants to be sure cannot happen silently. You can
verify this directly:

```{r determinism}
dat_repeat <- sim_multisite(design, seed = 1L)
identical(canonical_hash(dat), canonical_hash(dat_repeat))
```

See [Reproducibility and provenance](m7-reproducibility-provenance.html)
for how to wire the hash and provenance string into automated audit
trails.

## 10. Where to next

- [Choosing a preset](a2-choosing-a-preset.html) — the nine built-in
  presets and when each fits your design.
- [Diagnostics in practice](a3-diagnostics-in-practice.html) — the
  four diagnostic groups (A: scale, B: dependence, C: G-shape fit,
  D: feasibility) and the plots that go with each.
- [Calibrating to real data](a5-calibrating-real-data.html) — when
  you have observed point estimates and SEs you want the simulator to
  match.
- [`sim_multisite()`](../reference/sim_multisite.html) — full argument
  reference, including how to override preset fields.
- [`preset_jebs_paper()`](../reference/preset_jebs_paper.html) — the
  preset used here.

For the methodological foundations — the formal two-stage DGP, the
G-distribution catalog, the *site-size-driven* and *direct-precision*
paths — start with [The two-stage DGP](m1-statistical-dgp.html) and
proceed through M2 / M3 / M4 in order.

## References

The `multisiteDGP` data-generating process and the JEBS preset
implement the design in @lee2025improving. Empirical-Bayes context
for site-effect estimation is collected in @walters2024empirical;
applied calibration benchmarks for the education presets are taken
from @weiss2017howmuch.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
