---
title: "A2 · Choosing a preset"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For researchers who know they need a multisite or meta-analytic
  simulation but do not yet know which preset matches their study.
  This vignette is a scannable decision table over the nine built-in
  presets — read down the rows, find your scenario, lift the call.
  You leave with the preset name that fits, the source citation behind
  it, the override pattern for adapting it to a local design, and a
  provenance string suitable for a methods appendix.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A2 Choosing a preset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The problem with picking parameters by feel

A multisite simulation has roughly a dozen knobs that interact: $J$,
$\sigma_\tau$, $\bar n_j$, the latent G shape, the dependence target,
the engine. Pick them one at a time and the realized design rarely
matches what you intended — the implied informativeness $I$ is too
high, the heterogeneity ratio $R$ is too narrow, the shrinkage flag
fires unexpectedly. A *preset* packages a defensible combination tied
to a published reference or a documented benchmark. You start there,
then override the one or two fields your study departs on.

The package ships nine presets across two front doors. Seven are
*site-size-driven* (sampling variance comes from generated $n_j$, used
with [`sim_multisite()`](../reference/sim_multisite.html)); two are
*direct-precision* (sampling variance is set through $(I, R)$, used
with [`sim_meta()`](../reference/sim_meta.html)). The rest of this
vignette is a table you read down to find the row that matches your
scenario, plus the override pattern for adapting it.

## 2. What a preset locks for you

Every preset returns a `multisitedgp_design` with all four generative
layers already filled in. The constructor below shows what the nine
presets fix at the parameters that drive the headline diagnostics —
the paradigm and engine, the site count $J$, the between-site SD
$\sigma_\tau$, the latent G shape, the site-size mean (the
*site-size-driven* path, Paradigm A in the package's underlying
blueprint) or the $(I, R)$ pair (the *direct-precision* path, Paradigm
B in the same blueprint), and the citation that anchors the choice.

```{r preset-table}
preset_objects <- list(
  education_small        = preset_education_small(),
  education_modest       = preset_education_modest(),
  education_substantial  = preset_education_substantial(),
  jebs_paper             = preset_jebs_paper(),
  jebs_strict            = preset_jebs_strict(),
  walters_2024           = preset_walters_2024(),
  twin_towers            = preset_twin_towers(),
  meta_modest            = preset_meta_modest(),
  small_area_estimation  = preset_small_area_estimation()
)

preset_summary <- data.frame(
  preset      = names(preset_objects),
  paradigm    = vapply(preset_objects, `[[`, character(1), "paradigm"),
  engine      = vapply(preset_objects, `[[`, character(1), "engine"),
  J           = vapply(preset_objects, `[[`, integer(1),   "J"),
  sigma_tau   = vapply(preset_objects, `[[`, numeric(1),   "sigma_tau"),
  true_dist   = vapply(preset_objects, `[[`, character(1), "true_dist"),
  nj_mean     = vapply(preset_objects, function(p) p$nj_mean %||% NA_real_, numeric(1)),
  I           = vapply(preset_objects, function(p) p$I       %||% NA_real_, numeric(1)),
  R           = vapply(preset_objects, function(p) p$R       %||% NA_real_, numeric(1)),
  row.names   = NULL,
  stringsAsFactors = FALSE
)
preset_summary
```

Read across a row as the locked design. The two direct-precision
presets at the bottom carry the active $(I, R)$ pair that drives
their sampling-variance distribution; their `nj_mean` field surfaces
the constructor default and is not used at simulation time. The seven
site-size presets above them carry the active `nj_mean` (and the
implicit `cv`, `nj_min`, `R2` settings printed by the design object
itself); their $(I, R)$ entries surface constructor defaults and are
not used — realized $I$ and $R$ fall out of the site-size distribution
and the chosen $\sigma_\tau$ at simulation time.

The two engines you see — `A1_legacy` and `A2_modern` — produce
identical numerical results on the standardized residual scale; the
difference is internal plumbing. The two JEBS presets pin
`A1_legacy` because the published JEBS results were generated under
that engine and the parity check on the JEBS table requires it.
Everything else uses `A2_modern`. See
[Margin and SE models](m3-margin-se-models.html) for the engine
discussion.

## 3. The decision rubric

Read this table down, top to bottom, until a row describes your
study. The recommended preset is the call to lift; the locked
parameters column is what that preset will fix for you; the source
column is the published reference behind the calibration; the
override column is the field most studies need to change first.

| Scenario | Preset | Locked: J / sigma_tau / paradigm | Source | Most common override |
|---|---|---|---|---|
| Education trial, small effects (~0.05) on the standardized scale | `preset_education_small()` | J = 50 / 0.05 / site-size | @weiss2017howmuch | `J`, `nj_mean` |
| Education trial, modest effects (~0.20) — typical RCT scale | `preset_education_modest()` | J = 50 / 0.20 / site-size | @weiss2017howmuch | `J`, `sigma_tau` |
| Education trial, substantial effects (~0.30), large J | `preset_education_substantial()` | J = 100 / 0.30 / site-size | @weiss2017howmuch | `nj_mean` |
| Reproduce the JEBS paper design (mixture-shape latent effects) | `preset_jebs_paper()` | J = 50 / 0.20 / site-size | @lee2025improving | (none — verbatim) |
| JEBS strict parity grid (validation / benchmarking) | `preset_jebs_strict()` | J = 100 / 0.15 / site-size | @lee2025improving | (none — verbatim) |
| Replicate the Walters Handbook chapter calibration | `preset_walters_2024()` | J = 46 / 0.197 / site-size; locks $R^2 = 0.4$, $\bar n_j = 240$ | @walters2024empirical | `R2` |
| Large-J / extreme heterogeneity stress test | `preset_twin_towers()` | J = 1000 / 2.0 / site-size | Package-curated | `J` |
| Meta-analysis warm-up (direct-precision path) | `preset_meta_modest()` | J = 50 / 0.20 / direct; locks $I = 0.30$, $R = 1.5$ | Package-curated | `I`, `R` |
| Small-area-estimation prior calibration | `preset_small_area_estimation()` | J = 30 / 0.20 / direct; locks $I = 0.20$, $R = 3$ | Package-curated | `I`, `R` |

Three reading rules.

First — the *paradigm* column is the most consequential entry. If you
plan to call `sim_multisite()`, you need a row whose paradigm is
*site-size*. If you plan to call `sim_meta()`, you need *direct*.
Mixing them produces an immediate front-door error (Section 7).

Second — *source* tells you what published number you are inheriting.
The three education presets share Weiss et al. (2017) as their
sigma-tau anchor [@weiss2017howmuch]. The two JEBS presets implement
the simulation design from @lee2025improving. `preset_walters_2024()`
reproduces the calibration in the *Handbook of Labor Economics*,
volume 5 chapter by @walters2024empirical. The
remaining three presets are package-curated rather than
paper-anchored and carry no external citation.

Third — *override* is the field most studies change first. Education
presets retain $\sigma_\tau$ but adjust $J$ to match the local sample
of schools; the Walters preset retains $J$ and $\sigma_\tau$ but
adjusts the covariate $R^2$. JEBS rows show no
override because their value is reproducing the published number
exactly — change a JEBS field and the preset name no longer applies.

### A compact tree view (alternative)

Readers who prefer branching to tabular logic can walk the same
rubric as a four-question tree:

```
Q1 paradigm
├── multisite trial -> Q2 calibration source
│   ├── JEBS paper        -> Q3 grid
│   │   ├── paper-grid    -> preset_jebs_paper()
│   │   └── strict-grid   -> preset_jebs_strict()
│   ├── Walters Handbook  -> preset_walters_2024()
│   ├── Education applied -> Q4 effect-size scale
│   │   ├── small  ~0.05  -> preset_education_small()
│   │   ├── modest ~0.20  -> preset_education_modest()
│   │   └── subst. ~0.30  -> preset_education_substantial()
│   └── stress test       -> preset_twin_towers()
└── meta-analysis     -> Q5 special requirements
    ├── methods warm-up   -> preset_meta_modest()
    └── small-area prior  -> preset_small_area_estimation()
```

If a reviewer asks why a particular preset, you trace four nodes
through this tree and you have your answer.

### When to override versus pick a different row

If you find yourself overriding more than two fields, you are in a
different row. Move to the row whose locked parameters are closest to
your design, then override at most one or two. The presets are not
fragile — every preset accepts named overrides through `...` — but
their *defensibility* comes from the published anchor. Overriding the
sigma-tau of a Weiss-anchored preset to 0.40 silently breaks the
anchor. Pick `preset_education_substantial()` instead and tune from
there.

## 4. Override patterns

Overrides are flat, named arguments forwarded to the underlying
[`multisitedgp_design()`](../reference/multisitedgp_design.html)
constructor. Anything you do not name is left at the preset default.

```{r override-design}
local_design <- preset_education_modest(
  J          = 20L,
  nj_mean    = 60,
  sigma_tau  = 0.25
)

local_design
```

Three things to read off the printed object. The header shows $J$
moved from the preset default 50 to 20. The Layer 1 $\sigma_\tau$
moved from 0.20 to 0.25 — a modest tune-up to match a local pilot. In
Layer 2 (Margin), `nj_mean` moved from 50 to 60. Every unnamed field —
`cv = 0.5`, `nj_min = 10`, the dependence settings, the observation
hook — survives untouched from the preset.

Now simulate from the override design and check that the realized
diagnostics still pass:

```{r override-sim}
local_dat <- sim_multisite(local_design, seed = 2027L)
summary(local_dat)
```

The realized $\sigma_\tau$ landed at 0.273 against the target 0.250
(WARN, $\sim 9$ percent relative drift — small-J sampling noise, not
a design error). The dependence diagnostic on the rank scale passes
at $-0.103$ against target zero. Operational feasibility flags WARN
because $J = 20$ leaves the effective per-site sample size near 8 —
expected for the small-J override.

The general override rules:

- **One field at a time first.** Bump $\sigma_\tau$, re-render, check
  the diagnostics. Only move on if the realized values look right.
- **Re-render `summary()` after every override.** A field move can
  push a previously-passing diagnostic into WARN.
- **Document the override in your provenance string.** The
  [`canonical_hash()`](../reference/canonical_hash.html) and
  [`provenance_string()`](../reference/provenance_string.html) carry
  the override into the methods appendix automatically — you do not
  need to spell out the field changes in prose.

## 5. Visual comparison across presets

A scannable table loses one piece of context: how the realized site
effects actually look across presets. The plot below pulls three
presets covering the heterogeneity range — small (Weiss low end),
modest (JEBS / Weiss middle), and substantial (Weiss upper, large
J) — and overlays their $\widehat\tau_j$ distributions on a common
axis.

```{r preset-compare-data, include = FALSE}
.compare_presets <- function() {
  pres <- list(
    education_small       = preset_education_small(),
    education_modest      = preset_education_modest(),
    education_substantial = preset_education_substantial()
  )
  rows <- lapply(names(pres), function(nm) {
    d <- sim_multisite(pres[[nm]], seed = 1L)
    tibble::tibble(
      preset    = nm,
      site      = d$site_index,
      tau_hat   = d$tau_j_hat,
      se        = d$se_j,
      sigma_tau = pres[[nm]]$sigma_tau
    )
  })
  do.call(rbind, rows)
}
compare_df <- .compare_presets()
```

```{r preset-compare-plot, fig.cap = "Realized observed estimates across three Weiss-anchored presets at seed = 1L. Vertical spread widens as the preset's sigma_tau grows (0.05, 0.20, 0.30 — left to right). The fixed sigma_tau anchor of each preset is the dashed reference line. The small-effects column's cloud nearly collapses to zero because the latent SD is small relative to sampling error; the substantial column's cloud is widest, with J = 100 narrowing each site's bar and pulling the cloud closer to its dashed band.", fig.alt = "Three side-by-side strip plots of observed site effects across the three Weiss-anchored education presets, showing increasing vertical spread as sigma_tau grows."}
ggplot2::ggplot(
  compare_df,
  ggplot2::aes(x = factor(preset, levels = c("education_small",
                                             "education_modest",
                                             "education_substantial")),
               y = tau_hat)
) +
  ggplot2::geom_jitter(width = 0.18, alpha = 0.55, size = 1.4) +
  ggplot2::geom_hline(
    data    = unique(compare_df[, c("preset", "sigma_tau")]),
    mapping = ggplot2::aes(yintercept = sigma_tau),
    linetype = "dashed", color = "grey45"
  ) +
  ggplot2::geom_hline(
    data    = unique(compare_df[, c("preset", "sigma_tau")]),
    mapping = ggplot2::aes(yintercept = -sigma_tau),
    linetype = "dashed", color = "grey45"
  ) +
  ggplot2::labs(
    x        = NULL,
    y        = expression(hat(tau)[j]),
    title    = "Site-effect spread across three education presets",
    subtitle = "Dashed lines mark each preset's sigma_tau target"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

What to read off this plot, in this order.

1. **Vertical spread.** The substantial-effects column should be
   visibly wider than the modest and small columns; that is the
   $\sigma_\tau$ ladder you locked in by choosing one preset over
   another.
2. **Bar density at the dashed lines.** Roughly two-thirds of the
   site dots should fall inside each preset's $\pm \sigma_\tau$
   band — the empirical rule for an SD-scale parameter.
3. **The small column collapsing.** When $\sigma_\tau = 0.05$, latent
   variation is dominated by sampling noise; the dots scatter widely
   relative to the dashed band. That is the calibration warning
   `preset_education_small()` is meant to make explicit, not a bug
   in the preset.

## 6. Diagnostic per preset — funnel for `preset_jebs_paper()`

The visual contract for an individual preset is its funnel plot.
Below is the funnel for `preset_jebs_paper()` — the rank-correlation
target is zero, so a symmetric scatter top-to-bottom is the expected
signature. If the cloud tilts, the design has accidentally induced
precision–effect dependence and the rank-correlation diagnostic in
Group B will flip from PASS to FAIL.

```{r jebs-funnel, fig.cap = "Funnel plot for preset_jebs_paper() at seed = 1L. Standard error on the x-axis (smaller = more precise); observed estimate on the y-axis. The cloud tapers as SE shrinks because the design generates SE from site sizes — the funnel signature of the site-size-driven path. Symmetry top-to-bottom confirms the rank-correlation target rho_S = 0 (PASS at -0.193, well within sampling tolerance for J = 50).", fig.alt = "Funnel plot for preset_jebs_paper showing the standard error on the x-axis and observed estimate on the y-axis, with roughly symmetric scatter around zero."}
dat_jebs <- sim_multisite(preset_jebs_paper(), seed = 1L)
plot_funnel(dat_jebs, caption = FALSE) +
  ggplot2::labs(
    title    = "preset_jebs_paper() — funnel diagnostic",
    subtitle = "Symmetric cloud confirms rho_S target = 0"
  )
```

The same visual contract applies to every site-size preset; only the
spread and the funnel-mouth width change. For
`preset_education_small()` the cloud is narrower vertically; for
`preset_twin_towers()` it is enormous (J = 1000, $\sigma_\tau = 2$).
The diagnostic question — "is the cloud symmetric around zero" —
stays identical. See
[Diagnostics in practice](a3-diagnostics-in-practice.html) for the
full four-group rubric.

## 7. Wrong-door checks

Front doors are intentionally strict. A site-size preset belongs to
[`sim_multisite()`](../reference/sim_multisite.html), and a
direct-precision preset belongs to
[`sim_meta()`](../reference/sim_meta.html). Mixing them produces a
clean error rather than a silent miscalculation.

```{r wrong-door}
tryCatch(
  sim_meta(preset_education_modest(), seed = 1L),
  error = function(e) conditionMessage(e)
)
```

Read the error: the wrapper says it requires `paradigm = "direct"`,
notes the design carried `paradigm = "site_size"`, and points you to
the right front door. Reverse the example and the same guard fires
the other way.

The second wrong-door check is a calibration FAIL inside a preset
that the user picked at the wrong scale. If Maya picks
`preset_education_small()` ($\sigma_\tau = 0.05$) for a study whose
real heterogeneity is closer to 0.20, the simulation runs without
error — but the diagnostics catch it.

```{r small-fail-sim}
small_dat <- sim_multisite(preset_education_small(), seed = 1L)
summary(small_dat)
```

Read the diagnostics. The realized $\sigma_\tau$ comes in at 0.042
against a target of 0.050 — that is a FAIL, even though the small-J
($J = 50$) sampling noise is the dominant cause. The implied
operational feasibility flag at the bottom (mean shrinkage, average
MOE, feasibility index) fires a red status because the latent signal
is small relative to sampling error. The simulation ran cleanly — it
is the *design* that is wrong for this scenario. Switch to
`preset_education_modest()` and the same diagnostics return to PASS
or WARN territory at appropriate magnitudes.

## 8. Defending the choice

Once a preset is selected and any overrides are applied, the
provenance string carries the choice and the seed into the methods
appendix in one line.

```{r provenance}
chosen_dat <- sim_multisite(preset_jebs_paper(), seed = 1L)
provenance_string(chosen_dat)
```

The string contains the package version, the paradigm, the seed, the
canonical hash of the simulation, the design hash, the hashing
algorithm, the R version, and any active hooks. Pasting this into a
methods appendix gives a reviewer the exact contract: identical
package version + identical seed + identical preset → identical
canonical hash → byte-identical simulation. The hash itself is the
short anchor:

```{r canonical-hash}
canonical_hash(chosen_dat)
```

Cite the source for the chosen preset in prose. For
`preset_jebs_paper()` and `preset_jebs_strict()` cite
@lee2025improving; for `preset_walters_2024()` cite
@walters2024empirical; for the three education presets cite
@weiss2017howmuch. The three remaining presets are package-curated
and carry no external citation; describe each by role ("a stress-test
preset for J = 1000 with extreme heterogeneity"; "a meta-analysis
warm-up at the direct-precision front door"; "a small-area-estimation
prior calibration") rather than naming a paper.

A reviewer-ready methods-appendix line ties it all together:

> Simulation generated with multisiteDGP (version 0.1.0, Lee 2025)
> using `preset_education_modest(J = 20L, nj_mean = 60, sigma_tau =
> 0.25)`, seed = 2027L, canonical hash 4f9399b626d1c71f.

Three lines, every load-bearing claim grounded in either a citation
or a hash.

## 9. Where to next

- [A1 · Getting started](a1-getting-started.html) — if you have not
  already run a first simulation end-to-end, start there.
- [Diagnostics in practice](a3-diagnostics-in-practice.html) — the
  four diagnostic groups and the plots that go with each.
- [Covariates and precision dependence](a4-covariates-precision-dependence.html)
  — for the `R2` override on `preset_walters_2024()` and how
  precision–effect dependence enters Layer 3.
- [Calibrating to real data](a5-calibrating-real-data.html) — when
  you have observed point estimates and SEs you want the simulator
  to match.
- [Cookbook](a8-cookbook.html) — recipe walkthroughs for downstream
  analysis on a fitted preset.

For methodology — the formal two-stage DGP, the G-distribution
catalog, the *site-size-driven* and *direct-precision* paths — start
with [The two-stage DGP](m1-statistical-dgp.html) and read M2 / M3 /
M4 in order.

## References

The three education presets inherit their $\sigma_\tau$ ladder from
@weiss2017howmuch. The two JEBS presets reproduce the simulation
design in @lee2025improving. `preset_walters_2024()` is calibrated to
the Handbook chapter @walters2024empirical. The remaining three
presets — `preset_meta_modest()`, `preset_small_area_estimation()`,
`preset_twin_towers()` — are package-curated and carry no published
anchor at this writing.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
