---
title: "A3 · Diagnostics in practice"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists who can already run `sim_multisite()` and now need
  to decide whether the realized design is fit for the analysis that
  follows. This vignette walks the four-question diagnostic rubric on a
  single working design — scale, dependence, G-shape, feasibility —
  showing the function that answers each question, the plot that makes
  the answer legible, and the design move that fixes it when the answer
  comes back red. You leave with a one-page triage card mapping each
  red status to the change it implies, and a `scenario_audit()` call
  that scales the same logic across a design grid.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A3 Diagnostics in practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The four questions

A multisite simulation is only useful if four things land where the
designer asked. The four diagnostic groups answer one question each;
every Group A/B/C/D row in `summary(dat)` is a PASS / WARN / FAIL
verdict on one of them [@lee2025improving].

| Group | Question | Lead helpers | Threshold source |
|---|---|---|---|
| **A** | Did the simulation reach the target *information scale*? | `informativeness()`, `heterogeneity_ratio()`, `compute_I()`, `compute_kappa()` | `feasibility_min`, `R_max` |
| **B** | Did the *precision–effect dependence* land where the design asked? | `realized_rank_corr()`, `realized_rank_corr_marginal()` | qualitative (compare to design `rank_corr`) |
| **C** | Does the realized *G shape* match the target distribution? | `bhattacharyya_coef()`, `ks_distance()` | `bhattacharyya_min`, `ks_max` |
| **D** | Will the design *work downstream* — is shrinkage non-trivial? | `mean_shrinkage()`, `feasibility_index()`, `default_thresholds()` | `mean_shrinkage_min`, `feasibility_min` |

The rest of this vignette is a reference card: one section per group
with the call signatures, the ground-truth values for
`preset_jebs_paper(seed = 1L)`, what to read off the diagnostic plot,
and the design move that fixes it when the answer comes back red.
Sections 6 and 7 close with the triage card and a `scenario_audit()`
walkthrough.

## 2. Group A — did the simulation reach the target information scale?

[`informativeness()`](../reference/informativeness.html) and
[`heterogeneity_ratio()`](../reference/heterogeneity_ratio.html) are
the two headline scalars; both take a `multisitedgp_data` object and
return a single number.
[`compute_I()`](../reference/compute_I.html) and
[`compute_kappa()`](../reference/compute_kappa.html) are the
lower-level primitives that operate on raw vectors — useful when
sketching a candidate design before any data exist.

```{r groupA-design}
design <- preset_jebs_paper()
dat    <- sim_multisite(design, seed = 1L)
dat
```

The print header reports realized $\sigma_\tau = 0.207$ vs target
0.200 (PASS) and feasibility 13 (WARN). The two Group A scalars
directly:

```{r groupA-informativeness}
informativeness(dat)
```

```{r groupA-heterogeneity}
heterogeneity_ratio(dat$se2_j)
```

Realized $I = 0.250$ — about a quarter of the total variance in
$\widehat\tau_j$ is signal. Realized $R = 7.58$ — the largest sampling
variance across sites is roughly 7.6 times the smallest. Both clear
the public defaults
([`default_thresholds()`](../reference/default_thresholds.html) caps
$R$ at 30), so Group A passes for this design.

The lower-level primitives reproduce the same number from raw inputs
— the calculation you run *before* simulating, when sketching a
candidate $(J, \bar{n}_j, \sigma_\tau)$:

```{r groupA-compute-I}
compute_I(dat$se2_j, sigma_tau = 0.2)
```

```{r groupA-compute-kappa}
compute_kappa(p = 0.5, R2 = 0.2, var_outcome = 1)
```

[`compute_kappa()`](../reference/compute_kappa.html) gives the
precision constant feeding the *site-size-driven* path (Paradigm A in
the package's underlying blueprint) where
$\widehat{se}^2 \approx \kappa / \bar{n}$ — telling you, at the design
stage, how the implied per-site SE scales with $\bar{n}$.

### When Group A fires red — what to fix

If realized $I$ is low, either raise `nj_mean` (more data per site
lifts precision and pushes $I$ up) or revisit the $\sigma_\tau$
target — a low $I$ at a believed $\sigma_\tau$ often means the
believed value is too low. If realized $R$ is excessive, narrow the
site-size distribution by raising `nj_min` or lowering `cv` at the
[`multisitedgp_design()`](../reference/multisitedgp_design.html)
constructor. Both moves are picked up by the printed diagnostic on
the next simulation.

### Plot 1 — informativeness across a small preset shelf

A single $I = 0.25$ is hard to read in isolation. Plot 1 places it on
the same axis as three education presets so the JEBS design's
position on the field's working range is visible.

```{r groupA-plot-data, include = FALSE}
.preset_I_grid <- function() {
  pres <- list(
    education_small       = preset_education_small(),
    education_modest      = preset_education_modest(),
    education_substantial = preset_education_substantial(),
    jebs_paper            = preset_jebs_paper()
  )
  rows <- lapply(names(pres), function(nm) {
    d <- sim_multisite(pres[[nm]], seed = 1L)
    tibble::tibble(
      preset    = nm,
      sigma_tau = pres[[nm]]$sigma_tau,
      I_hat     = informativeness(d),
      R_hat     = heterogeneity_ratio(d$se2_j)
    )
  })
  do.call(rbind, rows)
}
groupA_df <- .preset_I_grid()
```

```{r groupA-plot, fig.cap = "Realized informativeness across four site-size presets at seed = 1L. Each bar is one preset; the dashed reference line at I = 0.10 marks the practical lower bound below which empirical-Bayes estimates collapse toward the grand mean. The JEBS preset (I_hat = 0.250) sits comfortably above the line; the small-effects preset sits well below it — a calibration warning, not a bug, because its sigma_tau = 0.05 target is intentionally signal-poor.", fig.alt = "Bar chart of realized informativeness for four presets, with a dashed reference line at I = 0.10 and the JEBS preset highlighted."}
ggplot2::ggplot(
  groupA_df,
  ggplot2::aes(x = stats::reorder(preset, I_hat), y = I_hat,
               fill = preset == "jebs_paper")
) +
  ggplot2::geom_col(width = 0.6) +
  ggplot2::geom_hline(yintercept = 0.10, linetype = "dashed",
                      color = "grey30") +
  ggplot2::scale_fill_manual(values = c("grey55", "steelblue4"),
                             guide = "none") +
  ggplot2::labs(
    x        = NULL,
    y        = expression(hat(I)),
    title    = "Realized informativeness across four presets",
    subtitle = "Dashed line: I = 0.10 working lower bound"
  ) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal(base_size = 11)
```

Bars above the 0.10 line are healthy; bars below it mean partial
pooling dominates regardless of analyst choice. The
substantial-effects preset clears the line by the largest margin —
$\sigma_\tau = 0.30$ pushes signal-to-noise up. The small-effects bar
is small by *design* — that preset's calibration warning made
legible [@weiss2017howmuch], not a pathology.

## 3. Group B — did the dependence land where I asked?

[`realized_rank_corr()`](../reference/realized_rank_corr.html) reads
the residual-scale dependence (Spearman correlation of $z_j$ with
$\widehat{se}_j^2$); the same function also accepts
`on = "marginal"` to switch to the marginal-scale version, and
[`realized_rank_corr_marginal()`](../reference/realized_rank_corr_marginal.html)
is a convenience alias for that case (Spearman of $\widehat\tau_j$
with $\widehat{se}_j^2$). Both take a `multisitedgp_data` object.

```{r groupB-residual}
realized_rank_corr(dat)
```

```{r groupB-marginal}
realized_rank_corr_marginal(dat)
```

Both come in at $-0.193$. The JEBS preset declares
`rank_corr = 0`, so the realized value is sampling noise around zero
at $J = 50$ — well within the rule-of-thumb $\pm 0.20$ band the
package treats as PASS. Values outside flip to WARN, and farther
outside, to FAIL.

The two scales coincide here because the JEBS preset declares no
covariate-induced effects in Layer 1: with no $X\beta$ shifting
$\tau_j$, the marginal and residual ranks coincide. They diverge for
designs with covariates; that divergence is what
[A4 · Covariates and precision dependence](a4-covariates-precision-dependence.html)
walks through end-to-end.

### When Group B fires red — what to fix

First, confirm the declared dependence target on the printed
`multisitedgp_design` object — the field is `rank_corr` (Layer 3). If
the declared target is correct, verify the alignment method actually
ran: the
[`align_rank_corr()`](../reference/align_rank_corr.html) family
couples $z_j$ to $\widehat{se}_j^2$ at the requested correlation. A
non-zero declared target with a near-zero realized correlation means
alignment was bypassed; a zero target with a far-from-zero realized
correlation is sampling noise at the chosen $J$ — increase $J$ or
accept the band. See
[A4 · Covariates and precision dependence](a4-covariates-precision-dependence.html)
for the alignment workflow and
[M4 · Precision dependence](m4-precision-dependence-theory.html) for
the three injection methods.

### Plot 2 — funnel view annotated with the dependence target

The funnel plot is the visual contract for Group B: standard error on
the x-axis, observed estimate on the y-axis, and a roughly symmetric
top-to-bottom scatter is the visual signature of $\rho_S = 0$.

```{r groupB-plot, fig.cap = "Funnel plot for preset_jebs_paper() at seed = 1L. Standard error on the x-axis (smaller = more precise); observed estimate on the y-axis. Symmetric scatter top-to-bottom across the precision range confirms rho_S target = 0; realized rank correlation -0.193 sits inside the ~0.20 PASS tolerance for J = 50. A FAIL would tilt the cloud — small-SE sites would skew toward one side of zero — and the rank-correlation status would flip.", fig.alt = "Funnel plot of 50 simulated site effects against their standard errors."}
plot_funnel(dat, caption = FALSE) +
  ggplot2::labs(
    title    = "Group B funnel — does dependence land at the target?",
    subtitle = "Symmetric cloud confirms rho_S target = 0 (realized -0.193, PASS)"
  )
```

Three things to read: top-to-bottom symmetry across the precision
range is the PASS signature for $\rho_S = 0$. A *tilt* — small-SE
sites piling up above zero with large-SE sites below (or vice
versa) — is the visual FAIL. Cloud narrowing toward small SE is the
*site-size-driven* signature; that is a Group A feature, read off
Plot 1 and the $R$ scalar.

## 4. Group C — does the G shape look right?

[`ks_distance()`](../reference/ks_distance.html) is the
Kolmogorov–Smirnov supremum distance between empirical CDFs;
[`bhattacharyya_coef()`](../reference/bhattacharyya_coef.html) is the
histogram overlap coefficient. Both *require* a target sample `y` —
Group C is the only group whose content depends on what the user is
calibrating *to*, and there is no neutral default for "the right G
shape." Called without `y`, they error cleanly:

```{r groupC-no-target, error = TRUE}
ks_distance(dat$z_j)
```

For illustration, compare the JEBS mixture residuals to a Gaussian
target sample. This is a deliberately mismatched reference: it shows
how Group C reacts when the comparison distribution does not match
the declared mixture shape. Reset the seed before drawing the target
so the values below are reproducible across renders:

```{r groupC-rng-seed, include = FALSE}
set.seed(1L)
y_target <- rnorm(50)
```

```{r groupC-ks}
ks_distance(dat$z_j, y = y_target)
```

```{r groupC-bhat}
bhattacharyya_coef(dat$z_j, y = y_target)
```

Realized KS $\approx 0.28$ against `ks_max = 0.10` and Bhattacharyya
$\approx 0.62$ against `bhattacharyya_min = 0.85`. Both look like
"failures" relative to the public gates, but the JEBS preset declares
a *mixture-shape* latent rather than a Gaussian one — flagging a
mixture against a Gaussian reference is the expected behavior, not a
calibration error. The Group C row in `summary(dat)` carries the
status `target_source = "not_available"` because the JEBS preset does
not register a deterministic target G grid; in this regime the audit
treats the row as reporting-only rather than a gate. The gates *do*
fire as documented for designs that declare a Gaussian (or other
deterministic-grid) target.

```{r groupC-thresholds}
default_thresholds()[c("bhattacharyya_min", "ks_max")]
```

### When Group C fires red — what to fix

Group C FAIL on a working design (large $J$, large target sample) is
a content-level mismatch between the declared G and the data being
mimicked. Revisit the latent G choice and recalibrate against a
real-data target through the workflow in
[A5 · Calibrating to real data](a5-calibrating-real-data.html). The
[G-distribution catalog](m2-g-distribution-catalog.html) lists the
eight built-in shapes. At small $J$ a Group C FAIL is more often a
power problem than a calibration problem; the quick check is to
repeat the diagnostic at a synthetically larger $J$ and see whether
the FAIL persists.

### Plot 3 — Q-Q view of realized $z_j$ against Gaussian quantiles

For a Gaussian-G design, points line up on the 45-degree reference;
for a mixture, heavy-tailed, or skewed design, the curve bends at
the tails or the mode in a shape-specific way.

```{r groupC-plot, fig.cap = "Q-Q plot of z_j (preset_jebs_paper, seed = 1L) against theoretical Gaussian quantiles. Departures from the 45-degree reference are the signature of the mixture shape; the central segment is approximately linear because the bulk of the latent is near-Gaussian, while the tails show the mixture's lift. Read tail departures as the structural component of the KS distance from a Gaussian comparison.", fig.alt = "Q-Q plot of z_j against theoretical Gaussian quantiles, with points departing from the 45-degree reference at the upper and lower tails."}
qq_df <- tibble::tibble(
  theoretical = qnorm(ppoints(length(dat$z_j))),
  empirical   = sort(dat$z_j)
)
ggplot2::ggplot(qq_df, ggplot2::aes(theoretical, empirical)) +
  ggplot2::geom_abline(slope = 1, intercept = 0,
                       linetype = "dashed", color = "grey45") +
  ggplot2::geom_point(size = 2.0, alpha = 0.78, color = "grey20") +
  ggplot2::labs(
    x = "Theoretical Gaussian quantile",
    y = expression(z[j] * " (sorted)"),
    title = "Q-Q plot: z_j vs Gaussian quantiles",
    subtitle = "Tail departures are the signature of the mixture latent"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

Three things to read: linearity of the central segment (points near
zero close to the 45-degree line means the *bulk* of the latent is
near-Gaussian); tail-segment slope and curvature (S-shape signals
heavier-than-Gaussian tails — a mixture or Student-t; a dog-leg
signals a discrete component such as point-mass-and-slab); and
sample-size noise at $J = 50$ (use this view in concert with the
KS / Bhattacharyya numbers, not as a sole verdict).

## 5. Group D — will it work downstream?

[`mean_shrinkage()`](../reference/mean_shrinkage.html) is the average
shrinkage factor an empirical-Bayes estimator will apply [@walters2024empirical];
[`feasibility_index()`](../reference/feasibility_index.html) is an
effective-sample-size metric that gates whether partial pooling
stabilizes at all.

```{r groupD-shrinkage}
mean_shrinkage(dat)
```

```{r groupD-feasibility}
feasibility_index(se2_j = dat$se2_j, sigma_tau = 0.2)
```

Realized mean shrinkage is 0.262 — slightly *below* the canonical
threshold `mean_shrinkage_min = 0.30`. This earns a WARN status: the
design does enough partial-pooling work to benefit from a hierarchical
estimator, but it is below the recommended floor. Feasibility index
13.1 — the Efron form, comfortably above the gate `feasibility_min =
5.0`. This is the equivalent of about 13 perfectly-informative sites'
worth of information after shrinkage. PASS.

### Threshold band — written out

`default_thresholds()` returns five gates. The two that bear directly
on Group D are `mean_shrinkage_min` and `feasibility_min`; the other
three live with Groups A and C.

```{r groupD-thresholds}
default_thresholds()
```

| Gate | Group | Value | Direction | Meaning of FAIL/WARN |
|---|---|---|---|---|
| `mean_shrinkage_min` | D | 0.30 | floor | Partial pooling does too little work for a hierarchical estimator to add value. |
| `feasibility_min` | A/D | 5.0 | floor | Fewer than five sites' worth of effective information after shrinkage. |
| `R_max` | A | 30.0 | ceiling | Ill-conditioned: a small subset of sites dominates the precision-weighted average. |
| `bhattacharyya_min` | C | 0.85 | floor | Realized G overlaps the target by less than 85 percent. |
| `ks_max` | C | 0.10 | ceiling | Worst-case CDF mismatch exceeds 10 percentage points. |

To override a gate, edit the returned list and pass it to
`scenario_audit()` via the `thresholds` argument:

```{r groupD-override, eval = FALSE}
th <- default_thresholds()
th$mean_shrinkage_min <- 0.25  # accept slightly lower S-bar
audit <- scenario_audit(grid, M = 50L, thresholds = th)
```

### When Group D fires red — what to fix

To lift the feasibility index, raise $J$ or `nj_mean`. To lift mean
shrinkage, raise $\sigma_\tau$ if the design plausibly supports it —
partial pooling does more work when between-site variation is large
relative to within-site noise. When both scalars sit far below
threshold, reconsider whether partial pooling is the right tool at
all: a fixed-effects estimator (no pooling) or a grand-mean
estimator (full pooling) may be more honest than empirical Bayes.

### Plot 4 — feasibility curve across average site size

The visual contract for Group D is a feasibility curve across average
site size at fixed $\sigma_\tau$. As $\bar{n}_j$ grows, each site's
$\widehat{se}_j^2 \approx \kappa / \bar{n}_j$ shrinks, $S_j$ grows,
and the feasibility index rises.

```{r groupD-plot-data, include = FALSE}
.feasibility_grid <- function() {
  nj_grid <- c(20, 30, 50, 80, 120, 200)
  rows <- lapply(nj_grid, function(nj) {
    d <- sim_multisite(preset_education_modest(nj_mean = nj), seed = 1L)
    tibble::tibble(
      nj_mean     = nj,
      mean_S      = mean_shrinkage(d),
      feasibility = feasibility_index(d, warn = FALSE),
      R_hat       = heterogeneity_ratio(d$se2_j)
    )
  })
  do.call(rbind, rows)
}
groupD_df <- .feasibility_grid()
```

```{r groupD-plot, fig.cap = "Efron feasibility index against average site size nj_mean (preset_education_modest, sigma_tau = 0.20, seed = 1L). The dashed reference at feasibility = 5 is the default lower gate; the curve grows because larger sites shrink each per-site SE and lift S_j toward 1. Use this curve to read the smallest defensible nj_mean for a target feasibility.", fig.alt = "Line plot of feasibility index against nj_mean, showing a monotone increase with a horizontal dashed reference at feasibility = 5."}
ggplot2::ggplot(groupD_df, ggplot2::aes(nj_mean, feasibility)) +
  ggplot2::geom_hline(yintercept = 5, linetype = "dashed", color = "grey45") +
  ggplot2::geom_line(linewidth = 0.9, color = "grey25") +
  ggplot2::geom_point(size = 2.6, color = "grey15") +
  ggplot2::geom_text(ggplot2::aes(label = sprintf("%.1f", feasibility)),
                     vjust = -0.9, size = 3.3) +
  ggplot2::labs(
    x = expression(bar(n)[j]),
    y = "Efron feasibility index",
    title = "Feasibility rises with average site size",
    subtitle = "preset_education_modest(), sigma_tau = 0.20; dashed line is the default gate (5.0)"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

What to read: position relative to the dashed gate (above = clears,
below = Group D FAIL); slope between adjacent points is the marginal
feasibility you buy by adding 10 students per site (a flat slope at
large $\bar{n}_j$ means diminishing returns — recruiting more sites
buys more feasibility than enlarging existing ones); and where the
curve crosses the gate is the *minimum defensible* average site size
at this $\sigma_\tau$.

## 6. Triage workflow — what to do when a diagnostic fires red

The four sections above each end with a short "what to fix" rule.
The card below collects all four. Paste it next to your simulation
script or run it as the checklist on your next design review.

| Group | What it asks | How to read | If FAIL — first move | Where the fix lives |
|---|---|---|---|---|
| **A — scale** | Did $I$ and $R$ hit their targets? | `informativeness(dat)`, `heterogeneity_ratio(dat$se2_j)`. Low $I$ (< 0.10), high $R$ (> 30). | Raise `nj_mean` (lifts $I$) or revisit $\sigma_\tau$. Narrow site sizes if $R$ excessive. | [`multisitedgp_design()`](../reference/multisitedgp_design.html); preset [overrides](a2-choosing-a-preset.html). |
| **B — dependence** | Did rank correlation match the declared target? | `realized_rank_corr(dat)`, `realized_rank_corr_marginal(dat)` vs. design `rank_corr`. | Verify Layer-3 alignment ran; confirm the declared target. Small $J$ widens the tolerance band. | [`align_rank_corr()`](../reference/align_rank_corr.html) family; [A4](a4-covariates-precision-dependence.html); [M4](m4-precision-dependence-theory.html). |
| **C — G shape** | Does realized $z_j$ match the target G? | `ks_distance(dat$z_j, y)`, `bhattacharyya_coef(dat$z_j, y)`. Both REQUIRE `y`. | Calibrate against a real-data target; reconsider the latent G shape. Gate is power-bound at small $J$. | [A5 · Calibrating to real data](a5-calibrating-real-data.html); [M2 · G-distribution catalog](m2-g-distribution-catalog.html). |
| **D — feasibility** | Will partial pooling work? | `mean_shrinkage(dat)`, `feasibility_index(dat$se2_j, sigma_tau)` vs. `default_thresholds()`. | Raise $J$ or `nj_mean`; reconsider the estimator if both sit deep below threshold. | [`feasibility_index()`](../reference/feasibility_index.html); estimator choice in [A6](a6-case-study-multisite.html). |

Three reading rules. *Answer Group A first* — scale is the earliest
design check and a Group A FAIL is almost always a design problem,
not a downstream-tool problem. *The Group B rule asks for the declared
target, not the realized number* — $-0.193$ against a declared $0$
is PASS at $J = 50$; the realized number alone is not actionable.
*The Group D rule has the longest fix list* — lifting feasibility
requires more sites, more data per site, or a different estimator,
none of which is free [@walters2024empirical].

## 7. Audit a grid before running the full study

A single working design is one row; a real simulation campaign covers
a grid. [`design_grid()`](../reference/design_grid.html) builds the
factorial grid; [`scenario_audit()`](../reference/scenario_audit.html)
runs the four-question rubric on every cell with $M$ replicates and
returns one summary row per cell. `scenario_audit()` requires its
first argument to be a `multisitedgp_design_grid` (the output of
`design_grid()`), not an arbitrary list of designs.

```{r grid}
g <- design_grid(
  base_design = preset_education_modest(),
  J           = c(30L, 50L),
  sigma_tau   = c(0.15, 0.20),
  seed_root   = 12345L
)
g
```

Four cells, one per $(J, \sigma_\tau)$ combination. The `design`
list-column carries the per-cell `multisitedgp_design`, and `seed`
carries a deterministic per-cell seed derived from `seed_root`. Now
the audit at $M = 20$ replicates per cell:

```{r audit}
audit <- scenario_audit(g, M = 20L)
audit[, c("cell_id", "J", "sigma_tau", "M",
          "status", "pass", "n_violations")]
```

`scenario_audit()` returns a tibble with 25 columns. The eight you
reach for first:

- **`cell_id`** — integer 1, 2, ...; matches the `design_grid()`
  printout.
- **`M`** — replicate count actually used; carries through so the
  audit is self-documenting.
- **`status`** — one-word verdict per cell: `PASS` / `WARN` / `FAIL`.
- **`pass`** — logical companion; `TRUE` only if no gate fired. Use
  it to filter the audit tibble.
- **`n_violations`** — count of distinct gates that fired. Three
  simultaneous gates point at where to look; one often resolves with
  a single override.
- **`fail_reasons`** — comma-separated names of gates at FAIL. The
  column to read when triaging a red cell — it points at which of
  the four groups broke.
- **`warn_reasons`** — same column for WARN-level gates.
- **`med_I_hat`, `q05_I_hat`, `q95_I_hat`** — median and 5/95
  quantiles of realized $I$ across the cell's $M$ replicates. The
  median is the headline; the bracket says how stable the median is.

The remaining 16 columns repeat the median/quantile pattern for $R$,
mean shrinkage, the two flavors of feasibility, Bhattacharyya, and
KS. Read them in groups of three, anchored on the medians:

```{r audit-medians}
audit[, c("cell_id", "J", "sigma_tau",
          "med_I_hat", "med_R_hat", "med_mean_shrinkage",
          "med_feasibility_efron", "fail_reasons")]
```

Read down the `fail_reasons` column to see which gate is firing in
each cell. The audit is honest about $M = 20$ being a *smoke* audit
— the across-replicate medians stabilize as $M$ grows, and a cell
that flags FAIL at $M = 20$ but clears at $M = 200$ is revealing
sampling noise, not a structural failure of the design.

### Recommended workflow

1. Build the grid with `design_grid(..., seed_root = ...)`.
2. Run `scenario_audit(grid, M = 50L)` for a fast pre-check.
3. Drop or revise cells that fail the gates that matter for your
   study — Group A and D are usually the binding constraints; Group
   C binds only when the design declares a deterministic target G.
4. Run the full simulation only on the surviving cells.

For an end-to-end case study walking this workflow, see
[A6 · Case study — multisite trial](a6-case-study-multisite.html).
For wiring the audit's hashes into a methods appendix, see
[Reproducibility and provenance](m7-reproducibility-provenance.html).

## 8. Where to next

- [A1 · Getting started](a1-getting-started.html) and
  [A2 · Choosing a preset](a2-choosing-a-preset.html) — prior steps.
- [A4 · Covariates and precision dependence](a4-covariates-precision-dependence.html)
  — the Layer-3 alignment workflow that drives Group B.
- [A5 · Calibrating to real data](a5-calibrating-real-data.html) —
  the workflow when Group C fires on a real-data target.
- [A6 · Case study — multisite trial](a6-case-study-multisite.html)
  — applies the §6 triage card to a realistic design grid.

For methodology, start with [The two-stage DGP](m1-statistical-dgp.html)
and read M2 / M3 / M4 in order.

## References

The four-question rubric is introduced in @lee2025improving; the
public thresholds in `default_thresholds()` are the package's
adaptation of the JEBS gates. Empirical-Bayes context for Group D is
collected in @walters2024empirical. Education calibration anchors
referenced in §2 are taken from @weiss2017howmuch.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
