---
title: "A4 · Covariates and precision dependence"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For trial designers whose multisite simulation must respect a known
  site-level covariate — district size, baseline performance, urbanicity —
  *and* whose analysis depends on the dependence between site effects
  and their sampling precision. This vignette walks through one
  covariate-bearing design end-to-end: build the design with
  `multisitedgp_design(formula = ~ x_site, beta = ...)`, target
  residual-scale precision dependence, and read the residual *and*
  marginal rank correlations side by side. You leave with a working
  pattern for reporting both numbers and a clear sense of when the two
  scales diverge.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A4 Covariates and precision dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why this matters for the trial designer

Most multisite simulations carry a covariate — the sites are not
exchangeable, and a designer who knows that district size, baseline
achievement, or urbanicity predicts site-level effects has to encode
that knowledge somewhere. The package encodes it through a one-sided
formula on `multisitedgp_design()`: `formula = ~ x_site` declares that
the latent site effect $\tau_j$ is the sum of a covariate term
$\beta\, x_j$ and a *standardized residual* $\sigma_\tau z_j$ with
$\mathrm{Var}(z_j) = 1$ [@lee2025improving]. The residual $z_j$ — not
the marginal $\tau_j$ — is the column the package's downstream
machinery treats as the unit of analysis: dependence alignment,
G-shape diagnostics, and shrinkage all read off $z_j$.

The implication is sharp. When the design declares a covariate, the
*residual* rank correlation between site effects and sampling variance
is what Layer 3 alignment targets, and what `realized_rank_corr()`
reads back. The *marginal* rank correlation — the Spearman correlation
of $\tau_j$ with $\widehat{se}_j^2$ that a downstream analyst who has
not been told about the covariate would compute — is a derived
quantity that depends on how $x_j$ correlates with sampling variance
and how strong $\beta$ is. The two numbers can differ in sign and
magnitude. A designer who reports only the marginal number (the one
the analyst will see) loses the information about the residual
contract the simulation actually honored. A designer who reports only
the residual number leaves the analyst guessing about the surface they
will encounter. Report both whenever the design carries a covariate.

## 2. Add a site-level covariate

Build a one-row-per-site `data.frame` of covariate values, then pass
it to [`multisitedgp_design()`](../reference/multisitedgp_design.html)
through the `data` argument. The `formula` argument is *one-sided* —
no left-hand side, because the package generates the response. With an
intercept implied by `~ x_site`, an unnamed `beta = 0.30` is
interpreted as the non-intercept coefficient.

```{r covariate-data}
site_data <- data.frame(
  x_site = as.numeric(scale(seq_len(30L)))
)

design <- multisitedgp_design(
  J       = 30L,
  formula = ~ x_site,
  beta    = 0.30,
  data    = site_data,
  seed    = 77L
)

design
```

The printed Layer 1 block reports `formula: ~x_site`, `beta: 0.3`, and
the standardized residual scale `sigma_tau: 0.2`. The covariate column
is now part of the design object; simulating populates `x_site`
alongside the residual `z_j` and the marginal $\tau_j$.

```{r covariate-sim}
dat <- sim_multisite(design)
head(tibble::as_tibble(dat)[,
  c("site_index", "x_site", "z_j", "tau_j", "tau_j_hat", "se2_j")
], 6)
```

`tau_j` carries the covariate component plus residual variation;
`z_j` is the residual that downstream machinery reads. Two columns,
two scales. The rest of this vignette is about which one to target,
and how to report both.

## 3. The two-scale split

When the design declares a covariate, the *residual* and *marginal*
scales diverge. Concretely:

- **Residual scale.** The standardized residual $z_j$ satisfies
  $\mathrm{Var}(z_j) = 1$ by convention. The Spearman correlation of
  $z_j$ with $\widehat{se}_j^2$ is the *residual rank correlation* —
  what the package's Layer 3 alignment actually targets, and what
  [`realized_rank_corr(dat, on = "residual")`](../reference/realized_rank_corr.html)
  reads back. This is the package's primary contract.
- **Marginal scale.** The site effect $\tau_j = \tau + \beta\,x_j +
  \sigma_\tau z_j$ is the marginal site effect — what a downstream
  analyst sees when the covariate is not modeled. The Spearman
  correlation of $\tau_j$ with $\widehat{se}_j^2$ is the *marginal
  rank correlation*, returned by
  [`realized_rank_corr_marginal(dat)`](../reference/realized_rank_corr_marginal.html).

Without a covariate, the two scales coincide — $\tau_j$ is a linear
function of $z_j$ alone, and Spearman correlations are invariant to
monotone transforms. With a covariate, the scales are different
random variables, and their rank correlations against
$\widehat{se}_j^2$ generally differ. The headline lesson: targeting
one scale does not pin the other.

## 4. Target residual precision dependence

Add `dependence = "rank"` and a residual-scale `rank_corr` target to
the design. The Layer 3 alignment family
([`align_rank_corr()`](../reference/align_rank_corr.html)) couples
$z_j$ to $\widehat{se}_j^2$ at the requested correlation; the
covariate component flows through unchanged.

```{r dep-design}
dep_design <- multisitedgp_design(
  J          = 30L,
  formula    = ~ x_site,
  beta       = 0.30,
  data       = site_data,
  dependence = "rank",
  rank_corr  = -0.30,
  seed       = 77L
)

dep_design
```

The Layer 3 block now reports `method: rank` and `rank_corr: -0.3`.
Simulate and read both correlations off the result:

```{r dep-sim}
dep_dat <- sim_multisite(dep_design)
canonical_hash(dep_dat)
```

The hash `5fc7e768a2581b35` is a determinism smoke check — a local
re-run at `seed = 77L` should produce the same value. Now the two
rank correlations:

```{r two-number-reporting}
data.frame(
  scale    = c("residual", "marginal"),
  spearman = c(
    realized_rank_corr(dep_dat, on = "residual"),
    realized_rank_corr_marginal(dep_dat)
  )
)
```

Residual realized $-0.306$ — a clean PASS against the declared
$-0.30$ target; alignment did its work on the scale the design
declared. Marginal realized $0.029$ — within sampling noise of zero,
and *opposite in sign* to the residual target. The covariate term
$\beta\,x_j$ is doing enough work in $\tau_j$ that the marginal-scale
ordering against $\widehat{se}_j^2$ is essentially decoupled from the
residual-scale ordering the alignment shaped.

The full diagnostic table records both rows side by side:

```{r diagnostic-table}
diag <- attr(dep_dat, "diagnostics")$target_vs_realized
diag[diag$diagnostic == "realized_rank_corr",
     c("diagnostic", "basis", "target", "realized", "status")]
```

Read down the `basis` column. The residual row carries `target = -0.3`
and `status = PASS`; the marginal row carries `target = NA` (no
declared target on that scale) and reports the realized number for
the analyst's information. Both rows are part of the artifact a
methods appendix should carry.

## 5. Two-scatterplot visualization

[`plot_dependence()`](../reference/plot_dependence.html) defaults to
the residual scale; flip `by_residual = FALSE` to inspect the marginal
scale. The residual plot is the visual contract for the alignment
target; the marginal plot is what the downstream analyst will see.

```{r residual-plot, fig.cap = "Residual-scale dependence: standardized residual z_j against sampling variance se2_j (J = 30, seed = 77, declared rank_corr = -0.30 on the residual scale). The downward smoother and realized Spearman -0.306 confirm the alignment landed on target. This is the surface Layer 3 alignment is designed to shape.", fig.alt = "Scatterplot of standardized residual site effects z_j against sampling variance se2_j with a downward smoother."}
plot_dependence(dep_dat, by_residual = TRUE, caption = FALSE) +
  ggplot2::labs(
    title    = "Residual scale — what alignment targets",
    subtitle = "Realized Spearman -0.306 against declared -0.30 (PASS)"
  )
```

```{r marginal-plot, fig.cap = "Marginal-scale dependence: site effect tau_j against sampling variance se2_j (same simulation as above). The cloud is essentially flat and the realized marginal rank correlation 0.029 is within sampling noise of zero — opposite in sign to the residual-scale target. This is the surface a downstream analyst who does not model the covariate would observe.", fig.alt = "Scatterplot of marginal site effects tau_j against sampling variance se2_j with a near-flat smoother."}
plot_dependence(dep_dat, by_residual = FALSE, caption = FALSE) +
  ggplot2::labs(
    title    = "Marginal scale — what the analyst sees",
    subtitle = "Realized Spearman 0.029 — covariate flattens the residual-scale tilt"
  )
```

The two plots tell the same story the two numbers told: alignment
shaped the residual-scale tilt; the covariate term in $\tau_j$ washes
that tilt out at the marginal scale. Showing both plots is the
discipline that prevents a methods reader from confusing the two
contracts.

## 6. Identifiability constraints

The standardized-residual convention is what makes the residual scale
identifiable: by declaring $\mathrm{Var}(z_j) = 1$, the package fixes
the units of the residual independently of the covariate component
$\beta\,x_j$. The residual is, by construction, what is left after
subtracting the covariate prediction from the latent site effect.

Two identifiability requirements follow.

- **Non-zero variation in `x_site`.** If $x_j$ is constant across
  sites, the covariate column is collinear with the intercept and
  $\beta$ is unidentified. Stating $\beta\,x_j = c$ for some constant
  $c$ is observationally equivalent to absorbing $c$ into the
  intercept, so the design constructor cannot distinguish the two.
- **`x_site` does not perfectly predict $z_j$.** The residual must
  carry genuinely orthogonal variation; if `x_site` perfectly
  reproduces $z_j$ up to sign, $\sigma_\tau z_j$ has no independent
  content beyond rescaling the covariate term. The package's residual
  generator draws $z_j$ from a declared G distribution independently
  of `x_site`, so this constraint is satisfied by construction in the
  shipped paths — flag it only if a custom `g_fn` callback is wired in
  to depend on the covariate.

The formula itself encodes the convention: `~ x_site` is
*one-sided* — there is no left-hand side because the package
generates the response. The model space is "right-hand side names a
linear combination of site-level covariates from `data`; the package
draws the standardized residual and assembles $\tau_j$ accordingly."
A two-sided formula is rejected by the constructor.

## 7. Collinearity guard

The constructor refuses to build a design when the covariate matrix
is rank-deficient. The minimal failure mode is a constant column:

```{r collinearity-illustration, eval = FALSE}
multisitedgp_design(
  J       = 30L,
  formula = ~ x,
  beta    = 0.30,
  data    = data.frame(x = rep(1, 30L)),
  seed    = 77L
)
#> Error: covariate column 'x' has zero variation; cannot identify beta.
#> Provide a column with non-zero variance, or drop the covariate.
```

The error fires inside the design constructor — before any simulation
is attempted — so the failure mode is caught up front. The same guard
catches more general collinearity (one column equal to a linear
combination of the others, or two perfectly correlated columns) by
checking the design matrix's rank against its column count. The
practical workflow is: pull the covariate frame from your real data,
inspect column variances and pairwise correlations, then pass it to
the constructor. Any rank deficiency is reported with the offending
column name in the error message.

## 8. When residual versus marginal is appropriate

The residual scale is the package's primary contract for three
reasons. The Layer 3 alignment family targets it directly; the
G-shape diagnostics
([`bhattacharyya_coef()`](../reference/bhattacharyya_coef.html),
[`ks_distance()`](../reference/ks_distance.html)) read off $z_j$; and
the standardized-residual convention $\mathrm{Var}(z_j) = 1$ is what
makes those diagnostics comparable across designs that declare
different covariates [@lee2025improving]. A design that declares
`rank_corr = -0.30` is making a residual-scale promise — the realized
value should land near $-0.30$ on the residual scale, regardless of
what the marginal-scale number does.

The marginal scale is what downstream analysts see. An analyst who
fits a hierarchical model without including `x_site` as a site-level
covariate is operating on the marginal $\tau_j$, and the dependence
they encounter between $\tau_j$ and $\widehat{se}_j^2$ is the marginal
rank correlation. For empirical-Bayes shrinkage and meta-analytic
pooling, that marginal dependence is what shapes the estimator's
behavior in finite samples [@walters2024empirical]. Reporting only
the residual number is therefore incomplete when covariates are
present — it tells the methods reader what the simulation contract
was, but not what the analyst will encounter.

The discipline is straightforward. When a design carries a covariate,
report both rank correlations every time:

```{r both-numbers-final}
data.frame(
  scale    = c("residual (alignment target)", "marginal (analyst surface)"),
  spearman = c(
    realized_rank_corr(dep_dat, on = "residual"),
    realized_rank_corr_marginal(dep_dat)
  )
)
```

Two scales, two numbers, one design. The diagnostic table
(`attr(dep_dat, "diagnostics")$target_vs_realized`) preserves both
rows; a methods appendix should carry both. When the design declares
no covariate, the two scales coincide and the second number adds
nothing — but in the covariate case the second number is the part the
analyst lives with.

## 9. Where to next

- [A1 · Getting started](a1-getting-started.html) — the no-covariate
  baseline case, where residual and marginal scales coincide.
- [A2 · Choosing a preset](a2-choosing-a-preset.html) — preset
  starting points for designs that may or may not carry covariates.
- [A3 · Diagnostics in practice](a3-diagnostics-in-practice.html) —
  the four-question rubric, including the Group B section on
  precision dependence.
- [M3 · Margin and SE models](m3-margin-se-models.html) — how the
  site-size-driven path generates the $\widehat{se}_j^2$ that the
  rank correlation is computed against.
- [M4 · Precision dependence](m4-precision-dependence-theory.html) —
  the three injection methods, the algorithmic detail behind Layer 3
  alignment.

## References

The standardized-residual convention is laid out in @lee2025improving;
the empirical-Bayes shrinkage context for the marginal scale is
collected in @walters2024empirical.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
