---
title: "A5 · Calibrating to real data"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For trial designers who already have a pilot study, an analogous
  trial, or a public summary of per-site standard errors and need the
  simulation to mimic that observed precision distribution rather than
  a guessed `sigma_tau`. This vignette walks the calibration trick
  end-to-end: back-calculate `sigma_tau` from the observed geometric
  mean of $\widehat{se}_j^2$ at a researcher-chosen target
  informativeness $I$, override a preset with the calibrated fields,
  audit the resulting design grid, and freeze the audit hash into a
  reproducible manifest. You leave with one defensible
  `multisitedgp_design`, two calibration plots, and a hash you can
  cite in a methods appendix.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A5 Calibrating to real data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why calibrate to real data?

A multisite simulation is only as honest as its precision. The
between-site variance $\sigma_\tau^2$ is the single number that
determines how much partial-pooling work a hierarchical estimator can
do; per-site sampling variances $\widehat{se}_j^2$ determine how
unequal that work is across sites. Pulling $\sigma_\tau$ from a guess
— even a literature-informed guess — leaves the simulation's
informativeness $I = \sigma_\tau^2 / (\sigma_\tau^2 + \mathrm{GM}(\widehat{se}^2))$
untethered from the precision distribution the design needs to mimic.
The downstream cost is concentrated where it hurts most: power
calculations and minimum-detectable-effect tables read
$\sigma_\tau$ directly, so a 50-percent miss on $\sigma_\tau$
translates into a several-fold miss on the smallest effect a study can
reliably detect [@lee2025improving].

The calibration target is therefore $I$, not $\sigma_\tau$.
Researchers have a defensible target for $I$ — a working lower bound
near $0.10$, an upper bound near $0.50$, and a typical education-trial
range of $0.20$–$0.40$ [@weiss2017howmuch]. They have a directly
observed value for $\mathrm{GM}(\widehat{se}^2)$ from the pilot, the
analogous trial, or a forest plot. What they do not have, and have no
reliable way to guess, is $\sigma_\tau$. The trick is that they do not
need it: $\sigma_\tau$ is algebraically determined by the other two.

## 2. The calibration trick — back-calculate $\sigma_\tau$ from observed $\widehat{se}^2$

The package's informativeness identity is the Lee–Che–Rabe-Hesketh
informativeness scalar [@lee2025improving]:

$$
I \;=\; \frac{\sigma_\tau^2}{\sigma_\tau^2 + \mathrm{GM}(\widehat{se}^2)}.
$$

This relationship is one equation in three quantities. Two are
determined: $\mathrm{GM}(\widehat{se}^2)$ is observed from the source
study; $I$ is chosen by the researcher as the calibration target.
Solving for the third gives the back-calculation:

$$
\sigma_\tau^2 \;=\; \frac{I \cdot \mathrm{GM}(\widehat{se}^2)}{1 - I}.
$$

This is the calibration trick. It is *defensible* for three reasons.
First, neither input is a guess: $\mathrm{GM}(\widehat{se}^2)$ is
observed; $I$ is the researcher's explicit target with public bounds
in the literature. Second, the algebraic step is invertible — the
package's [`compute_I()`](../reference/compute_I.html) takes
$\widehat{se}^2$ and $\sigma_\tau$ and returns $I$, so the
back-calculated $\sigma_\tau$ can be re-checked by passing it through
`compute_I()`. Third, the simulation honors the precision distribution
of the source by construction so long as the site-size distribution is
matched. The vignette ends with a hash pinning the calibrated design
to a reproducible artifact.

The minimal back-calculation chunk:

```{r back-calc}
observed_se2 <- c(0.060, 0.075, 0.080, 0.095, 0.120, 0.150)
target_i     <- 0.30
gm_se2       <- exp(mean(log(observed_se2)))
sigma_tau    <- sqrt(target_i * gm_se2 / (1 - target_i))

data.frame(
  target_I  = target_i,
  GM_se2    = gm_se2,
  sigma_tau = sigma_tau,
  check_I   = compute_I(observed_se2, sigma_tau = sigma_tau)
)
```

The `observed_se2` vector is six per-site sampling variances pulled
from a hypothetical pilot — replace it with the vector your source
study reports. `target_i = 0.30` is the calibration target;
`gm_se2 = 0.0922` is the observed geometric mean. The back-calculated
`sigma_tau = 0.199` is what the simulation needs in order for its
realized $I$ to land near $0.30$ — and the `check_I` column confirms
it: feeding the back-calculated $\sigma_\tau$ back through
`compute_I()` returns $I = 0.300$ exactly. That round-trip is the
self-check; if the columns disagree, the back-calc has a typo.

## 3. Override a preset with calibrated fields

A preset is a thin wrapper around
[`multisitedgp_design()`](../reference/multisitedgp_design.html) that
ships with sensible education-trial defaults. The calibration workflow
is to start from the preset that matches the source study's structure,
then override the calibrated fields:

```{r override}
design <- preset_jebs_paper(
  J         = 30L,
  sigma_tau = sigma_tau,
  true_dist = "Mixture",
  theta_G   = list(delta = 4.5, eps = 0.25, ups = 1.8)
)
design
```

Three overrides. `J = 30L` matches the source study's site count; if
your source has $J = 50$, set it to that. `sigma_tau` is the
back-calculated value from §2 — the bound between $\sigma_\tau$ and
$\mathrm{GM}(\widehat{se}^2)$ at the target $I$. `true_dist =
"Mixture"` and `theta_G = list(delta = 4.5, eps = 0.25, ups = 1.8)`
are optional; supply them when the source study has G-shape evidence
(a histogram of site effects, qualitative bimodality, prior knowledge
of subgroup-by-site interaction). When absent, drop both overrides
and the preset default persists.

The printed Layer 1 block reports `true_dist: Mixture`, `tau: 0`, and
`sigma_tau: 0.199` — the calibrated residual scale. Layer 2 reports
the *site-size-driven* path (`Paradigm A` in the package's underlying
blueprint) with `nj_mean: 40` from the preset. Simulate and print the
diagnostic summary:

```{r override-sim}
dat <- sim_multisite(design, seed = 7301L)
summary(dat)
```

The Group A row reports realized $\sigma_\tau = 0.200$ against target
$0.199$ — PASS at relative error $0.7\%$, the calibration's primary
contract. Realized $I = 0.252$ sits below the $0.30$ target because
the simulation generates fresh per-site sampling variances under the
preset's site-size model, so its $\mathrm{GM}(\widehat{se}^2)$ differs
from the observed geometric mean. Realized $R = 10$ is the
heterogeneity ratio across simulated per-site sampling variances; the
Group B rank correlation $-0.166$ is sampling noise around the
declared $\rho_S = 0$ at $J = 30$. Group D reports mean shrinkage
$0.264$ — a WARN signaling the partial-pooling boundary at this $J$.

## 4. Calibration plots

Two plots make the calibration legible: the *informativeness curve*
(how realized $I$ depends on the assumed $\sigma_\tau$) and the
*target overlay* (whether the simulation's per-site sampling variances
match the observed distribution).

### Plot 1 — informativeness curve, with the calibrated point highlighted

The relationship $I(\sigma_\tau \mid \mathrm{GM}(\widehat{se}^2))$ is a
strictly increasing function of $\sigma_\tau$ at fixed
$\mathrm{GM}(\widehat{se}^2)$. The back-calculation finds the unique
$\sigma_\tau$ at which the curve crosses the chosen target $I$.
Plotting that curve makes the precision of the calibration visible —
small misses on $\sigma_\tau$ map to small misses on $I$ in the
neighborhood of the target, but large guesses translate to large
informativeness misses.

```{r plot1-data, include = FALSE}
sigma_grid <- seq(0.01, 0.40, length.out = 80)
I_grid     <- vapply(
  sigma_grid,
  function(s) compute_I(observed_se2, sigma_tau = s),
  numeric(1)
)
guesses <- c(0.05, 0.10, 0.15, sigma_tau, 0.25, 0.30)
guess_I <- vapply(
  guesses,
  function(s) compute_I(observed_se2, sigma_tau = s),
  numeric(1)
)
plot1_curve <- data.frame(sigma_tau = sigma_grid, I = I_grid)
plot1_pts   <- data.frame(
  sigma_tau = guesses,
  I         = guess_I,
  label     = c("guess 0.05", "guess 0.10", "guess 0.15",
                "calibrated 0.199", "guess 0.25", "guess 0.30")
)
```

```{r plot1, fig.cap = "Realized informativeness as a function of the assumed sigma_tau, holding GM(observed se^2) = 0.0922 fixed. The dashed horizontal at I = 0.30 is the calibration target; the calibrated point sigma_tau = 0.199 sits exactly at the crossing. A guess of sigma_tau = 0.10 lands at I ~ 0.10 — a working lower bound; a guess of sigma_tau = 0.30 overshoots to I ~ 0.49. The curve is steepest in the [0.10, 0.30] band, so a small miss on sigma_tau in this region produces a large miss on I — the empirical case for back-calculating rather than guessing.", fig.alt = "Curve of realized I against assumed sigma_tau, with six labeled candidate points and a dashed line at I = 0.30 marking the calibration target."}
ggplot2::ggplot(plot1_curve, ggplot2::aes(sigma_tau, I)) +
  ggplot2::geom_line(linewidth = 0.9, color = "grey25") +
  ggplot2::geom_hline(yintercept = 0.30, linetype = "dashed",
                      color = "grey45") +
  ggplot2::geom_point(
    data = plot1_pts,
    ggplot2::aes(color = label == "calibrated 0.199"),
    size = 3
  ) +
  ggplot2::geom_text(
    data = plot1_pts,
    ggplot2::aes(label = sprintf("%.2f", I)),
    vjust = -1.0, size = 3.2
  ) +
  ggplot2::scale_color_manual(values = c("grey55", "steelblue4"),
                              guide = "none") +
  ggplot2::labs(
    x        = expression(sigma[tau]),
    y        = expression(hat(I)),
    title    = "Calibration curve: I as a function of sigma_tau",
    subtitle = "Dashed line is the I = 0.30 target; calibrated point sits at the crossing (GM(se^2) fixed)"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

What to read: the calibrated point on the dashed line confirms the
algebra; the steepness of the curve in the target band is the case
for back-calculation; and the spread of the surrounding guesses
($I = 0.10$ at $\sigma_\tau = 0.10$, $I = 0.49$ at $\sigma_\tau =
0.30$) is the empirical cost of guessing wrong.

### Plot 2 — target overlay of observed and simulated $\widehat{se}^2$

The second plot answers the question, "does the simulation's
per-site precision distribution actually match the observed one?"
Layer the observed sampling variances over the simulated ones; if
the calibration is honest, the two distributions overlap near the
geometric mean, and the simulation's tail behavior is consistent with
the source.

```{r plot2-data, include = FALSE}
obs_df <- data.frame(
  source = "observed",
  se2    = observed_se2
)
sim_df <- data.frame(
  source = "simulated",
  se2    = dat$se2_j
)
plot2_df <- rbind(obs_df, sim_df)
gm_obs   <- exp(mean(log(observed_se2)))
gm_sim   <- exp(mean(log(dat$se2_j)))
gm_lines <- data.frame(
  source = c("observed", "simulated"),
  gm     = c(gm_obs, gm_sim)
)
```

```{r plot2, fig.cap = "Per-site sampling variance se^2 — observed (n = 6 sites) vs simulated (J = 30 sites at the calibrated design). Vertical lines mark the geometric mean of each distribution; the simulated GM(se^2) = 0.117 sits close to the observed GM(se^2) = 0.092, consistent with the calibration. The simulated distribution carries a longer right tail because the preset's nj_mean = 40, cv = 0.5 site-size distribution generates a wider precision range than the six pilot sites — the design choice to preserve heterogeneity over a tighter fit. A perfectly matched calibration would tighten the simulated distribution by raising nj_min or lowering cv at the constructor.", fig.alt = "Two density curves of per-site sampling variance, one for observed and one for simulated, with vertical reference lines at the geometric means."}
ggplot2::ggplot(plot2_df, ggplot2::aes(se2, color = source, fill = source)) +
  ggplot2::geom_density(alpha = 0.18, linewidth = 0.9, adjust = 1.4) +
  ggplot2::geom_rug(alpha = 0.6) +
  ggplot2::geom_vline(
    data = gm_lines,
    ggplot2::aes(xintercept = gm, color = source),
    linetype = "dashed", linewidth = 0.7
  ) +
  ggplot2::scale_color_manual(values = c(observed = "steelblue4",
                                         simulated = "darkorange3")) +
  ggplot2::scale_fill_manual(values = c(observed = "steelblue4",
                                        simulated = "darkorange3")) +
  ggplot2::labs(
    x        = expression(widehat(se)^2),
    y        = "Density",
    title    = "Target overlay: observed vs simulated per-site precision",
    subtitle = "Vertical lines are the geometric means; rug is the raw site-level se^2"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

What to read: distance between the two dashed lines is the calibration
miss on $\mathrm{GM}(\widehat{se}^2)$ — small distance means the
informativeness identity holds in the simulation as it did in the
observed data. Spread of the orange (simulated) density relative to
the blue (observed) rug is the heterogeneity premium charged by the
preset's site-size distribution; tightening this is a Layer 2
adjustment ($\bar{n}_j$, `cv`, `nj_min`) covered in
[M3 · Margin and SE models](m3-margin-se-models.html).

## 5. Audit a calibrated grid

A single calibrated design is one row; a real simulation campaign
covers a grid. Use the *direct-precision* path
([`design_grid(paradigm = "direct", ...)`](../reference/design_grid.html))
when the design is to be defined in $(I, R)$ space directly, holding
$\sigma_\tau$ and the G shape fixed across cells. The
direct paradigm sidesteps the site-size machinery and lets the
audit answer "for which $(I, R, J)$ combinations does the rest of the
diagnostic ladder pass?" before any simulation effort is spent.

```{r grid}
direct_grid <- design_grid(
  paradigm  = "direct",
  I         = c(0.20, 0.40),
  R         = c(1, 3),
  J         = c(14L, 30L),
  sigma_tau = 0.20,
  true_dist = "Gaussian",
  seed_root = 20260507L
)
direct_grid
```

Eight cells, one per $(I, R, J)$ combination. Each row carries a
`design` list-column with the per-cell `multisitedgp_design` and a
deterministic `seed` derived from `seed_root`. Now the audit at
$M = 1$ replicate per cell — a *smoke audit* to discover which cells
fail any gate before committing to a full $M = 50$ or $M = 200$ run:

```{r audit}
audit <- scenario_audit(direct_grid, M = 1L, parallel = FALSE)
audit[, c("cell_id", "J", "I", "R", "pass", "fail_reasons")]
```

Read down the `fail_reasons` column. The four cells at $I = 0.20$ all
fail `mean_shrinkage:median<0.300` — at this informativeness, the
partial-pooling fraction sits below the Group D floor regardless of
$J$ or $R$. The four cells at $I = 0.40$ all fail
`bhattacharyya:median<0.850` and `ks:median>0.100` — Group C is firing
because the direct paradigm's Gaussian declaration is being audited
against a power-bound G-shape gate at $M = 1$ (the Group C numbers
stabilize as $M$ grows; at $M = 50$ many of these cells clear). The
audit's job at $M = 1$ is not to make the final go/no-go call — it is
to surface where the real audit at $M = 50$ should focus. None of the
eight cells passes, so the smoke audit's recommendation is to revisit
the calibration target: either lift $I$ above the partial-pooling
floor ($I \ge 0.30$ with this $\sigma_\tau$) or accept a Group D WARN
and proceed.

For an end-to-end campaign that runs the full audit, see
[A6 · Case study — multisite trial](a6-case-study-multisite.html). For
the diagnostic rubric the audit is reading off, see
[A3 · Diagnostics in practice](a3-diagnostics-in-practice.html).

## 6. Why is this defensible?

The case for back-calculating $\sigma_\tau$ rather than guessing it
rests on three properties of the workflow.

First, *every calibrated quantity is either observed or chosen
explicitly*. $\mathrm{GM}(\widehat{se}^2)$ is computed from the source
study's reported per-site sampling variances — it is not a parameter
to be estimated, it is an arithmetic summary. $I$ is the researcher's
calibration target, declared up front and defensible against the
public bounds in the empirical-Bayes literature [@walters2024empirical].
The only quantity that flows out of the algebra — $\sigma_\tau$ — is
exactly the quantity the researcher had no defensible way to guess
directly. The workflow trades a guess for a derivation.

Second, *the algebraic step is invertible*, and the diagnostic ladder
enforces the round-trip. [`compute_I()`](../reference/compute_I.html)
takes $\widehat{se}^2$ and $\sigma_\tau$ and returns $I$; the
back-calculation is its inverse on the third variable. The §2
`check_I` column confirms the round-trip on observed inputs; the §3
Group A row confirms it on simulated inputs (realized $\sigma_\tau =
0.200$ against target $0.199$, PASS). When either fails, the
calibration has a coding error, not a methodological gap.

Third, *the resulting design is auditable via canonical_hash*. The §7
manifest freezes the audit's hash (`f305c1bb926e3d2a`) and the
simulated data's hash (`5eac731a6c96e9d1`); a reader who re-runs the
code at the same package version and seed reproduces both. Every
input either appears in the back-calc chunk (observed
$\widehat{se}^2$, target $I$) or in the override chunk (`J`,
`true_dist`, `theta_G`) — nothing is hidden in the prose.

Three caveats. The workflow does not estimate $\sigma_\tau$ from data
— there is no inference, no posterior, no likelihood. It does not
require the researcher's $I$ target to match any specific historical
estimate; a sensitivity sweep across $I \in \{0.20, 0.30, 0.40\}$ is
the honest accompaniment when the target is uncertain. And the
calibration is honest only to the precision distribution of the
source study — if the source is small or unrepresentative, the
calibration inherits that limitation [@weiss2017howmuch]. None of
these undermines the core claim: at fixed $\mathrm{GM}(\widehat{se}^2)$
and fixed target $I$, the back-calculated $\sigma_\tau$ is the unique
value that makes the simulation's informativeness identity hold.

## 7. Freeze provenance

The two hashes that anchor the calibration to a reproducible artifact
are `canonical_hash(dat)` for the simulated data and
`canonical_hash(audit)` for the smoke audit. The
[`provenance_string()`](../reference/provenance_string.html) helper
returns a single-line summary suitable for a methods footnote.

```{r provenance}
manifest <- list(
  project       = "calibration-A5-vignette",
  source_se2    = observed_se2,
  target_I      = target_i,
  GM_se2_obs    = gm_se2,
  sigma_tau     = sigma_tau,
  J             = 30L,
  preset        = "preset_jebs_paper",
  data_hash     = canonical_hash(dat),
  n_audit_cells = nrow(direct_grid),
  audit_M       = 1L,
  audit_hash    = canonical_hash(audit)
)
manifest
```

```{r provenance-string}
provenance_string(dat)
```

The `manifest` list is a self-documenting record: source per-site
sampling variances, target informativeness, back-calculated
$\sigma_\tau$, the calibrated design's data hash, and the audit's
hash. Saving it as an `.rds` next to the manuscript draft makes the
calibration auditable by a methods reviewer without re-running the
simulation. `provenance_string()` is the same information in a single
line — useful for a figure caption or footnote. The two hashes —
`5eac731a6c96e9d1` for the simulated data and `f305c1bb926e3d2a` for
the audit — are deterministic at this package version, this seed, and
this source `observed_se2`; they are the calibration's contract with
the reader.

For a deeper treatment of the manifest pattern, the determinism
contract, and the relationship between `canonical_hash()` and the
package version, see
[Reproducibility and provenance](m7-reproducibility-provenance.html).

## 8. Where to next

- [A2 · Choosing a preset](a2-choosing-a-preset.html) — the preset
  shelf the calibration starts from; pick the preset whose Layer 2
  defaults best match your source study's site-size distribution.
- [A3 · Diagnostics in practice](a3-diagnostics-in-practice.html) —
  the four-question rubric (the Group A scale check is the calibration's
  primary contract; the §3 walkthrough above reads off Group A directly).
- [A6 · Case study — multisite trial](a6-case-study-multisite.html) —
  the next step after calibration: take the calibrated design, run the
  full audit at $M = 50$, drop the failing cells, and run the
  surviving cells through the analyst's pipeline.
- [M3 · Margin and SE models](m3-margin-se-models.html) — the formal
  definitions of $I$, $R$, and the geometric-mean identity, including
  the relationship between Layer 2 site-size choices and the realized
  $\mathrm{GM}(\widehat{se}^2)$.
- [Reproducibility and provenance](m7-reproducibility-provenance.html)
  — the manifest pattern, the determinism contract, and the
  relationship between `canonical_hash()` and the package version.

## References

The informativeness identity $I = \sigma_\tau^2 / (\sigma_\tau^2 +
\mathrm{GM}(\widehat{se}^2))$ and the calibrated-design rationale are
laid out in @lee2025improving. The empirical-Bayes context for the
calibration's Group D consequences is collected in
@walters2024empirical. The education-trial calibration thresholds —
the $I \in [0.10, 0.50]$ working band that frames the §1 case for
calibrating rather than guessing — are reviewed in @weiss2017howmuch.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
