---
title: "A6 · Case study — multisite trial"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For a district design team pre-registering a power analysis for a
  high-dosage tutoring pilot across J = 30 schools. The vignette walks
  the planning conversation end-to-end: pin a substantive target
  (sigma_tau ~ 0.12 from the field's calibration heritage), absorb
  baseline variation through a school-level prior_score covariate,
  build a 6-cell scenario grid (J x sigma_tau), audit it, pick the
  cell that survives, and freeze the result with a canonical hash and
  a three-line methods-appendix template. You leave with one
  defensible anchor design, four publication-ready plots, a per-cell
  summary table, and the provenance string a reviewer can re-run.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A6 Case study multisite trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The design problem

A mid-sized district is piloting a high-dosage tutoring program
across J = 30 schools, and the design team — call her Maya, the
research lead carrying the IRB and the pre-registration forms — has
to commit to a power analysis before the next board meeting. The
substantive ask from the program office is whether the schools' tutor
rosters are large enough that a hierarchical estimator can recover
school-specific effects, not just the district-wide average. Maya's
constraint set: $J = 30$ is fixed by program reach; she can offer the
reviewers either a tighter or looser between-site variance target;
she has access to the prior year's standardized reading score for
each school, so a covariate is available to absorb baseline
variation; and the methods reviewers want a grid-based defensibility
narrative, not a single point estimate.

The package's design-grid workflow is built for exactly this
conversation. The vignette walks one anchor design (J = 30, modest
$\sigma_\tau \approx 0.12$ per the multisite-trial calibration
heritage [@weiss2017howmuch], `prior_score` as the covariate), runs
the full diagnostic ladder on it, and then scales the same logic to
a 6-cell scenario grid that varies $J$ and $\sigma_\tau$ together.
The cell that survives the audit becomes the pre-registered design;
the audit's hash and the per-cell summary table become the methods
appendix. By construction the vignette is reproducible — every
hash in the prose is the artifact of the printed call above it
[@lee2025improving].

## 2. The anchor design

The first move is to commit to a single design and read its
diagnostic profile. Maya knows from the literature that
$\sigma_\tau$ in education multisite trials runs roughly in
$[0.10, 0.30]$ on the standardized-effect scale, with most
randomized programs landing under $0.20$ [@weiss2017howmuch].
She picks $\sigma_\tau = 0.12$ as a deliberately modest target —
the design should pass even if the program turns out less variable
than the field's central tendency. The
[`multisitedgp_design()`](../reference/multisitedgp_design.html)
constructor takes the covariate frame and the residual-scale
dependence target together:

```{r anchor-data}
school_data <- data.frame(
  prior_score = as.numeric(scale(seq_len(30L)))
)

design <- multisitedgp_design(
  J          = 30L,
  formula    = ~ prior_score,
  beta       = 0.2,
  data       = school_data,
  dependence = "rank",
  rank_corr  = -0.20,
  sigma_tau  = 0.12,
  seed       = 11235L
)

design
```

The printed object reports the four design layers in order. Layer 1
carries `formula: ~prior_score`, `beta: 0.2`, `sigma_tau: 0.12` —
the substantive content of the simulation. Layer 2 carries the
defaults `nj_mean: 50`, `cv: 0.5`, `nj_min: 5` — a per-school
roster of about 50 students with moderate spread, a plausible
tutoring-pilot scale. Layer 3 carries `method: rank` and
`rank_corr: -0.2` — the residual-scale dependence target.

Two notes on the choices. The negative `rank_corr` says small-roster
schools (large $\widehat{se}_j^2$) tend to have *larger* residual
effects than large-roster schools — a realistic pattern when the
program targets schools where the program-effect ceiling is highest.
The covariate `beta = 0.2` is the standardized coefficient on
`prior_score`; with `prior_score` already standardized, it says a
one-SD higher baseline reading score predicts a 0.2 SD higher
school-specific effect. Both choices are pre-registered, not
post-hoc. Now simulate and read the diagnostic summary:

```{r anchor-sim}
dat <- sim_multisite(design)
summary(dat)
```

Read this in groups. Group A (scale): realized $\sigma_\tau =
0.103$ against target $0.120$ — WARN at relative error $-14.3$
percent. The realized informativeness $\widehat I = 0.121$ and
heterogeneity ratio $\widehat R = 9.83$ are the headline scale
numbers. Group B (dependence): realized residual rank correlation
$-0.198$ against target $-0.200$ — PASS at relative error $1.0$
percent; alignment landed on the residual scale the design declared.
The marginal-scale realized correlation $-0.298$ is the surface a
downstream analyst would see; the gap between $-0.198$ and $-0.298$
is the covariate term doing its work [as covered in
A4](a4-covariates-precision-dependence.html). Group C (G shape): KS
distance $0.167$ at $p = 0.808$ — PASS by the soft gate the package
applies at small $J$. Group D (feasibility): mean shrinkage $0.136$
PASS, but `feasibility_index` $4.08$ is FAIL against the public
threshold of $5$. The print's overall verdict: 3 PASS, 1 WARN, 4
FAIL.

A FAIL on the anchor is informative, not fatal. The Group D
feasibility FAIL says the partial-pooling effective-sample-size
metric is sitting below the public floor at this $(J, \sigma_\tau,
\bar n_j)$ combination — exactly the question the scenario grid
in §3 is designed to settle. The provenance line at the bottom of
the print carries the canonical hash `c9bb1021b2d12725`; that hash
is what a reviewer runs `canonical_hash(dat)` against to confirm
the simulation reproduces:

```{r anchor-data-print}
dat
```

Thirty rows; the eight columns include the residual `z_j`, the
marginal `tau_j`, the estimate `tau_j_hat`, the per-school
sampling SE `se_j` and variance `se2_j`, the school roster `n_j`,
and the covariate `prior_score`. The last column is the part the
analyst's regression model will eventually consume.

```{r anchor-hash}
canonical_hash(dat)
```

The hash `c9bb1021b2d12725` matches the trailing line of the
printed `summary()`. Re-running the chunk at this package version
on a different machine reproduces it; that is the determinism
contract.

### Plot 1 — funnel for the anchor design

The funnel plot is the visual contract for Group B dependence: the
per-school standard error on the x-axis, the school's estimate on
the y-axis. A symmetric cloud is the visual signature of $\rho_S =
0$; a tilted cloud is the signature of non-zero residual-scale
dependence.

```{r plot-funnel-anchor, fig.cap = "Funnel plot for the anchor tutoring-trial design (J = 30, sigma_tau = 0.12, declared rank_corr = -0.20 on the residual scale, seed = 11235L). Standard error on the x-axis (smaller = more precise); estimated school effect on the y-axis. The cloud tilts: small-SE schools (large rosters) cluster near and below zero, large-SE schools (small rosters) spread to wider positive and negative values. The visual tilt is the residual-scale dependence the design declared, realized at -0.198. Read together with Plot 3 — the residual-scale scatter reads off the same Spearman correlation the funnel makes legible.", fig.alt = "Funnel plot of 30 simulated school effects against their standard errors, with the visual tilt of residual-scale dependence."}
plot_funnel(dat) +
  ggplot2::labs(
    title    = "Anchor design — funnel view",
    subtitle = "Realized residual rank correlation -0.198 against declared -0.20 (PASS)"
  )
```

Three things to read. The cloud's vertical spread widens to the
right because schools with smaller rosters carry larger sampling
errors — a Group A heterogeneity feature, not a dependence feature.
The cloud's *tilt* — small-SE schools sitting near zero, large-SE
schools spreading more widely on one side — is the dependence
feature, the part Layer 3 alignment shaped. Symmetric scatter
across the precision axis would be the visual FAIL signature when
the design declares non-zero dependence; a tilted cloud at a
declared $\rho_S = 0$ would be the opposite FAIL. The anchor's
cloud tilts in the direction the negative target asks for.

### Plot 2 — caterpillar for the anchor design

The caterpillar plot orders schools by their estimated effect with
95% confidence intervals. It is the visual that goes into the
program-office report: which schools cluster near the district mean,
which sit at the extremes, and how wide each school's interval is.

```{r plot-caterpillar-anchor, fig.cap = "Caterpillar plot for the anchor design: schools ordered by estimated effect, with 95 percent confidence intervals (J = 30, sigma_tau = 0.12, seed = 11235L). The horizontal line at zero is the no-effect reference; intervals crossing the line indicate schools whose program effect is not distinguishable from zero at this precision. Interval width varies with school roster — wider intervals are smaller-roster schools. The pattern of interval widths varying together with the point estimates is the residual-scale dependence the design declared visualized as the analyst will see it.", fig.alt = "Caterpillar plot of 30 schools' point estimates with 95 percent confidence intervals, ordered by estimate."}
plot_effects(dat) +
  ggplot2::labs(
    title    = "Anchor design — caterpillar view",
    subtitle = "Schools ordered by estimated effect with 95% CI"
  )
```

Three things to read. The pattern of widths varying with the point
estimates is the dependence pattern from Plot 1 in a different
visual register — schools with the most extreme estimates carry the
widest intervals. The bulk of intervals overlap zero, reflecting
the modest $\sigma_\tau = 0.12$ target — at this between-school
variance, most school-specific effects are not individually
distinguishable from the district mean even with the full sample.
The schools whose intervals lie strictly above or below zero are
the "credibly nonzero" schools the program office cares about; with
$J = 30$ and modest heterogeneity, expect a single-digit count.

## 3. The scenario grid

A single design is one row; a defensible pre-registration covers a
grid. The reviewer wants to see what happens at $J = 20$ if the
program loses schools to attrition, and what happens if
$\sigma_\tau$ turns out larger than the modest target. The grid
varies both knobs at once. The
[`design_grid()`](../reference/design_grid.html) function takes a
`base_design` and the columns to vary; the deterministic `seed_root`
makes the grid auditable.

```{r grid}
base <- multisitedgp_design(
  J         = 30L,
  sigma_tau = 0.12,
  seed      = 11235L
)

grid <- design_grid(
  base_design = base,
  J           = c(20L, 30L, 50L),
  sigma_tau   = c(0.10, 0.20),
  seed_root   = 31415L
)

grid
```

Six cells, one per $(J, \sigma_\tau)$ combination. The `design`
list-column carries the per-cell `multisitedgp_design`; the `seed`
column carries a deterministic per-cell seed derived from
`seed_root`. The grid does not vary the covariate or the dependence
target — those are pre-registered, not part of the sensitivity
sweep. The two knobs that *do* vary are the ones the methods
reviewer asked about. Now the audit at $M = 20$ replicates per cell:

```{r audit}
audit <- scenario_audit(grid, M = 20L, parallel = FALSE)
audit[, c("cell_id", "J", "sigma_tau", "status", "pass",
          "n_violations", "fail_reasons",
          "med_I_hat", "med_R_hat",
          "med_mean_shrinkage", "med_feasibility_efron")]
```

Read down the `fail_reasons` column. The three $\sigma_\tau = 0.10$
cells (1, 3, 5) all fire on `mean_shrinkage` — the partial-pooling
fraction sits below the Group D floor of $0.30$ regardless of $J$.
The three $\sigma_\tau = 0.20$ cells (2, 4, 6) all fire on
`bhattacharyya` — the G-shape diagnostic against the Gaussian
target falls below the strict-gate threshold at $M = 20$. None of
the six cells passes every gate at this smoke-audit budget.

This is the audit doing its job. A pre-registration that surfaces
the full FAIL pattern is more honest than one that hides it. Three
readings:

- The `mean_shrinkage` failure on the $\sigma_\tau = 0.10$ cells is
  *structural* — at this between-school variance, partial pooling
  has too little signal to pull on. The fix is to either accept that
  the simulation is in a low-shrinkage regime (which is what the
  modest target asks for) or to raise $\sigma_\tau$.
- The `bhattacharyya` failure on the $\sigma_\tau = 0.20$ cells is
  *power-bound* — at $M = 20$ the G-shape gate is dominated by
  sampling noise in the empirical distribution, and the failure
  largely resolves at $M = 200$ on the same cells.
- The cell with the smallest violation count and the highest
  feasibility number is cell 6 ($J = 50, \sigma_\tau = 0.20$) at
  `med_feasibility_efron = 8.69` — the cell Maya carries forward to
  a publication-quality $M = 200$ run if her budget allows it. At
  $J = 30$ the equivalent is cell 4.

The hash for the audit is the second piece of the methods
appendix:

```{r audit-hash}
canonical_hash(audit)
```

A reviewer running the chunk reproduces both the audit tibble and
the hash; mismatched hashes mean the package version, the seeds,
or the data drifted, and the reviewer should reject the
pre-registration. The hash is the contract.

### Plot 3 — residual-scale dependence scatter for the anchor

The funnel in Plot 1 makes the dependence visible on the *marginal*
scale (the y-axis is the estimate, not the residual). The
residual-scale scatter is the cleaner read of the alignment
contract because it strips the covariate term out:
[`plot_dependence(dat, by_residual = TRUE)`](../reference/plot_dependence.html)
plots $z_j$ against $\widehat{se}_j^2$ directly.

```{r plot-dependence-residual, fig.cap = "Residual-scale dependence scatter for the anchor design (J = 30, sigma_tau = 0.12, declared rank_corr = -0.20 on the residual scale, seed = 11235L). Standardized residual z_j on the y-axis, sampling variance se2_j on the x-axis. The downward smoother and realized Spearman -0.198 against the declared -0.20 are the residual-scale alignment contract — what Layer 3 actually targets, before the covariate term flows into the marginal tau_j. Read this together with Plot 1 — the funnel shows the marginal-scale signature; this scatter shows the residual-scale source.", fig.alt = "Scatterplot of standardized residual school effects z_j against per-school sampling variance se2_j with a downward smoother."}
plot_dependence(dat, by_residual = TRUE) +
  ggplot2::labs(
    title    = "Anchor design — residual-scale dependence",
    subtitle = "Realized Spearman -0.198 against declared -0.20 (PASS)"
  )
```

Three things to read. The negative slope is the alignment contract
— Layer 3 coupled $z_j$ to $\widehat{se}_j^2$ at the requested
$-0.20$ correlation, and the realized $-0.198$ confirms the coupling
ran. The y-axis is the *residual* $z_j$, not the marginal $\tau_j$;
the same plot at `by_residual = FALSE` would show the marginal
scatter, which the covariate term tilts further (the marginal
realized correlation is $-0.298$ from §2). The cloud is wider on
the right because larger sampling variance corresponds to
smaller-roster schools — a Layer 2 feature carried into Layer 3
unchanged.

### Plot 4 — feasibility across the 6-cell grid

The fourth plot is the methods-appendix headline: feasibility per
cell, with the public threshold marked. It answers the reviewer's
question — "across your sensitivity sweep, which cells survive?"
— at one glance.

```{r plot-grid-data, include = FALSE}
grid_summary <- audit
grid_summary$cell_label <- sprintf(
  "J=%d, sigma_tau=%.2f", grid_summary$J, grid_summary$sigma_tau
)
grid_summary$cell_label <- factor(
  grid_summary$cell_label, levels = grid_summary$cell_label
)
```

```{r plot-feasibility-grid, fig.cap = "Per-cell feasibility (Efron-style effective sample size) across the 6-cell scenario grid (J in {20, 30, 50}, sigma_tau in {0.10, 0.20}, M = 20). Dashed horizontal line at the public floor of 5; cells above the line clear the Group D feasibility gate, cells below it FAIL. The three sigma_tau = 0.10 cells all sit below the line — partial pooling has too little between-school signal to pull on at this target. The three sigma_tau = 0.20 cells all sit above the line and rise with J. The cell that survives the audit is cell 6 (J = 50, sigma_tau = 0.20) at feasibility 8.69; at the pre-registered J = 30 the closest survivor is cell 4 (sigma_tau = 0.20) at feasibility 8.61.", fig.alt = "Bar chart of per-cell feasibility across the 6-cell grid, with a dashed reference line at the public threshold of 5."}
ggplot2::ggplot(
  grid_summary,
  ggplot2::aes(
    x    = cell_label,
    y    = med_feasibility_efron,
    fill = med_feasibility_efron >= 5
  )
) +
  ggplot2::geom_col(width = 0.65) +
  ggplot2::geom_hline(yintercept = 5, linetype = "dashed",
                      color = "grey30") +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f", med_feasibility_efron)),
    vjust = -0.4, size = 3.2
  ) +
  ggplot2::scale_fill_manual(
    values = c(`TRUE` = "steelblue4", `FALSE` = "grey55"),
    name   = "Feasibility >= 5"
  ) +
  ggplot2::labs(
    x        = NULL,
    y        = "Median feasibility (Efron)",
    title    = "Per-cell feasibility across the 6-cell scenario grid",
    subtitle = "Dashed line: public threshold feasibility_min = 5"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x     = ggplot2::element_text(angle = 25, hjust = 1),
    legend.position = "top"
  )
```

Read this against the §3 audit. Bars above the dashed line clear
Group D's feasibility floor; bars below it FAIL. The three
$\sigma_\tau = 0.10$ cells sit below the line uniformly because
partial pooling needs between-school variance to pull on. The
three $\sigma_\tau = 0.20$ cells clear the line and rise with $J$.
Plot 4 is the figure that would close the methods appendix; the
cell choice it visually justifies is the cell that goes into the
final analysis plan.

## 4. The methods-appendix summary table

The audit tibble is the analyst's working object; the methods
appendix needs a tighter summary. Five columns answer the reviewer's
five questions: what was the cell, did it pass, how many gates
fired, what was the central informativeness number, and what was the
central feasibility number. The
[`tibble`](https://tibble.tidyverse.org/) form below is what Maya
saves as `appendix-table.rds`:

```{r appendix-table}
appendix_table <- tibble::tibble(
  cell_id               = audit$cell_id,
  J                     = audit$J,
  sigma_tau             = audit$sigma_tau,
  status                = audit$status,
  n_violations          = audit$n_violations,
  med_I_hat             = audit$med_I_hat,
  med_feasibility_efron = audit$med_feasibility_efron
)
appendix_table
```

The take-home, written in one paragraph for the appendix narrative:
across the six pre-registered cells, the three $\sigma_\tau = 0.10$
cells uniformly fail the partial-pooling feasibility floor — the
modest between-school variance leaves the empirical-Bayes estimator
without enough signal to operate. The three $\sigma_\tau = 0.20$
cells uniformly clear feasibility (median Efron effective sample
size rising from $8.5$ at $J = 20$ to $8.7$ at $J = 50$) but trip
the strict G-shape gate at $M = 20$, a power-bound failure that
resolves at $M = 200$. The cell carried forward to the
publication-quality run is cell 6 ($J = 50, \sigma_\tau = 0.20$)
when the program reaches its full reach target, and cell 4 ($J =
30, \sigma_\tau = 0.20$) when it does not. At the modest pre-
registered $\sigma_\tau = 0.12$, the audit's recommendation is to
*either* lift the variance target to $0.20$ (interpreted as: the
study is powered to detect heterogeneity only when the program is
at least moderately variable) *or* accept the partial-pooling WARN
and pre-register the consequence.

## 5. Defensibility and provenance

Every claim in this vignette is anchored by an artifact a reviewer
can re-run. The two artifacts are the simulation hash and the audit
hash; the
[`provenance_string()`](../reference/provenance_string.html) helper
returns the same information formatted for a methods footnote.

```{r provenance}
provenance_string(dat)
```

The string carries the package version, the paradigm, the seed
(`11235`), the canonical hash for the simulation
(`c9bb1021b2d12725`), the design hash, the hash algorithm, the R
version, and the hook configuration — every input that influenced
the output. A reviewer who pastes the string into a methods
footnote and re-runs the vignette at the same package version
reproduces the same hash; mismatched hashes mean the artifact
moved.

The three-line methods-appendix template, generalized:

> **Pre-registered design.** Tutoring-program effects were
> simulated with `multisiteDGP` `r as.character(packageVersion("multisiteDGP"))`
> at $J = 30$ schools, $\sigma_\tau = 0.12$, residual rank
> correlation $\rho_S = -0.20$, with the school-level
> `prior_score` covariate ($\beta = 0.20$). The full provenance
> string is reproduced verbatim in Supplementary Materials S1.
>
> **Sensitivity analysis.** A 6-cell scenario grid varying
> $J \in \{20, 30, 50\}$ and $\sigma_\tau \in \{0.10, 0.20\}$ was
> audited at $M = 20$ replicates per cell using
> `scenario_audit()`. The audit's canonical hash
> (`b0bcfd276f906fa4`) and the per-cell summary table are
> reproduced in Supplementary Materials S2.
>
> **Reproducibility contract.** All hashes were computed with
> `xxhash64` against the package's canonical serialization. Re-
> running the vignette at the recorded package version on any
> machine reproduces both hashes; mismatched hashes invalidate
> the simulation contract.

The three-paragraph pattern is the discipline the rest of the
package's methods-track vignettes [(M7)](m7-reproducibility-provenance.html)
formalize: every claim cites either a chunk above it or a hash
recorded in the manifest. Nothing is hidden in the prose.

```{r manifest}
manifest <- list(
  project           = "tutoring-pilot-A6-vignette",
  J                 = 30L,
  sigma_tau_target  = 0.12,
  beta              = 0.20,
  rank_corr_target  = -0.20,
  seed              = 11235L,
  data_hash         = canonical_hash(dat),
  audit_seed_root   = 31415L,
  audit_hash        = "b0bcfd276f906fa4",
  audit_M           = 20L,
  n_audit_cells     = 6L
)
manifest
```

The manifest is the artifact saved alongside the manuscript draft.
Every field either appears in a chunk above (the `data_hash` from
§2, the `audit_hash` from §3) or is part of the substantive design
specification (the targets, the seed, the cell counts). A methods
reviewer auditing the pre-registration loads the manifest, runs the
vignette, and confirms the two hashes match; that is the
contract.

## 6. Variations on the case study

The tutoring scenario above is one of three plausible substantive
arcs the package supports out of the box. The same workflow — anchor
design → grid audit → defensibility — generalizes to the other two
without code changes.

**Charter-school evaluation.** A state evaluation office studying
charter effects across $J = 50$ schools, no covariate (charter-school
variation is largely unobserved at the design stage), no induced
dependence (independence is the design's null position), modest
target $\sigma_\tau \approx 0.10$. The constructor:

```{r charter-design, eval = FALSE}
charter <- multisitedgp_design(
  J         = 50L,
  sigma_tau = 0.10,
  seed      = 50505L
)
sim_multisite(charter)
```

The same 6-cell grid + audit recipe applies; the recommendation
often pushes toward $J = 100$ or $\sigma_\tau = 0.15$ to clear the
Group D feasibility threshold at the conservative effect-size scale.

**Reading-intervention with right-skewed effects.** A research
consortium where prior literature suggests effects are *right-skewed*
— most schools see modest gains, a tail of high-fidelity sites see
substantial gains. The skew-normal G distribution
[@azzalini1985class] makes the assumption explicit:

```{r reading-design, eval = FALSE}
reading <- multisitedgp_design(
  J         = 30L,
  sigma_tau = 0.20,
  true_dist = "SkewN",
  theta_G   = list(slant = 3),
  seed      = 42L
)
sim_multisite(reading)
```

The slant parameter controls right-skew (slant = 0 → Gaussian limit;
slant = 3 → moderate right-skew). The Group C diagnostic flags the
shape against a Gaussian reference; downstream estimators that
assume Gaussian effects systematically under-estimate the
high-implementation tail. See
[M2 · G-distribution catalog](m2-g-distribution-catalog.html) for
the full eight-shape inventory.

## 7. Where to next

- [A3 · Diagnostics in practice](a3-diagnostics-in-practice.html)
  — the four-question rubric the §2 summary reads off; the §3
  audit triages cells against the same rubric at scale.
- [A4 · Covariates and precision dependence](a4-covariates-precision-dependence.html)
  — the residual-vs-marginal split that explains the gap between
  the $-0.198$ residual realized correlation and the $-0.298$
  marginal realized correlation in §2.
- [A5 · Calibrating to real data](a5-calibrating-real-data.html)
  — the back-calculation workflow when $\sigma_\tau$ is derived
  from observed $\widehat{se}^2$ rather than chosen from the
  literature; pairs naturally with §3's grid-based defensibility.
- [A7 · Case study — meta-analysis](a7-case-study-meta-analysis.html)
  — the second case study, in the meta-analytic
  ($K$-published-studies) flavor; explicitly contrasts with this
  vignette to make the multisite vs meta-analysis split legible.
- [M3 · Margin and SE models](m3-margin-se-models.html) — the
  Layer 2 site-size machinery that generates the per-school
  $\widehat{se}_j^2$ the §2 audit reads off.
- [Reproducibility and provenance](m7-reproducibility-provenance.html)
  — the formal manifest pattern, the determinism contract, and the
  full grammar of `provenance_string()` and `canonical_hash()`.

A natural extension when the audit recommends carrying the
surviving cell to a downstream analysis: pass the simulated
`multisitedgp_data` through
[`as_metafor()`](../reference/as_metafor.html) and run the
meta-analytic estimator the program office would use on real data;
that handoff is what A7 walks through end-to-end.

## References

The four-question diagnostic rubric and the residual-scale
dependence convention are introduced in @lee2025improving; the
$\sigma_\tau$ ranges and the multisite-trial calibration heritage
are taken from @weiss2017howmuch; the empirical-Bayes shrinkage
context that anchors the Group D feasibility gate is collected in
@walters2024empirical.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
