---
title: "A7 · Case study — meta-analysis"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For a meta-analyst preparing a small synthesis of K = 12 published
  studies that report effect-size estimates and standard errors but no
  per-study sample sizes. The vignette walks the planning conversation
  end-to-end on the direct-precision path: declare the precision profile
  through (I, R) instead of through site rosters, simulate with
  sim_meta(), read the four diagnostic groups, layer in a
  precision-effect coupling that mimics the small-study selection
  signature, hand the simulated data off to metafor for the random-effects
  estimate, and freeze the result with a canonical hash and a three-line
  methods-appendix template. Sibling vignette A6 walks the same workflow
  on the site-size-driven path; the contrast is the headline.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A7 Case study meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",   # never "hide" for primary calls
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The meta-analytic design problem

A methodologist — call her Priya, the senior author on a synthesis
manuscript — has assembled a small-K meta-analysis of $K = 12$
published studies of a school-based reading intervention. Each
included study reports a standardized effect-size estimate and its
standard error, but most do not report the per-study sample size in
a form the synthesis can use. The pre-registration committee asks for
a simulation that *mirrors the precision profile of the included
studies* — the geometric mean of the sampling variances and the
spread of those variances across studies — without inventing
fictitious per-study $n_j$ values that would have to be defended on
their own. The substantive target is informativeness $I = \rho_S$
roughly $0.30$ (modest but non-trivial signal-to-noise) and a
standard-error-heterogeneity ratio $R = 2$ (the largest study is
twice as precise as the smallest).

The committee also wants the simulation to be honest about the
small-study selection literature. Smaller, less-precise studies in
the published record tend to *under-report* effect sizes — null and
small-effect findings file-drawer at higher rates, and the surviving
small studies that do appear in print often correspond to
larger-effect estimates (a documented pattern in the published
meta-analytic record). Priya's
pre-registration commits to *respecting* that signature — simulating
data that exhibits the same precision-effect coupling the published
record exhibits — rather than correcting it post-hoc. The simulation
target on that scale is a residual rank correlation
$\rho_S \approx 0.30$ between the standardized effect and the per-
study sampling variance.

The package's direct-precision path is built for exactly this
conversation. Sibling vignette
[A6 · Case study — multisite trial](a6-case-study-multisite.html)
walks the *site-size-driven* path (the package's blueprint
abbreviation: the per-school roster $n_j$ is the front-door knob, and
the per-school sampling SE is computed from $n_j$ and a per-pupil
residual variance). The meta-analytic case study switches paradigms:
the front-door knob is the precision pair $(I, R)$, not the per-
study $n_j$, and the package's
[`sim_meta()`](../reference/sim_meta.html) function is the entry
point. The contrast is the headline of this vignette — same
workflow shape (anchor → diagnostics → dependence → adapter →
provenance), different front door.

## 2. Direct-precision specification

The first move is to commit to a single design and read its
diagnostic profile. With no per-study $n_j$ available, Priya
specifies the precision profile directly: $I = 0.30$ on the
informativeness scale, $R = 2$ on the SE-heterogeneity scale,
$\sigma_\tau = 0.20$ as the between-study heterogeneity target. The
[`sim_meta()`](../reference/sim_meta.html) function takes the four
arguments together and returns a `multisitedgp_data` tibble with the
direct-precision provenance attached:

```{r anchor-sim}
dat <- sim_meta(
  J         = 12L,
  I         = 0.30,
  R         = 2,
  sigma_tau = 0.20,
  seed      = 1L
)
dat
```

Twelve rows, one per study. The seven columns include the
standardized residual `z_j`, the marginal effect `tau_j`, the
estimated effect `tau_j_hat`, the per-study sampling SE `se_j` and
sampling variance `se2_j`, the (here `NA`) site-size column `n_j`,
and the integer `site_index`. The print's tibble header carries the
direct-precision summary:
`Realized vs intended` for $I$, $R$, $\sigma_\tau$, $\rho_S$, plus
the operational feasibility verdict. The full diagnostic ladder is
in the `summary()` print:

```{r anchor-summary}
summary(dat)
```

Read this in groups. Group A (scale): realized $I = 0.300$ against
target $0.300$ — PASS at relative error $0.0$ percent; realized
$R = 2.000$ against target $2.000$ — PASS at the same threshold;
realized $\sigma_\tau = 0.162$ against target $0.200$ — FAIL at
relative error $-18.9$ percent; geometric mean of $\widehat{se}_j^2$
$0.093$ against the implied target $0.093$ — PASS. Group B
(dependence): realized residual rank correlation $0.063$ against
target $0.000$ — PASS; the marginal-scale realized correlation
$0.063$ matches because no covariate is in play. Group C (G shape):
KS distance $0.250$ at $p = 0.869$ — PASS by the soft gate; the
Bhattacharyya coefficient $0.333$ FAILs the strict gate at $K = 12$.
Group D (feasibility): mean shrinkage $0.302$ PASS, but
`feasibility_index` $3.624$ FAIL against the public floor of $5$ —
expected at $K = 12$ studies and modest informativeness.

The print's overall verdict is 6 PASS, 1 WARN, 4 FAIL. The
$\sigma_\tau$ FAIL and the feasibility FAIL are *informative, not
fatal*. Both are downstream consequences of the small-K regime the
synthesis lives in: at $K = 12$ the realized between-study
heterogeneity moment estimator has wide sampling distribution, and
the empirical-Bayes feasibility metric reflects the modest pooled
signal. The pre-registration's job is to surface those numbers, not
to hide them. The trailing provenance line carries the canonical
hash:

```{r anchor-hash}
canonical_hash(dat)
```

The hash `fea865758616c492` is the artifact a reviewer runs against;
re-running the chunk at this package version on a different machine
reproduces it. That is the determinism contract.

## 3. The contrast with A6

Two case-study vignettes use the same workflow shape against the
same artifact (`multisitedgp_data` tibble). The difference is the
front door — what the analyst commits to in the first call. The
table below is the "what changed" reference card; copy it into a
notebook when switching between paradigms.

| Dimension | A6 (multisite trial) | A7 (this vignette) |
|-----------|----------------------|--------------------|
| Paradigm  | site-size-driven (the blueprint's abbreviation: Paradigm A) | direct-precision (the blueprint's abbreviation: Paradigm B) |
| Front door| `sim_multisite()`    | `sim_meta()`       |
| Sample-size knob | `nj_mean`, `cv`, `nj_min` (per-site rosters) | `(I, R)` (informativeness and SE-heterogeneity) |
| `n_j` column | populated integer rosters | `NA` (no per-study sample sizes) |
| Reportable from | a planned trial with per-site rosters in hand | a published record with effect sizes and SEs |
| Output    | `multisitedgp_data` tibble | `multisitedgp_data` tibble (same class) |

The `multisitedgp_data` class is the contract that both paradigms
honor — the same diagnostic ladder, the same plot recipes, the same
adapter handoffs work on either. The paradigm choice is the choice
of what the *front-door knob* is, not what the simulation produces.
A meta-analyst forced into the site-size paradigm would have to
defend fictitious per-study $n_j$ to the reviewer. A multisite-trial
designer forced into the direct-precision paradigm would lose the
operational handle (bigger rosters, smaller SEs) the program office
expects to negotiate over. Each paradigm earns its keep on its own
front door. The methodological case for the split is documented in
[M3 · Margin and SE models](m3-margin-se-models.html); the
substantive case is the difference between A6's tutoring-pilot
narrative and the meta-analytic narrative this vignette walks.

A subtle consequence of the switch: the A6 anchor reports
`feasibility_index = 4.08` at $J = 30$ schools, and the A7 anchor
reports `feasibility_index = 3.624` at $K = 12$ studies. Different
front doors, similar feasibility regime — both case studies live
near the public floor of $5$, and both vignettes' surviving
recommendation pivots on lifting one knob (here, raising
$\sigma_\tau$ at fixed $K$ would not help because the feasibility
gate at $K = 12$ is dominated by $K$, not by $\sigma_\tau$). A
meta-analyst pushing for power-by-feasibility would pre-register a
larger $K$ as the eligibility-criteria sensitivity sweep, the way
A6's reviewer pushed for a larger $J$.

## 4. Add precision–effect dependence

The anchor design declares no precision-effect coupling
(`dependence = "none"`, the default). The pre-registration's
"respect the small-study selection signature" commitment requires a
second design that targets a positive residual rank correlation
between the standardized effect and the sampling variance:
$\rho_S = 0.30$. The
[`sim_meta()`](../reference/sim_meta.html) function takes the same
front-door arguments plus a `dependence` method and a `rank_corr`
target:

```{r dep-sim}
dat_dep <- sim_meta(
  J          = 12L,
  I          = 0.30,
  R          = 2,
  sigma_tau  = 0.20,
  dependence = "rank",
  rank_corr  = 0.30,
  seed       = 1L
)
dat_dep
```

The same twelve studies, the same realized $I = 0.300$, the same
realized $R = 2.000$, the same realized $\sigma_\tau = 0.162$
(under-realization is the small-K Group A FAIL from §2 carried
forward). What changes is the `rho_S` row in the tibble header — now
$0.287$ against the declared $0.300$. The full diagnostic readout:

```{r dep-summary}
summary(dat_dep)
```

The Group B reads tell the dependence story. Realized residual rank
correlation $0.287$ against target $0.300$ — PASS at relative error
$-4.4$ percent; the alignment landed within the $5$-percent gate
the package uses by default. The Pearson correlation $0.325$ is
*not* targeted (Layer 3's contract is rank-scale; Pearson is the
diagnostic-output column, not the alignment target), so its FAIL
against $0.000$ is a feature of the print, not a feature of the
realized design. Groups A, C, and D are unchanged from §2 — the
dependence injection runs on top of the residuals without disturbing
the scale or shape contracts. The hash:

```{r dep-hash}
canonical_hash(dat_dep)
```

Two hashes are now in play. `fea865758616c492` is the anchor (no
dependence); `862fe27147e750df` is the dependence-aware design.
Both go into the methods appendix because both are part of the pre-
registered sensitivity story — the anchor establishes that the scale
and shape diagnostics behave as expected at $\rho_S = 0$, and the
second design establishes that the precision-effect coupling
realizes within tolerance when declared.

A small but important diagnostic check, requested by name to
underline the contract:

```{r dep-realized}
realized_rank_corr(dat_dep)
```

The number `0.2867133` is the residual-scale Spearman correlation
between $z_j$ and $\widehat{se}_j^2$ in the simulated data; it is
the realized-side of the same row the `summary()` print reports as
$0.287$. The two numbers agree by construction — `summary()` calls
`realized_rank_corr()` internally — but the standalone helper is
useful when an analyst wants the bare scalar for a logging line or
a unit test.

## 5. Plots

Four plots support the two designs. Plot 1 is the meta-analysis
"forest" view of the dependence-aware design — point estimates with
95-percent intervals, ordered by estimate. Plot 2 is the funnel —
the visual signature of the small-study selection effect. Plot 3 is
the residual-scale dependence scatter — the alignment contract
visualized as Spearman corr 0.287. Plot 4 is the heterogeneity
view — a histogram of the per-study sampling variances, contrasting
the two designs at a glance.

### Plot 1 — meta-analysis forest

The traditional forest plot of meta-analysis orders the studies by
estimated effect with 95-percent intervals; the package's
[`plot_effects()`](../reference/plot_effects.html) function produces
the exact figure for the `multisitedgp_data` class. Each row is one
of the $K = 12$ included studies, the bullet is the point estimate,
and the horizontal segment is the 95-percent interval based on the
per-study SE:

```{r plot-forest, fig.cap = "Forest plot for the dependence-aware meta-analytic design (K = 12 studies, I = 0.30, R = 2, sigma_tau = 0.20, declared rank_corr = 0.30 on the residual scale, seed = 1L). Studies ordered by estimated effect, with 95 percent confidence intervals based on the per-study SE. The horizontal line at zero is the no-effect reference; intervals crossing the line indicate studies whose effect estimate is not distinguishable from zero at this precision. Interval widths vary modestly because R = 2 — the largest study is twice as precise as the smallest. The pattern of widths varying with the point estimates is the residual-scale dependence the design declared (rho_S = 0.287 PASS).", fig.alt = "Forest-style plot of 12 simulated meta-analytic studies' point estimates with 95 percent confidence intervals, ordered by estimate."}
plot_effects(dat_dep) +
  ggplot2::labs(
    title    = "Dependence-aware design — forest view",
    subtitle = "K = 12 studies; realized residual rank corr 0.287 against declared 0.30 (PASS)"
  )
```

Three things to read. The forest is ordered by estimated effect, so
the leftmost rows are the smallest (most negative) point estimates
and the rightmost are the largest (most positive). The interval
widths vary modestly across studies because the SE-heterogeneity
ratio $R = 2$ is *moderate* — the most-precise study has SE roughly
half the SE of the least-precise study, but the spread is not
extreme. The pattern of *interval widths varying together with the
estimates* — wider intervals at the larger-effect end of the order —
is the small-study-selection signature the dependence target asked
the design to encode. A reviewer expecting a publication-bias-shaped
forest sees the shape on the figure, then confirms the alignment by
the rank correlation in §4.

### Plot 2 — funnel for the dependence-aware design

The funnel plot is the visual contract of meta-analytic publication-
bias diagnostics: the per-study SE on the x-axis, the study's
estimate on the y-axis. A symmetric inverted-V is the visual
signature of $\rho_S = 0$. A *tilted* cloud — small-SE studies
clustering near zero, large-SE studies skewing toward larger
estimates on one side — is the visual signature of non-zero
precision-effect dependence:

```{r plot-funnel, fig.cap = "Funnel plot for the dependence-aware meta-analytic design (K = 12, I = 0.30, R = 2, sigma_tau = 0.20, declared rank_corr = 0.30 on the residual scale, seed = 1L). Standard error on the x-axis (smaller = more precise); estimated study effect on the y-axis. Small-SE studies cluster near and below the central tendency; large-SE studies skew toward larger positive estimates — the visual signature of small-study selection. The realized residual Spearman correlation of 0.287 against the declared 0.30 confirms the alignment ran (PASS). Read together with Plot 3 — the funnel shows the marginal-scale signature; the residual-scale scatter shows the source.", fig.alt = "Funnel plot of 12 simulated study estimates against their standard errors, with the visual tilt of precision-effect dependence."}
plot_funnel(dat_dep) +
  ggplot2::labs(
    title    = "Dependence-aware design — funnel view",
    subtitle = "Realized residual rank corr 0.287 against declared 0.30 (PASS)"
  )
```

Three things to read. The cloud's vertical spread widens to the right
because larger sampling variance corresponds to less-precise studies
— a Group A scale feature, not a dependence feature. The cloud's
*tilt* — the asymmetric scatter of the right-hand (less-precise)
studies relative to the central tendency — is the dependence
feature. The same plot rendered on the anchor design (no dependence)
would show a symmetric inverted-V; a side-by-side comparison is what
a reviewer would look at to confirm that the small-study selection
signature is *deliberately encoded* in the simulation, not an
accident of the sampling distribution.

### Plot 3 — residual-scale dependence scatter

The funnel in Plot 2 makes the dependence visible on the *marginal*
scale (the y-axis is the estimate, not the residual). The
residual-scale scatter is the cleaner read of the alignment contract
because it strips out the marginal effect:
[`plot_dependence(dat, by_residual = TRUE)`](../reference/plot_dependence.html)
plots $z_j$ against $\widehat{se}_j^2$ directly:

```{r plot-dependence, fig.cap = "Residual-scale dependence scatter for the dependence-aware design (K = 12, I = 0.30, R = 2, sigma_tau = 0.20, declared rank_corr = 0.30 on the residual scale, seed = 1L). Standardized residual z_j on the y-axis, sampling variance se2_j on the x-axis. The upward smoother and realized Spearman 0.287 against the declared 0.30 are the residual-scale alignment contract — what Layer 3 actually targets. Read this together with Plot 2 — the funnel shows the marginal-scale signature; this scatter shows the residual-scale source.", fig.alt = "Scatterplot of standardized residual study effects z_j against per-study sampling variance se2_j with an upward smoother."}
plot_dependence(dat_dep, by_residual = TRUE) +
  ggplot2::labs(
    title    = "Dependence-aware design — residual-scale scatter",
    subtitle = "Realized Spearman 0.287 against declared 0.30 (PASS)"
  )
```

Three things to read. The upward slope is the alignment contract —
Layer 3 coupled $z_j$ to $\widehat{se}_j^2$ at the requested $0.30$
correlation, and the realized $0.287$ confirms the coupling ran.
The y-axis is the *residual* $z_j$, not the marginal $\tau_j$; with
no covariate in play here the marginal-scale plot looks
indistinguishable, but the package convention reports the
*residual-scale* number as the contract because that is the surface
the alignment targets. The sample is small ($K = 12$) so the cloud
is sparse; what matters for the diagnostic is the realized Spearman,
not the visual eye-test.

### Plot 4 — heterogeneity view (sampling-variance histogram)

The fourth plot is the methods-appendix headline: the per-study
sampling variance distribution, contrasting the two designs side by
side. It answers the reviewer's question — "are your two simulation
draws sampling from the same precision profile?" — at one glance:

```{r plot-heterogeneity-data, include = FALSE}
heterog_df <- rbind(
  data.frame(
    se2_j  = dat$se2_j,
    design = "Anchor (rho_S = 0)"
  ),
  data.frame(
    se2_j  = dat_dep$se2_j,
    design = "Dependence-aware (rho_S = 0.30)"
  )
)
heterog_df$design <- factor(
  heterog_df$design,
  levels = c("Anchor (rho_S = 0)", "Dependence-aware (rho_S = 0.30)")
)
```

```{r plot-heterogeneity, fig.cap = "Heterogeneity view: histogram of per-study sampling variances se2_j for the anchor design and the dependence-aware design (K = 12 studies each, I = 0.30, R = 2, sigma_tau = 0.20, seed = 1L). The two distributions overlap because the (I, R) precision profile is shared between the designs — the dependence injection runs on the residual layer without altering the marginal precision distribution. The geometric mean of se2_j realized at 0.093 in both designs against the implied target of 0.093 (PASS in Group A). The R = 2 SE-heterogeneity ratio shows up as a roughly 4-fold spread in se2_j between the smallest and largest study (R is on the SE scale; the variance scale is its square).", fig.alt = "Histogram of per-study sampling variances for two simulated meta-analytic designs, faceted by design."}
ggplot2::ggplot(heterog_df, ggplot2::aes(x = se2_j, fill = design)) +
  ggplot2::geom_histogram(bins = 8L, color = "white",
                          show.legend = FALSE) +
  ggplot2::facet_wrap(~ design, ncol = 2L) +
  ggplot2::scale_fill_manual(
    values = c("Anchor (rho_S = 0)"             = "grey55",
               "Dependence-aware (rho_S = 0.30)" = "steelblue4")
  ) +
  ggplot2::labs(
    x        = expression(hat(se)[j]^2),
    y        = "Studies",
    title    = "Per-study sampling variance across the two designs",
    subtitle = "Same (I, R) precision profile; dependence injection touches residuals only"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

Read this against the §4 dependence claim. The two histograms occupy
the same support and have the same coarse shape because the
precision profile is *not* what changes between the anchor and the
dependence-aware design — the change is the residual-effect coupling
to that profile. The geometric mean of $\widehat{se}_j^2$ is $0.093$
in both designs (Group A PASS), and the realized $R = 2.00$ is
identical. A reviewer worried that the dependence injection was
secretly changing the precision profile reads this figure and
confirms it is not. The dependence-aware design's $\rho_S = 0.287$
realization is what visualizes on Plots 2 and 3; the marginal
precision distribution on Plot 4 is shared.

## 6. Adapter handoff to metafor

The pre-registered simulation is the input to the meta-analytic
estimator the synthesis manuscript reports. Priya's pipeline is:
simulate, audit, then run a random-effects meta-analytic model with
[metafor::rma](https://metafor-project.org/) on the simulated
`multisitedgp_data`. The
[`as_metafor()`](../reference/as_metafor.html) adapter performs the
column rename — the `multisitedgp_data` columns `tau_j_hat`, `se2_j`,
`se_j` become metafor's expected `yi`, `vi`, `sei` — and returns a
plain tibble:

```{r adapter-handoff}
meta_input <- as_metafor(dat_dep)
meta_input
names(meta_input)
```

Three columns: `yi` (the effect-size estimate), `vi` (the sampling
variance), `sei` (the sampling standard error). These are the column
names metafor's
[`rma()`](https://wviechtb.github.io/metafor/reference/rma.uni.html)
function expects, which is the entire purpose of the rename — the
analyst's downstream call is now a one-liner. The estimator call is
guarded with `requireNamespace()` because metafor is a *soft*
dependency in `multisiteDGP` (Suggests, not Imports), and the
vignette must build whether or not metafor is installed:

```{r metafor-call, eval = requireNamespace("metafor", quietly = TRUE)}
fit <- metafor::rma(yi = yi, vi = vi, data = meta_input,
                    method = "REML")
fit
```

The `rma()` print reports the random-effects model summary: a pooled
estimate $\widehat\mu$ with its standard error and 95-percent
interval; the residual heterogeneity statistic $\widehat\tau^2$ and
its $Q$-test; and the moment-based heterogeneity ratio $I^2$. The
pooled estimate is the meta-analytic statistic the synthesis
manuscript reports; the heterogeneity row is the headline of how
much between-study variation the pooled estimate is averaging over.

A complementary view that does not require metafor is the per-study
column rename, which is informative on its own:

```{r adapter-rename}
data.frame(
  multisitedgp = c("tau_j_hat", "se2_j", "se_j"),
  metafor      = c("yi",        "vi",    "sei")
)
```

The rename is the entire interface contract between the simulation
and the downstream estimator. The same adapter pattern works for the
two other supported downstream packages — `as_baggr()` for Bayesian
hierarchical estimation, `as_multisitepower()` for the multisitepower
power-calculation API — and the per-package column conventions are
documented in
[M6 · Adapters and downstream packages](m6-adapters-downstream.html).
Switching downstream estimators is a one-line change in the analyst's
script, never a re-simulation.

The contrast with the anchor design is one short check: re-fit on
the anchor data, compare the pooled estimate. The anchor lacks the
precision-effect coupling, so a small-study-corrected estimator
(such as Egger's regression run on top of `rma()`) would *not* fire
on the anchor and *would* fire on the dependence-aware design. That
contrast is the discipline the simulation enables — every claim
about "what does the small-study selection do to the pooled
estimate" is grounded in two simulated draws, not in a single
observed dataset that conflates the two regimes.

```{r metafor-anchor-call, eval = requireNamespace("metafor", quietly = TRUE)}
anchor_input <- as_metafor(dat)
fit_anchor <- metafor::rma(yi = yi, vi = vi, data = anchor_input,
                           method = "REML")
fit_anchor
```

Two pooled estimates, two confidence intervals, one substantive
comparison. The pre-registration commits to the comparison *before*
the synthesis runs the estimator on the real published record; the
simulation's job is to anchor the reviewer's expectations about
what the estimator does at the declared precision profile.

## 7. Defensibility and provenance

Every claim in this vignette is anchored by an artifact a reviewer
can re-run. The two artifacts are the anchor's canonical hash and
the dependence-aware design's canonical hash; the
[`provenance_string()`](../reference/provenance_string.html) helper
returns the same information formatted for a methods footnote:

```{r provenance}
provenance_string(dat_dep)
```

The string carries the package version, the paradigm
(`paradigm=direct`), the seed (`seed=1`), the canonical hash for the
simulation (`canonical_hash=862fe27147e750df`), the design hash, the
hash algorithm, the R version, and the hook configuration — every
input that influenced the output. A reviewer who pastes the string
into a methods footnote and re-runs the vignette at the same package
version reproduces the same hash.

The three-line methods-appendix template, generalized for the
meta-analytic case study:

> **Pre-registered simulation.** Pooled-effect simulations were run
> with `multisiteDGP` `r as.character(packageVersion("multisiteDGP"))`
> on the direct-precision path at $K = 12$ studies, $I = 0.30$,
> $R = 2$, $\sigma_\tau = 0.20$, with no precision-effect dependence
> in the anchor design and a residual rank correlation
> $\rho_S = 0.30$ in the dependence-aware design. The anchor's
> canonical hash is `fea865758616c492`; the dependence-aware
> design's canonical hash is `862fe27147e750df`. The full provenance
> strings are reproduced verbatim in Supplementary Materials S1.
>
> **Estimator.** Pooled effects were estimated by random-effects
> meta-analysis with `metafor::rma()` (REML), called on the
> simulation tibbles after the
> [`as_metafor()`](../reference/as_metafor.html) column rename.
>
> **Reproducibility contract.** All hashes were computed with
> `xxhash64` against the package's canonical serialization. Re-
> running the vignette at the recorded package version on any
> machine reproduces both hashes; mismatched hashes invalidate the
> simulation contract.

The three-paragraph pattern is the discipline the rest of the
package's methods-track vignettes
[(M7)](m7-reproducibility-provenance.html) formalize: every claim
cites either a chunk above it or a hash recorded in the manifest.
Nothing is hidden in the prose. The manifest below is the artifact
saved alongside the manuscript draft:

```{r manifest}
manifest <- list(
  project          = "meta-analysis-A7-vignette",
  K                = 12L,
  I_target         = 0.30,
  R_target         = 2,
  sigma_tau_target = 0.20,
  rho_S_target     = c(anchor = 0, dependence = 0.30),
  seed             = 1L,
  anchor_hash      = canonical_hash(dat),
  dependence_hash  = canonical_hash(dat_dep),
  paradigm         = "direct"
)
manifest
```

Every field either appears in a chunk above (the two hashes from §2
and §4) or is part of the substantive design specification (the
targets, the seed, the paradigm). A methods reviewer auditing the
pre-registration loads the manifest, runs the vignette, and confirms
the two hashes match; that is the contract.

A small note on the `preset_meta_modest()` shortcut: the package
ships a preset that lifts the boilerplate of the §2 anchor
specification:

```{r preset, eval = FALSE}
preset_meta_modest()
```

The preset returns a `multisitedgp_design` object on the
direct-precision path with $J = 50$, $I = 0.30$, $R = 1.5$,
$\sigma_\tau = 0.20$ — modest informativeness, modest SE-
heterogeneity, modest between-study spread. A pre-registration that
adopts the preset instead of the hand-built call is *more*
defensible because the preset is package-authored and version-
controlled; the trade is that the preset's $J = 50$ is not the
$K = 12$ this case study runs at, so the pre-registration would
override $J$ in the call. See
[A2 · Choosing a preset](a2-choosing-a-preset.html) for the full
preset catalog.

## 8. Where to next

- [A6 · Case study — multisite trial](a6-case-study-multisite.html)
  — the sibling case study on the site-size-driven path; the
  paradigm contrast in §3 makes the workflow split legible. The two
  vignettes are designed to be read together when an analyst is
  unsure which paradigm the project sits in.
- [A8 · Cookbook](a8-cookbook.html) — the recipe-book extension of
  the case studies; the meta-analytic case study covered here maps
  to the cookbook's "small-K synthesis with publication-bias
  signature" recipe, which adds a back-calculation step from the
  observed published record to $(I, R)$ targets.
- [A2 · Choosing a preset](a2-choosing-a-preset.html) — the full
  catalog of nine presets; `preset_meta_modest()` is the
  direct-precision preset; the eight others are site-size-driven
  presets calibrated to specific empirical heritages.
- [A3 · Diagnostics in practice](a3-diagnostics-in-practice.html)
  — the four-question rubric the §2 and §4 summary prints read off;
  pairs naturally with this vignette when the reader wants to dig
  into the FAILs the small-K regime surfaces.
- [A4 · Covariates and precision dependence](a4-covariates-precision-dependence.html)
  — the residual-vs-marginal split that the dependence diagnostics
  in §4 rely on; relevant when the meta-analyst decides to absorb
  study-level moderators (publication year, design type) into the
  simulation as covariates.
- [M3 · Margin and SE models](m3-margin-se-models.html) — the
  formal Layer 2 split between the site-size-driven and
  direct-precision paths the §3 contrast table sketches.
- [M6 · Adapters and downstream packages](m6-adapters-downstream.html)
  — the full grammar of `as_metafor()`, `as_baggr()`,
  `as_multisitepower()`, and the column-rename conventions; the
  go-to reference when extending the §6 metafor handoff to other
  estimators.
- [Reproducibility and provenance](m7-reproducibility-provenance.html)
  — the formal manifest pattern, the determinism contract, and the
  full grammar of `provenance_string()` and `canonical_hash()`.

A natural extension when the synthesis manuscript advances to the
review stage: re-run the §6 metafor handoff against the *observed*
published record (real `yi`, `vi`, `sei` from the included studies),
compare the pooled estimate to the simulated pooled estimate, and
report the gap as a sensitivity number. The simulation's role is to
calibrate the reviewer's expectations *before* the real estimator
runs; the calibration's defensibility is the entire reason for the
pre-registration.

## References

The four-question diagnostic rubric and the residual-scale
dependence convention are introduced in @lee2025improving; the
empirical-Bayes shrinkage context that anchors the Group D
feasibility gate is collected in @walters2024empirical. The
random-effects estimator that the §6 `metafor::rma()` handoff fits
is documented in the `metafor` package itself; classic
meta-analytic methodology references are intentionally left to the
analyst's choice rather than prescribed by this package.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
