---
title: "A8 · Cookbook"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For analysts who already know what to compute and want a working
  pattern they can copy. Each of the nine recipes is a self-contained
  script under `inst/cookbook/` that runs a small end-to-end workflow
  and prints a `COOKBOOK_RESULT` marker the audit harness checks. The
  walkthroughs below run each recipe, show the printed output, and add
  one figure of context per recipe. Use this vignette as a reference
  card; the case studies in
  [A6](a6-case-study-multisite.html) and
  [A7](a7-case-study-meta-analysis.html) carry the longer narratives.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A8 Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. What is a cookbook recipe?

A cookbook recipe is a short, self-contained R script under
`inst/cookbook/` that loads the package, runs a small end-to-end
workflow, asserts a few `stopifnot()` invariants, and prints a single
`COOKBOOK_RESULT` line at the end. The audit harness in
`inst/scripts/cookbook_audit.R` runs each recipe and matches its
output against `expected_marker` in
`inst/cookbook/cookbook-manifest.csv`; a missing or malformed marker
is a smoke-test failure. Recipes therefore play two roles at once —
they are pedagogy (a paste-ready pattern for one common task) and
they are smoke tests (the package builds and behaves as documented).
The walkthroughs in this vignette run each recipe inline, surface the
marker, and add a figure of context.

## 2. Recipe index

```{r manifest}
manifest_path <- system.file(
  "cookbook", "cookbook-manifest.csv", package = "multisiteDGP"
)
recipes <- utils::read.csv(manifest_path, stringsAsFactors = FALSE)
recipes[, c("id", "title", "file")]
```

Use the table below when scanning for a pattern you can copy.

| ID | Title | What it demonstrates | When to reach for it |
|---|---|---|---|
| R5.1 | Hello-world JEBS-style workflow | `sim_multisite()` + `plot_effects()` from a published preset | First exposure; smoke check after install |
| R5.2 | Custom calibration with `sigma_tau` and `R2` | Preset override of two anchor knobs | Pinning a substantive heterogeneity target with a covariate proxy |
| R5.3 | Compare two presets side by side | `preset_education_small()` vs `preset_education_substantial()` | Bracketing an evidence claim across the field's regime |
| R5.4 | Bridge to downstream adapters | `as_metafor()` and `as_baggr()` with namespace guards | Handing simulated data to an external estimator |
| R5.5 | User `g_fn` callback | A custom Layer 1 G-distribution callback | A site-effect distribution that none of the eight built-ins covers |
| R5.6 | Custom `dependence_fn` | A custom Layer 3 alignment callback | Replacing the built-in copula/rank/hybrid alignment |
| R5.7 | Design grid and scenario audit | `design_grid()` + `scenario_audit()` for a small factorial | Pre-registration; defending a chosen cell |
| R5.8 | Canonical hash reproducibility | Hash determinism across reruns and column orderings | Provenance; cache invalidation |
| R5.9 | Debugging feasibility diagnostics | Reading `feasibility_index()` across seeds | Diagnosing a flagged diagnostic before changing the design |

Each section below loads the recipe verbatim from the installed
script, runs it, and reads off the `COOKBOOK_RESULT` line. The
exported function names used in every recipe are entries in
`NAMESPACE` (verified before drafting).

## 3. R5.1 — Hello-world JEBS-style workflow

What it does: simulates a 50-site dataset from
`preset_jebs_paper()`, builds the caterpillar plot, and prints
$\hat I$ and the canonical hash.

When to reach for it: a 30-second smoke check after install, or the
shortest possible reference for the
`sim_multisite()` + `plot_effects()` pair.

The recipe ships verbatim as `R5-1-hello-world.R`:

```{r r51-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-1-hello-world.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r51-run}
dat_r51 <- sim_multisite(preset_jebs_paper(), seed = 4719L)
diag_r51 <- attr(dat_r51, "diagnostics")
effect_plot_r51 <- plot_effects(dat_r51, caption = FALSE)
```

```{r r51-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.1 status=PASS J=%d I=%.3f hash=%s\n",
  nrow(dat_r51),
  diag_r51$I_hat,
  canonical_hash(dat_r51)
))
```

The marker reports $J = 50$, $\hat I = 0.265$, and a stable hash;
the assertions in the recipe confirm `dat` is a `multisitedgp_data`
object, `effect_plot` inherits from `ggplot`, and $\hat I$ is finite.
Render the caterpillar:

```{r r51-plot, fig.cap = "R5.1: caterpillar of estimated site effects with shrinkage targets — sites are ordered by point estimate, and the y-axis spread reads as the across-site signal width on the residual scale.", fig.alt = "Caterpillar plot of estimated site effects from the JEBS-style preset, with sites ordered by point estimate along the x-axis and 95 percent intervals around each estimate; vertical spread of the dots represents the realized between-site heterogeneity."}
effect_plot_r51
```

This is the JEBS-paper preset's calibration heritage
[@lee2025improving] rendered in one figure: with $\sigma_\tau = 0.20$
and $J = 50$, the caterpillar shows the across-site spread and the
signal envelope you would publish.

## 4. R5.2 — Custom calibration with `sigma_tau` and `R2`

What it does: overrides two anchor knobs of
`preset_education_modest()` — between-site SD `sigma_tau = 0.25` and
covariate proxy `R2 = 0.40` — and reads the feasibility index off the
resulting dataset.

When to reach for it: pinning a substantive heterogeneity target
(say, $\sigma_\tau = 0.25$ from a prior trial) while letting a
school-level covariate absorb baseline variation.

```{r r52-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-2-custom-calibration.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r52-run}
design_r52 <- preset_education_modest(sigma_tau = 0.25, R2 = 0.40)
dat_r52 <- sim_multisite(design_r52, seed = 91207L)
fi_r52 <- feasibility_index(dat_r52, warn = FALSE)
```

```{r r52-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.2 status=PASS sigma_tau=%.2f R2=%.2f feasibility=%.3f\n",
  attr(dat_r52, "design")$sigma_tau,
  attr(dat_r52, "design")$R2,
  fi_r52
))
```

The override survives the round-trip: `attr(dat, "design")$R2` is
`0.40` exactly (the recipe asserts this). The realized feasibility
index sits well above the conventional `n_eff = 1` floor, consistent
with the multisite-trial calibration heritage at modest $\sigma_\tau$
[@weiss2017howmuch]. To see what changes when only `R2` moves, sweep
it on a coarse grid:

```{r r52-sweep}
r2_grid <- c(0.0, 0.2, 0.4, 0.6, 0.8)
sweep_r52 <- vapply(r2_grid, function(r2) {
  d <- sim_multisite(
    preset_education_modest(sigma_tau = 0.25, R2 = r2),
    seed = 91207L
  )
  c(I = informativeness(d), feasibility = feasibility_index(d, warn = FALSE))
}, numeric(2))
sweep_r52 <- data.frame(R2 = r2_grid, t(sweep_r52))
sweep_r52
```

```{r r52-plot, fig.cap = "R5.2: residual-scale informativeness rises with R2 — with sigma_tau = 0.25 fixed, absorbing more baseline variation through the covariate proxy moves I from below 0.30 toward 0.40, the regime where shrinkage starts paying.", fig.alt = "Line plot showing realized informativeness I on the y-axis against covariate-explained variance R-squared on the x-axis at fixed sigma_tau = 0.25; the curve rises monotonically across the R-squared range."}
ggplot(sweep_r52, aes(R2, I)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = r2_grid) +
  labs(x = expression(R^2), y = expression(hat(I))) +
  theme_minimal(base_size = 11)
```

## 5. R5.3 — Compare two presets side by side

What it does: simulates `preset_education_small()` and
`preset_education_substantial()` at the same $J$ and seed, then
compares informativeness.

When to reach for it: bracketing an evidence claim across the
field's regime — "even at the small end of the literature, $\hat I$
is x" or "at the substantial end, shrinkage costs y".

```{r r53-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-3-compare-presets.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r53-run}
small_r53 <- sim_multisite(preset_education_small(J = 40L), seed = 33113L)
substantial_r53 <- sim_multisite(
  preset_education_substantial(J = 40L), seed = 33113L
)
small_i <- informativeness(small_r53)
substantial_i <- informativeness(substantial_r53)
```

```{r r53-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.3 status=PASS small_I=%.3f substantial_I=%.3f\n",
  small_i,
  substantial_i
))
```

The recipe asserts `substantial_i > small_i` — at $J = 40$ on the
same seed, $\hat I$ moves from roughly 0.02 (small heterogeneity, no
useful between-site signal to recover) to roughly 0.62 (substantial
heterogeneity, shrinkage carries real information). Pair the two
caterpillars to see the contrast:

```{r r53-plot, fig.width = 9, fig.height = 5, fig.cap = "R5.3: caterpillars at small versus substantial heterogeneity — the substantial-preset spread (right) is materially wider than the small-preset spread (left), confirming the I gap reported in the marker.", fig.alt = "Two side-by-side caterpillar plots, one for preset_education_small (left, narrow vertical spread reflecting sigma_tau = 0.05) and one for preset_education_substantial (right, wide vertical spread reflecting sigma_tau = 0.30)."}
p_small <- plot_effects(small_r53, caption = FALSE) +
  ggtitle(sprintf("Small (I = %.3f)", small_i))
p_substantial <- plot_effects(substantial_r53, caption = FALSE) +
  ggtitle(sprintf("Substantial (I = %.3f)", substantial_i))
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_small, p_substantial, ncol = 2)
} else {
  p_small
}
```

The small-vs-substantial gap aligns with the empirical Bayes
calibration patterns in Walters' synthesis [@walters2024empirical] —
the substantial-heterogeneity end is where between-site signal
recovery becomes the dominant inferential question.

## 6. R5.4 — Bridge to downstream adapters

What it does: simulates from `preset_education_modest()`, hands the
result to `as_metafor()` and `as_baggr()` (with namespace guards so
the recipe still runs without those packages installed), and verifies
that the residual-scale informativeness $\hat I$ is preserved exactly
across the rename.

When to reach for it: passing simulated data to an external
estimator without losing the design's calibration.

```{r r54-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-4-downstream-adapters.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r54-run}
dat_r54 <- sim_multisite(preset_education_modest(), seed = 50421L)
sigma_tau_r54 <- attr(dat_r54, "design")$sigma_tau

if (requireNamespace("metafor", quietly = TRUE)) {
  mf_r54 <- as_metafor(dat_r54)
} else {
  mf_r54 <- tibble::tibble(
    yi  = dat_r54$tau_j_hat,
    vi  = dat_r54$se2_j,
    sei = dat_r54$se_j
  )
}

if (requireNamespace("baggr", quietly = TRUE)) {
  bg_r54 <- as_baggr(dat_r54, include_truth = TRUE)
} else {
  bg_r54 <- tibble::tibble(
    tau      = dat_r54$tau_j_hat,
    se       = dat_r54$se_j,
    tau_true = dat_r54$tau_j
  )
}

i_before_r54 <- compute_I(dat_r54$se2_j, sigma_tau = sigma_tau_r54)
i_after_r54  <- compute_I(mf_r54$vi,    sigma_tau = sigma_tau_r54)
```

```{r r54-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.4 status=PASS I_delta=%.3e metafor_guard=%s baggr_guard=%s\n",
  abs(i_before_r54 - i_after_r54),
  requireNamespace("metafor", quietly = TRUE),
  requireNamespace("baggr", quietly = TRUE)
))
```

The first three rows of the metafor frame use the canonical
column names `yi / vi / sei`:

```{r r54-mf-head}
head(mf_r54, 3)
```

The baggr frame uses `tau / se / tau_true`:

```{r r54-bg-head}
head(bg_r54, 3)
```

The recipe's invariant — $\hat I$ before and after rename agree to
within `1e-12` — guards against the row F-3 failure mode (post-adapter
column drift). A funnel of the renamed estimates makes the bridge
visible:

```{r r54-plot, fig.cap = "R5.4: funnel of metafor-renamed estimates — the yi versus sei cloud is the same cloud we would have plotted from tau_j_hat versus se_j, confirming the rename is a relabel rather than a transform.", fig.alt = "Funnel plot with sampling standard error sei on the x-axis and effect estimate yi on the y-axis after as_metafor() column rename; the point cloud tapers as sei shrinks, the standard funnel signature."}
ggplot(mf_r54, aes(yi, sei)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_y_reverse() +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  labs(x = expression(y[i]~"(estimate)"),
       y = expression(s*e[i]~"(reverse axis)")) +
  theme_minimal(base_size = 11)
```

## 7. R5.5 — User `g_fn` callback

What it does: defines a beta-difference G-distribution callback,
wires it into a `multisitedgp_design()` with `true_dist = "User"`,
and verifies the realized standardized SD recovers $\sigma_\tau$.

When to reach for it: a site-effect distribution that none of the
eight built-in G distributions covers — a bimodal or lopsided shape
the analyst wants to inject directly. The callback contract (input
`J` plus `...`, return either the standardized $z$-vector or a list
with `z`) is documented at
[`multisitedgp_design()`](../reference/multisitedgp_design.html).

```{r r55-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-5-user-g-fn.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r55-run}
beta_diff_g <- function(J, ...) {
  flip <- stats::rbinom(J, 1, 0.5)
  raw <- ifelse(flip == 1, stats::rbeta(J, 2, 5), -stats::rbeta(J, 5, 2))
  (raw - mean(raw)) / stats::sd(raw)
}

design_r55 <- multisitedgp_design(
  J          = 60L,
  true_dist  = "User",
  g_fn       = beta_diff_g,
  g_returns  = "standardized",
  sigma_tau  = 0.20
)
dat_r55 <- sim_multisite(design_r55, seed = 71823L)
realized_sigma_r55 <- stats::sd(dat_r55$z_j) * design_r55$sigma_tau
```

```{r r55-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.5 status=PASS realized_sigma=%.3f I=%.3f\n",
  realized_sigma_r55,
  informativeness(dat_r55)
))
```

Because the callback returns standardized values
(`g_returns = "standardized"`), the realized SD on the $\tau$-scale
is exactly `sd(z_j) * sigma_tau`, and the recipe's `stopifnot()`
catches drift if the callback ever returned an un-standardized
vector. Inspect the realized shape:

```{r r55-plot, fig.cap = "R5.5: density of standardized site-effect draws z_j from the bimodal beta-difference callback — the two-mode signature is preserved post-standardization, confirming the callback returned what it advertised.", fig.alt = "Kernel density estimate of standardized site-effect draws z_j produced by a beta-difference user callback, showing two distinct modes either side of zero."}
ggplot(dat_r55, aes(z_j)) +
  geom_density(fill = "grey70", colour = "grey20", alpha = 0.6) +
  geom_rug(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  labs(x = expression(z[j]~"(standardized site effect)"),
       y = "density") +
  theme_minimal(base_size = 11)
```

The two-mode signature is what the bimodal callback promised; the
mean-zero, unit-SD constraint is the standardization contract.

## 8. R5.6 — Custom `dependence_fn`

What it does: registers a deterministic rank-alignment callback that
co-orders site effects with site-level standard errors, then verifies
that supplying a malformed callback (one that mutates `se2_j` or
returns a non-permutation) is caught by
`validate_multisitedgp_design()` before the simulator runs.

When to reach for it: replacing the built-in copula / rank / hybrid
alignment with a custom monotone scheme — useful when alignment is
dictated by an external mechanism (e.g., a measurement-precision
relationship the analyst wants to inject deterministically rather
than match in correlation). The callback contract is documented at
[`multisitedgp_design()`](../reference/multisitedgp_design.html).

```{r r56-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-6-custom-dependence-fn.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r56-run}
custom_rank <- function(z_j, se2_j, target, ...) {
  se_order <- order(se2_j)
  z_order  <- order(z_j)
  perm <- integer(length(se2_j))
  if (target >= 0) {
    perm[z_order] <- se_order
  } else {
    perm[z_order] <- rev(se_order)
  }
  list(se2_j = se2_j[perm], perm = as.integer(perm))
}

design_r56 <- multisitedgp_design(
  J             = 40L,
  sigma_tau     = 0.20,
  dependence    = "rank",
  rank_corr     = 0.40,
  dependence_fn = custom_rank,
  seed          = 4719L
)
dat_r56 <- sim_multisite(design_r56)
rho_r56 <- realized_rank_corr(dat_r56)

bad_dep <- function(z_j, se2_j, target, ...) {
  list(se2_j = se2_j * 2, perm = seq_along(se2_j))
}
bad_design_r56 <- update_multisitedgp_design(
  design_r56, dependence_fn = bad_dep
)
bad_caught_r56 <- tryCatch(
  {
    sim_multisite(bad_design_r56)
    FALSE
  },
  error = function(e) TRUE
)
```

```{r r56-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.6 status=PASS rho=%.3f bad_callback_caught=%s\n",
  rho_r56,
  bad_caught_r56
))
```

The deterministic rank-permutation produces a near-exact rank
alignment (`rho` close to 1), well above the `rank_corr = 0.40`
target — the callback is doing what it advertises rather than what
the target requested. The `bad_callback_caught = TRUE` half of the
marker is the second invariant: the malformed callback that scales
`se2_j` is rejected at validation time, not silently absorbed.
Visualize the alignment:

```{r r56-plot, fig.cap = "R5.6: site-level se versus standardized site-effect rank under the custom callback — the monotone band indicates near-perfect rank co-ordering, the callback's deterministic signature.", fig.alt = "Scatterplot of standardized site-effect rank on the x-axis and per-site sampling standard error on the y-axis under a custom dependence callback; the points line up in a tight monotone band."}
ggplot(dat_r56, aes(rank(z_j), se_j)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "loess", se = FALSE, formula = y ~ x,
              colour = "grey30", linewidth = 0.5) +
  labs(x = expression(rank(z[j])), y = expression(s*e[j])) +
  theme_minimal(base_size = 11)
```

## 9. R5.7 — Design grid and scenario audit

What it does: builds an 8-cell direct-precision design grid over
$I \in \{0.20, 0.40\}$, $R \in \{1, 3\}$, $J \in \{14, 30\}$, then
runs `scenario_audit()` with `M = 1` replicate to label which cells
pass.

When to reach for it: pre-registration — defending a chosen design
cell with a per-cell summary table and a hashable audit object.

```{r r57-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-7-design-grid-audit.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r57-run}
grid_r57 <- design_grid(
  paradigm  = "direct",
  I         = c(0.20, 0.40),
  R         = c(1, 3),
  J         = c(14L, 30L),
  sigma_tau = 0.20,
  true_dist = "Gaussian",
  seed_root = 20260507L
)
audit_r57 <- scenario_audit(grid_r57, M = 1L, parallel = FALSE)
```

```{r r57-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.7 status=PASS cells=%d audit_hash=%s\n",
  nrow(audit_r57),
  canonical_hash(audit_r57)
))
```

Inspect the per-cell pass/fail summary:

```{r r57-summary}
audit_r57[, c("cell_id", "I", "R", "J", "pass",
              "med_I_hat", "med_feasibility_efron")]
```

The full audit object carries 29 columns — design factors, status
flags, and the diagnostic medians and tail quantiles per cell.
Render the pass/fail mosaic:

```{r r57-plot, fig.cap = "R5.7: scenario-audit pass/fail mosaic across the I x J grid faceted by R — green cells pass the diagnostic ladder at M = 1; the cells that fail concentrate where I is small and J is small, the regime where shrinkage cannot recover the between-site signal.", fig.alt = "Faceted mosaic plot of scenario_audit pass/fail status across an informativeness I and study-count J grid, faceted by heterogeneity ratio R; cells are color-coded green for pass and red for fail, with failures concentrated at low I and low J."}
ggplot(audit_r57, aes(factor(J), factor(I), fill = pass)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  facet_wrap(~ paste0("R = ", R)) +
  scale_fill_manual(
    values = c(`TRUE` = "#4daf4a", `FALSE` = "#e41a1c"),
    name   = "audit pass"
  ) +
  labs(x = "J", y = "I") +
  theme_minimal(base_size = 11)
```

`M = 1` is a smoke setting — for a defensible pre-registration plan
on `M >= 50` and read off `med_I_hat`, `med_feasibility_efron`, and
the tail quantiles. The longer narrative is in
[A6](a6-case-study-multisite.html).

## 10. R5.8 — Canonical hash reproducibility

What it does: computes the canonical hash of the same design three
times and confirms a single unique value, then permutes column order
and confirms the hash is invariant.

When to reach for it: provenance and cache invalidation — a hash
change on what should be the same artifact is a determinism signal
worth chasing before publication.

```{r r58-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-8-reproducibility-hash.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r58-run}
hashes_r58 <- vapply(
  seq_len(3),
  function(i) canonical_hash(
    sim_multisite(preset_education_modest(), seed = 12345L)
  ),
  character(1)
)
dat_r58 <- sim_multisite(preset_education_modest(), seed = 12345L)
reordered_r58 <- dat_r58[, sort(names(dat_r58))]
```

```{r r58-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.8 status=PASS unique_hashes=%d hash=%s\n",
  length(unique(hashes_r58)),
  hashes_r58[[1L]]
))
```

The three reruns produce one unique hash:

```{r r58-hashes}
hashes_r58
```

And the column-permuted dataset hashes to the same value as the
original, by design — the hash canonicalizes column order before
serialization:

```{r r58-perm}
identical(canonical_hash(dat_r58), canonical_hash(reordered_r58))
```

To make the determinism story visible, sweep across seeds and read
off the per-seed hash:

```{r r58-sweep}
hash_sweep_r58 <- data.frame(
  seed = 12345L + 0:5,
  hash = vapply(
    12345L + 0:5,
    function(s) canonical_hash(
      sim_multisite(preset_education_modest(), seed = s)
    ),
    character(1)
  )
)
hash_sweep_r58
```

```{r r58-plot, fig.cap = "R5.8: per-seed hash uniqueness across six adjacent seeds — every seed produces a distinct hash, while same-seed reruns produce one stable hash, the two halves of canonical determinism.", fig.alt = "Bar chart with one bar per seed value (six adjacent seeds), each labeled with its 16-character canonical hash; every bar shows a distinct hash, illustrating seed-to-hash injectivity."}
hash_sweep_r58$short <- substr(hash_sweep_r58$hash, 1, 8)
ggplot(hash_sweep_r58, aes(factor(seed), 1, fill = short)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = short), size = 3) +
  scale_fill_grey(start = 0.6, end = 0.95) +
  guides(fill = "none") +
  labs(x = "seed", y = "") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The provenance idiom is one line:

```{r r58-prov}
provenance_string(dat_r58)
```

That string is the artifact a reviewer can pin to a methods
appendix [@lee2025improving].

## 11. R5.9 — Debugging feasibility diagnostics

What it does: simulates `preset_education_modest()` at $J = 14$,
$\bar n_j = 60$ — small-trial conditions where the feasibility
diagnostic does real work — and reads `informativeness()` and
`feasibility_index()` across three adjacent seeds.

When to reach for it: a flagged feasibility warning. The first
question is whether it is the design or the seed; this recipe is the
shortest sanity check.

```{r r59-script, echo = FALSE, comment = ""}
cat(readLines(system.file(
  "cookbook", "R5-9-debugging-feasibility.R", package = "multisiteDGP"
)), sep = "\n")
```

Run it:

```{r r59-run}
design_r59 <- preset_education_modest(J = 14L, nj_mean = 60)
dat_r59 <- sim_multisite(design_r59, seed = 12345L)

checks_r59 <- data.frame(
  seed = 12345L + 0:2,
  I = vapply(
    12345L + 0:2,
    function(s) informativeness(sim_multisite(design_r59, seed = s)),
    numeric(1)
  ),
  feasibility = vapply(
    12345L + 0:2,
    function(s) feasibility_index(
      sim_multisite(design_r59, seed = s), warn = FALSE
    ),
    numeric(1)
  )
)
```

```{r r59-marker}
cat(sprintf(
  "COOKBOOK_RESULT R5.9 status=PASS J=%d mean_I=%.3f mean_feasibility=%.3f\n",
  design_r59$J,
  mean(checks_r59$I),
  mean(checks_r59$feasibility)
))
```

The per-seed table is the actual debugging artifact:

```{r r59-table}
checks_r59
```

The three seeds produce $\hat I$ and feasibility values that move by
under 5%, which is what you want — the feasibility warning is a
property of the design, not the seed. If $\hat I$ jumped by an order
of magnitude across adjacent seeds, the design's seed-to-seed
sensitivity would be the diagnosis. Plot the variability:

```{r r59-plot, fig.cap = "R5.9: feasibility index across three adjacent seeds at J = 14, n_j = 60 — bar heights are within 5% of each other, indicating the warning is a stable property of the design rather than a seed-specific artifact.", fig.alt = "Bar chart of Efron feasibility index for three adjacent seed values at J = 14 and nj_mean = 60; the three bars have nearly identical heights, indicating the feasibility warning is design-driven not seed-driven."}
ggplot(checks_r59, aes(factor(seed), feasibility)) +
  geom_col(fill = "grey60", colour = "grey20") +
  geom_text(aes(label = sprintf("%.2f", feasibility)),
            vjust = -0.4, size = 3.5) +
  labs(x = "seed", y = "feasibility index") +
  theme_minimal(base_size = 11) +
  expand_limits(y = max(checks_r59$feasibility) * 1.10)
```

A small-trial regime ($J = 14$, $\bar n_j = 60$) is where the
feasibility diagnostic earns its keep — diagnostics in practice are
covered in [A3](a3-diagnostics-in-practice.html).

## 12. The audit harness

The same harness that runs in CI is exposed for local use:

```{r harness}
audit_script <- system.file(
  "scripts", "cookbook_audit.R", package = "multisiteDGP"
)
env <- new.env(parent = globalenv())
sys.source(audit_script, envir = env)
env$run_cookbook_audit(execute = FALSE, verbose = FALSE)
```

`execute = FALSE` parses the manifest and confirms every recipe is
discoverable; `execute = TRUE` runs each recipe end-to-end and
matches the printed marker against `expected_marker`. A failed match
is a smoke-test failure — the audit harness is the canonical
definition of "the recipes still work".

## 13. Where to next

The case studies pick up where the recipes stop, and the M-track
vignettes document the four-layer mechanics behind every recipe.

- [A6 · Case study — multisite trial](a6-case-study-multisite.html)
  — full pre-registration flow for a multisite trial: anchor design,
  scenario grid, audit, and methods-appendix template.
- [A7 · Case study — meta-analysis](a7-case-study-meta-analysis.html)
  — meta-analysis analogue: forest, funnel, dependence, and a
  heterogeneity-ratio narrative.
- [M1 · The two-stage DGP](m1-statistical-dgp.html) — formal DGP
  behind every recipe; the canonical-column factorization.
- [M3 · Margin and SE models](m3-margin-se-models.html) — Layer 2
  mechanics (site sizes and standard errors) that drive R5.4 / R5.7 /
  R5.9.
- [M5 · Custom G distributions](m5-custom-g-distributions.html) —
  full detail on the custom-callback contract used in R5.5 and R5.6.
- [M7 · Reproducibility and provenance](m7-reproducibility-provenance.html)
  — hash + provenance machinery behind R5.8.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
