---
title: "M1 · The two-stage DGP — formal specification"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists and reviewers who need the exact specification
  behind every `multisiteDGP` call. This vignette derives the joint
  distribution the package generates from first principles, factors it
  into the four generative layers, lists twelve testable invariants
  that pin the implementation to the specification, and maps every
  symbol used in the package onto the notation of the JEBS paper. You
  leave with a one-page commutative diagram, a kappa-rate identity
  check on real simulated output, and a citation-ready notation table.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M1 The two-stage DGP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. What an analyst sees, what a design encodes

For a multisite trial across $j = 1, \ldots, J$ sites, an analyst who
opens a published table sees a tuple per site:

$$
\mathcal{D}_j \;=\; \bigl(\hat\tau_j, \;\widehat{se}_j, \;n_j, \;\mathbf{x}_j\bigr).
$$

`tau_j_hat` is the site's estimated treatment effect, `se_j` is its
sampling standard error, `n_j` is the within-site sample size, and
$\mathbf{x}_j$ collects optional site-level covariates. The covariate
vector is genuinely optional: `multisiteDGP`'s default `formula = NULL`
omits it.

What the analyst does *not* see — but needs in order to interpret
$\hat\tau_j$ — is the latent site effect $\tau_j$ itself: the value
$\hat\tau_j$ is estimating. The sampling story says $\hat\tau_j$ varies
around $\tau_j$ with magnitude $\widehat{se}_j$. The cross-site story
says the $\tau_j$'s vary around a grand mean $\tau$ with cross-site
standard deviation $\sigma_\tau$. The package generates both stories
simultaneously, and the rest of this vignette specifies exactly how.

Concretely, a `multisiteDGP` simulation tibble has the seven canonical
columns shown below — every Phase 7 vignette uses the same names.

```{r columns-preview}
dat_preview <- sim_multisite(preset_jebs_paper(), seed = 1L)
head(tibble::as_tibble(dat_preview)[, c(
  "site_index", "z_j", "tau_j", "tau_j_hat", "se_j", "se2_j", "n_j"
)], 4)
```

The standardized residual `z_j` and the latent effect `tau_j` are the
two latent quantities the design generates; `tau_j_hat`, `se_j`,
`se2_j`, and `n_j` are the four observable quantities. The
specification's job is to make every conditional and marginal in the
seven-column joint precise.

The data-generating process must therefore specify the full joint

$$
p\bigl(\hat\tau_j, \widehat{se}_j^{\,2}, n_j, \tau_j \mid \mathbf{x}_j, \boldsymbol\theta\bigr),
$$

so that a downstream user can answer at least these five questions
[blueprint Ch. 4 §1]:

1. *What population does $\hat\tau_j$ inform about?* — needs $p(\tau_j)$
   with a defined mean and variance.
2. *How does precision vary across sites?* — needs $p(\widehat{se}_j^{\,2} \mid n_j)$
   together with $p(n_j)$.
3. *What is the sampling distribution of $\hat\tau_j$ given a site's
   true effect and reported precision?* — needs
   $p(\hat\tau_j \mid \tau_j, \widehat{se}_j^{\,2})$.
4. *Do covariates explain part of the latent effect?* — needs
   $p(\tau_j \mid \mathbf{x}_j, \cdot)$.
5. *Is precision systematically related to the latent effect?* —
   needs the joint structure of $(\tau_j, \widehat{se}_j^{\,2})$.

The four-layer factorization in Section 4 is one natural way to answer
all five questions in a single object. Stages 1 and 2 in the next two
sections are the conceptual grouping that motivates that factorization.

## 2. Stage 1 — the sampling distribution

Conditional on the latent site effect $\tau_j$ and on the reported
sampling variance $\widehat{se}_j^{\,2}$, the observed estimate is
generated as a Gaussian draw:

$$
\hat\tau_j \mid \tau_j, \widehat{se}_j^{\,2}
\;\sim\; \mathcal{N}\bigl(\tau_j, \;\widehat{se}_j^{\,2}\bigr).
\tag{Stage 1}
$$

Two points fix the convention. First, the second argument of
$\mathcal{N}(\mu, \sigma^2)$ in this package is the *variance*, never
the standard deviation; every formula in this vignette obeys that
convention. Second, the Gaussian form is justified by within-site
randomization plus a central-limit argument on the difference-in-means
estimator $\hat\tau_j = \bar Y_{j,1} - \bar Y_{j,0}$ when the per-arm
sample sizes are moderate and the outcome has a finite fourth moment
(blueprint Ch. 4 §2.1). The package treats $\widehat{se}_j^{\,2}$ as
*known* (no measurement-error layer on top of it) — this matches the
standard meta-analytic convention in which reported standard errors are
plugged in [@walters2024empirical].

The implementation of this stage lives in
[`gen_observations()`](../reference/gen_observations.html), which is
the Layer 4 generator in the package's four-layer factorization.
Section 4.4 below makes that connection explicit.

The unbiasedness implied by Stage 1 is the package's testable
invariant T1 (Section 7).

### 2.1 Stage 1 as a picture

A small simulation makes the conditional concrete. Below, 30 sites are
drawn from a Gaussian-shape design with $\sigma_\tau = 0.25$, then the
observed estimates $\hat\tau_j$ are overlaid on the latent effects
$\tau_j$ on the same axis. The vertical bars give the $\pm 1.96 \cdot
\widehat{se}_j$ Gaussian envelope from the Stage-1 conditional.

```{r stage1-build}
set.seed(1L)
design_M1 <- multisitedgp_design(
  J            = 30L,
  sigma_tau    = 0.25,
  true_dist    = "Gaussian",
  nj_mean      = 80,
  cv           = 0.5,
  nj_min       = 5,
  p            = 0.5,
  R2           = 0,
  var_outcome  = 1,
  dependence   = "none"
)
dat_M1 <- sim_multisite(design_M1, seed = 1L)
dat_M1
```

The print method's diagnostic block confirms the realized
between-site SD landed near the target 0.25, and that the
sampling-variance product $\widehat{se}_j^{\,2} \cdot n_j$ is constant
across sites (this is the defining identity of the *site-size-driven*
path — see Section 5.2).

```{r stage1-plot, fig.cap = "Stage 1 — observed estimates (filled dots) sit around their latent effects (open circles), with sampling envelopes scaled by site-specific se_j. Vertical spread of bar lengths reflects the precision heterogeneity from variation in n_j (range 25 to 139); the overlap between observed and latent values shrinks for the largest sites (narrowest bars). Sampling error is centered on tau_j, confirming the unbiasedness contract of T1.", fig.alt = "Forest-style plot of 30 simulated sites: open circles mark latent tau_j ordered along the y-axis, filled dots mark observed tau_j_hat, and horizontal bars show the 95 percent Stage-1 sampling envelope plus or minus 1.96 times se_j around tau_j_hat."}
df_stage1 <- tibble::tibble(
  site = seq_len(nrow(dat_M1)),
  tau_j     = dat_M1$tau_j,
  tau_hat_j = dat_M1$tau_j_hat,
  se_j      = dat_M1$se_j
)
df_stage1 <- df_stage1[order(df_stage1$tau_j), ]
df_stage1$rank <- seq_len(nrow(df_stage1))

ggplot(df_stage1, aes(y = rank)) +
  geom_segment(aes(x    = tau_hat_j - 1.96 * se_j,
                   xend = tau_hat_j + 1.96 * se_j,
                   yend = rank),
               colour = "grey70", linewidth = 0.4) +
  geom_point(aes(x = tau_j),     shape = 1, size = 2, colour = "black") +
  geom_point(aes(x = tau_hat_j), shape = 16, size = 1.6, colour = "#1B4965") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(breaks = NULL) +
  labs(
    x        = expression(tau[j]~"  /  "~hat(tau)[j]),
    y        = "Sites (sorted by tau_j)",
    title    = "Stage 1 — sampling envelope around latent effect",
    subtitle = "Open = tau_j (latent) · Filled = tau_j_hat (observed) · Bar = +/- 1.96 * se_j"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

Three observations to read off the figure. First, the filled dots
scatter symmetrically around the open circles — consistent with
$E[\hat\tau_j \mid \tau_j] = \tau_j$ (T1). Second, bar widths vary
substantially across sites — this is precision heterogeneity, set in
Stage 2 by the site-size module (Section 4.2). Third, the bar widths
are not correlated with vertical position in the figure — consistent
with the design's `dependence = "none"` choice (T7).

## 3. Stage 2 — the generative model on $(\tau_j, \widehat{se}_j^{\,2})$

Stage 1 takes $\tau_j$ and $\widehat{se}_j^{\,2}$ as inputs. Stage 2
is where they *come from*. The Stage-2 joint factors as a marginal on
the latent effect, a precision module, and an explicit dependence
coupling:

$$
p\bigl(\tau_j, \widehat{se}_j^{\,2} \mid \mathbf{x}_j, \boldsymbol\theta\bigr)
= \underbrace{p\bigl(\tau_j \mid \mathbf{x}_j, \boldsymbol\theta_G\bigr)}_{\text{latent effect}}
\;\cdot\;
\underbrace{p\bigl(\widehat{se}_j^{\,2} \mid n_j, \kappa\bigr)
\cdot p\bigl(n_j \mid \bar n_j, \mathrm{CV}, n_{j,\min}\bigr)}_{\text{precision via site size}}
\;\cdot\;
\underbrace{c\bigl(\tau_j, \widehat{se}_j^{\,2}\bigr)}_{\text{coupling}}.
$$

The default — and what Stage-1 unbiasedness reduces the package to in
the absence of any dependence injection — is $c(\cdot, \cdot) \equiv 1$,
i.e. $\tau_j$ and $\widehat{se}_j^{\,2}$ are independent given the
design parameters [blueprint Ch. 4 §2.5]. The package allows that
independence to be broken via a post-hoc rearrangement step (Section
5); this is the package's *precision dependence* feature, kept
optional and off by default so that the JEBS-paper specification is
recovered exactly when the user does not opt in.

The next section makes the four pieces of this Stage-2 joint into
four named layers and gives each one its conditional distribution.

## 4. The four generative layers

The full joint, written across all of Stages 1 and 2, factors into
four layers in a fixed order: latent effects → site-size margins →
precision dependence → observation draws. Each layer corresponds to
one named generator family in the package, and one section of this
vignette below.

$$
\underbrace{p\bigl(\tau_j \mid \tau, \sigma_\tau, \boldsymbol\theta_G, \mathbf{x}_j, \boldsymbol\beta\bigr)}_{\text{Layer 1 (latent effects)}}
\;\cdot\;
\underbrace{p\bigl(n_j \mid \bar n_j, \mathrm{CV}, n_{j,\min}\bigr) \cdot p\bigl(\widehat{se}_j^{\,2} \mid n_j, \kappa\bigr)}_{\text{Layer 2 (site-size margins)}}
\;\cdot\;
\underbrace{c\bigl(\tau_j, \widehat{se}_j^{\,2}\bigr)}_{\text{Layer 3 (precision dependence)}}
\;\cdot\;
\underbrace{p\bigl(\hat\tau_j \mid \tau_j, \widehat{se}_j^{\,2}\bigr)}_{\text{Layer 4 (observation draws)}}.
$$

The order is operational: Layer 1 fills `z_j` and `tau_j`, Layer 2
fills `n_j` and `se2_j` (and therefore `se_j`), Layer 3 (when active)
permutes the precisions to induce a target rank correlation with the
effects, and Layer 4 fills `tau_j_hat`. The four layers are also the
four S3 generators
[`gen_effects()`](../reference/gen_effects.html),
[`gen_site_sizes()`](../reference/gen_site_sizes.html),
[`align_rank_corr()`](../reference/align_rank_corr.html) (and its
copula and hybrid siblings), and
[`gen_observations()`](../reference/gen_observations.html).

### 4.1 Layer 1 — latent effects

Site labels carry no information by design — they are interchangeable.
de Finetti's exchangeability theorem then implies that the latent
effects are conditionally i.i.d. from some mixing distribution $G$:

$$
\tau_j \mid \tau, \sigma_\tau, \boldsymbol\theta_G
\;\stackrel{\text{i.i.d.}}{\sim}\; G\bigl(\,\cdot\,;\, \tau, \sigma_\tau, \boldsymbol\theta_G\bigr),
\qquad E[\tau_j] = \tau, \;\; \mathrm{Var}(\tau_j) = \sigma_\tau^{\,2}.
$$

This corresponds to JEBS Equation (2) when $G$ is taken to be Gaussian
[@lee2025improving]. The package implements eight choices for $G$ —
Gaussian, two-component mixture, asymmetric Laplace
[@yu2005three], Student-t, skew-normal [@azzalini1985class],
point-mass-plus-slab, a user-supplied callback, and a Dirichlet
process mixture stub. The catalog and standardization details live in
[M2: G-distribution catalog](m2-g-distribution-catalog.html); for the
purposes of M1, the only contract that matters is that every member
of the catalog has mean $\tau$ and variance $\sigma_\tau^{\,2}$ on the
output scale.

The package factors $\tau_j$ as a location plus a scaled standardized
residual:

$$
\tau_j \;=\; \tau + \sigma_\tau \, z_j,
\qquad z_j \sim G_0, \;\; E[z_j] = 0, \;\; \mathrm{Var}(z_j) = 1,
$$

where $G_0$ is the standardized version of the chosen $G$ shape. The
column `z_j` in the simulation tibble is exactly this standardized
residual. Var $z_j = 1$ is the package's testable invariant T2
(Section 7), and writing the latent effect this way is what makes the
covariate-augmented form

$$
\tau_j \;=\; \tau + \mathbf{x}_j^{\!\top}\boldsymbol\beta + \sigma_\tau\, z_j
$$

a transparent extension rather than a reparameterization. Under
covariates, $\sigma_\tau$ is the *residual* between-site SD — the
marginal SD also picks up $\mathrm{Var}(\mathbf{x}_j^{\!\top}\boldsymbol\beta)$.
This is why
[`print.multisitedgp_data()`](../reference/print.multisitedgp_data.html)
shows two $\sigma_\tau$ values when covariates are active. The
covariate-augmented case is the subject of
[A4: Covariates and precision dependence](a4-covariates-precision-dependence.html);
the covariate-off identity $\sigma_\tau^{\text{resid}} =
\sigma_\tau^{\text{marg}}$ is the package's testable invariant T6.

### 4.2 Layer 2 — site-size margins

Two pieces. First, site sample sizes $n_j$ are drawn from a truncated
Gamma distribution moment-matched to the user-supplied mean $\bar n_j$
and coefficient of variation $\mathrm{CV}$, with a hard lower bound at
$n_{j,\min}$ [blueprint Ch. 4 §2.4]:

$$
n_j \;\stackrel{\text{i.i.d.}}{\sim}\; \mathrm{TruncGamma}\bigl(\bar n_j,\, \mathrm{CV};\; n_{j,\min}\bigr).
$$

The default $n_{j,\min} = 5$ is enforced as the package's testable
invariant T10 (Section 7); the truncation is the reason no
multisite simulation ever produces an $n_j$ smaller than 5 by
default. The truncated-Gamma form is the modern Engine A2; the legacy
JEBS-appendix Engine A1 uses a censor-and-round form instead, and the
two engines exist precisely so that the JEBS appendix is reproducible
bit-for-bit (T1a) without giving up modern moment-matching (T1b). See
[M3: Margin and SE models](m3-margin-se-models.html) for the engine
contracts.

Second, conditional on $n_j$, the per-site sampling variance is set
deterministically:

$$
\widehat{se}_j^{\,2} \;=\; \frac{\kappa}{n_j},
$$

with the proportionality constant $\kappa$ derived from a balanced
two-arm randomization with allocation share $p$, residual outcome
variance $s_Y^2$, and covariate-explained share $R^2$:

$$
\kappa \;=\; \left(\frac{1}{p} + \frac{1}{1-p}\right)(1 - R^2)\, s_Y^2.
$$

At the package defaults $p = 0.5$, $R^2 = 0$, $s_Y^2 = 1$, this
collapses to $\kappa = 4$. The deterministic form is what makes the
identity $\widehat{se}_j^{\,2} \cdot n_j = \kappa$ hold for every
site in every simulation under the *site-size-driven* path — this is
the package's testable invariant T3 (Section 7) and the defining
contract of the path itself. A minute of code makes it visible:

```{r kappa-check}
range(dat_M1$se2_j * dat_M1$n_j)
unique(round(dat_M1$se2_j * dat_M1$n_j, 8))
compute_kappa(p = 0.5, R2 = 0, var_outcome = 1)
```

The first line shows the kappa-rate product is constant. The third
line confirms the closed form returns 4.

The alternative *direct-precision* path (`gen_se_direct()`) discards
the deterministic kappa identity in exchange for direct control of
the informativeness scalar $I$ and the precision heterogeneity ratio
$R$; that path is the subject of M3 §3 and is outside the scope of
this vignette except as part of the JEBS notation table in Section 6.

### 4.3 Layer 3 — precision dependence

By default, $\tau_j$ and $\widehat{se}_j^{\,2}$ are independent
because they are generated from independent random draws. When a user
asks for a non-zero target Spearman or Pearson correlation, the
package performs a *post-hoc rearrangement* of the precisions so that
their multiset is preserved exactly while their pairing with effects
matches the target.

Three coupling functions are exposed:
[`align_rank_corr()`](../reference/align_rank_corr.html) (rank
permutation),
[`align_copula_corr()`](../reference/align_copula_corr.html) (Gaussian
copula re-pairing), and
[`align_hybrid_corr()`](../reference/align_hybrid_corr.html)
(copula initialization plus rank refinement). The methodological
contrast across the three lives in
[M4: Precision dependence theory](m4-precision-dependence-theory.html).

The contracts that matter for M1 are:

- When `dependence = "none"`, the coupling reduces to the identity
  permutation (T7). No rearrangement, no change to either marginal.
- The marginal of $\tau_j$ is preserved exactly (its mean, variance,
  and shape are not touched).
- The multiset of $\widehat{se}_j^{\,2}$ is preserved exactly (its
  realized values are not touched, only their pairing with sites is).
- The realized informativeness $I$ is preserved exactly across any
  active rearrangement.

The reason the package generates first and rearranges second — rather
than building dependence into a joint distribution upfront — is to
keep the user's specification of the marginals from drifting. T7 is
the testable contract that the rearrangement is genuinely identity
when the user opts out.

### 4.4 Layer 4 — observation draws

Layer 4 is the implementation of Stage 1 from Section 2: given the
latent effects produced by Layer 1 and the sampling variances produced
by Layer 2 (possibly rearranged in Layer 3), draw

$$
\hat\tau_j \;\sim\; \mathcal{N}\bigl(\tau_j, \;\widehat{se}_j^{\,2}\bigr),
\qquad j = 1, \ldots, J,
$$

independently across $j$. The implementation is
[`gen_observations()`](../reference/gen_observations.html); the
column `tau_j_hat` in the simulation tibble is its output.

Independence of the per-site $\hat\tau_j$ draws (conditional on Layer
1–3 outputs) is part of the same blueprint Ch. 4 §2.5 default. It is
not a separate testable invariant — it is the *defining* form of the
observation generator. T1 in Section 7 covers the unbiasedness
contract; the conditional independence of the draws is structural.

### 4.5 The Stage-2 joint as a picture

The default joint $p(\tau_j, \widehat{se}_j^{\,2})$ — independent
marginals, no dependence coupling — looks like this. The two density
panels show the marginal of each axis, and the central scatter shows
the realized pairs.

```{r stage2-build, results = "hide"}
set.seed(2L)
design_marg <- multisitedgp_design(
  J            = 200L,
  sigma_tau    = 0.25,
  true_dist    = "Gaussian",
  nj_mean      = 60,
  cv           = 0.5,
  nj_min       = 5,
  p            = 0.5,
  R2           = 0,
  var_outcome  = 1,
  dependence   = "none"
)
dat_marg <- sim_multisite(design_marg, seed = 2L)
```

```{r stage2-marginals}
sd_tau    <- round(sd(dat_marg$tau_j), 4)
gm_se2    <- round(exp(mean(log(dat_marg$se2_j))), 4)
range_se2 <- range(dat_marg$se2_j)
sd_tau; gm_se2; range_se2
```

```{r stage2-plot, fig.height = 5.6, fig.cap = "Stage 2 default joint p(tau_j, se_j squared) at J = 200. Top: marginal density of latent effects, centered near 0 with SD near sigma_tau = 0.25 (T2). Right: marginal density of sampling variances, right-skewed because n_j is right-skewed and se squared = kappa over n_j (T3). Center: scatter of realized (tau_j, se squared) pairs is roughly cloud-shaped, with no visible tilt — independence at the sample level, consistent with T7.", fig.alt = "Composite figure with three panels: a top density of tau_j, a right-side density of se squared, and a central scatter of (tau_j, se_j squared) pairs."}
df_marg <- tibble::tibble(
  tau_j  = dat_marg$tau_j,
  se2_j  = dat_marg$se2_j
)

p_center <- ggplot(df_marg, aes(x = tau_j, y = se2_j)) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#1B4965") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x = expression(tau[j]~"  (latent effect)"),
    y = expression(widehat(se)[j]^{2}~"  (sampling variance)")
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p_top <- ggplot(df_marg, aes(x = tau_j)) +
  geom_density(fill = "#1B4965", alpha = 0.35, colour = "#1B4965") +
  labs(x = NULL, y = "density") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x  = element_blank(),
        axis.ticks.x = element_blank())

p_right <- ggplot(df_marg, aes(x = se2_j)) +
  geom_density(fill = "#62929E", alpha = 0.45, colour = "#62929E") +
  coord_flip() +
  labs(x = NULL, y = "density") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank())

# Compose without depending on patchwork (soft dep).
gridExtra::grid.arrange(
  p_top,                                grid::nullGrob(),
  p_center,                             p_right,
  ncol = 2, widths = c(4, 1), heights = c(1, 4)
)
```

Three things to read off the joint:

- The top panel's density is centered near zero and has spread close
  to $\sigma_\tau = 0.25$. This is T2 in action: the standardized
  residual has unit variance, so $\tau_j$'s marginal SD lands at
  $\sigma_\tau$.
- The right panel's density is right-skewed. That is the deterministic
  consequence of $\widehat{se}_j^{\,2} = \kappa / n_j$ (T3) plus the
  truncated-Gamma marginal on $n_j$: large $n_j$ pile up near the
  mean, but small $n_j$ pull the SE-squared distribution out to the
  right.
- The central scatter is roughly cloud-shaped. There is no systematic
  tilt — the realized rank correlation between $\tau_j$ and
  $\widehat{se}_j^{\,2}$ is statistically indistinguishable from zero
  at $J = 200$. This is T7 in action.

## 5. Default-mode reduction

When every option is at its default value, the package reduces to a
short, vanilla two-stage random-effects model. Defaults are
`true_dist = "Gaussian"`, `formula = NULL`, `dependence = "none"`,
the *site-size-driven* path (Paradigm A in the blueprint), and
$p = 0.5$, $R^2 = 0$, $s_Y^2 = 1$ giving $\kappa = 4$:

$$
\boxed{
\begin{aligned}
n_j &\;\stackrel{\text{i.i.d.}}{\sim}\; \mathrm{TruncGamma}(\bar n_j, \mathrm{CV};\, n_{j,\min} = 5),\\
\widehat{se}_j^{\,2} &= 4 / n_j,\\
\tau_j &\;\stackrel{\text{i.i.d.}}{\sim}\; \mathcal{N}(\tau, \sigma_\tau^{\,2}),\\
\hat\tau_j \mid \tau_j, \widehat{se}_j^{\,2} &\;\sim\; \mathcal{N}\bigl(\tau_j, \widehat{se}_j^{\,2}\bigr).
\end{aligned}}
$$

This is the *site-size-driven* path's default form; the JEBS-paper
specification is contained inside it as a special case
[@lee2025improving]. Section 6 maps every symbol to JEBS notation,
and Section 7 lists the twelve invariants the package carries.

### 5.1 Two paths through Stage 2

The blueprint distinguishes two ways to set the precision marginal.
The *site-size-driven* path (Paradigm A in the blueprint) is what
Section 4.2 derived: a user supplies $\bar n_j$, $\mathrm{CV}$, and
$\kappa$; sites sizes $n_j$ are drawn; precisions follow as
$\widehat{se}_j^{\,2} = \kappa / n_j$. The *direct-precision* path
(Paradigm B in the blueprint) inverts: a user supplies the
informativeness scalar $I$ and the precision heterogeneity ratio $R$,
and the precision marginal is reconstructed (log-evenly spaced) to
match. The default is the *site-size-driven* path; the
*direct-precision* path is the subject of
[M3: Margin and SE models](m3-margin-se-models.html). The notation
table in Section 6 covers symbols from both paths.

### 5.2 Three deviations from the default

Three knobs, each with a default that recovers the JEBS-paper
specification, expand the package beyond the JEBS scope:

- **Covariate-augmented Stage 2** — passing a `formula` with `data`
  and `beta` introduces the linear term $\mathbf{x}_j^{\!\top}
  \boldsymbol\beta$ (Section 4.1). With `formula = NULL`, the
  unconditional and covariate-augmented code paths are identical to
  floating-point precision (T5).
- **Precision dependence** — setting `dependence` to `rank`, `copula`,
  or `hybrid` activates Layer 3's rearrangement (Section 4.3). With
  `dependence = "none"`, the rearrangement reduces to identity (T7).
- **G-shape choice** — setting `true_dist` to one of the seven
  non-Gaussian shapes (Section 4.1) selects from
  [M2's catalog](m2-g-distribution-catalog.html). With
  `true_dist = "Gaussian"`, the latent effects follow JEBS Equation
  (2) exactly.

## 6. JEBS notation map

This section maps every symbol the package uses onto the notation of
the JEBS paper [@lee2025improving]. The table is split into three
parts: equations from the body of the JEBS paper, simulation-design
formulas from the JEBS appendix code, and package-only extensions
that have no JEBS counterpart.

### 6.1 JEBS paper body equations

| JEBS reference | Symbol / equation | Package mapping | Notes |
|----------------|---------------------------------|---------------------|-------|
| Stage-1 setup (unnumbered) | $\hat\tau_j \mid \tau_j, \widehat{se}_j^{\,2} \sim \mathcal{N}(\tau_j, \widehat{se}_j^{\,2})$ | [`gen_observations()`](../reference/gen_observations.html) | Justified by within-site randomization + CLT |
| Eq. (1) | $\widehat{se}_j^{\,2} = \bigl(\tfrac{1}{p_j} + \tfrac{1}{1-p_j}\bigr) s_j^{\,2} (1 - R^2) / n_j$ | `compute_kappa()` (closed form) | Per-site SE under balanced randomization |
| Eq. (2) | $\tau_j \mid \tau, \sigma^2 \sim \mathcal{N}(\tau, \sigma^2)$ | [`gen_effects()`](../reference/gen_effects.html) with `true_dist = "Gaussian"` | Default G shape; mean $\tau$, variance $\sigma_\tau^{\,2}$ |
| Eq. (3) | Posterior shrinkage estimator | *(out of DGP scope)* | Estimator side; covered in `siteBayes2`, not `multisiteDGP` |
| Eq. (4) | $I = \sigma^2 / (\sigma^2 + \exp(\overline{\ln \widehat{se}_j^{\,2}}))$ | [`informativeness()`](../reference/informativeness.html) | Generative-side diagnostic |

### 6.2 JEBS appendix simulation formulas

| Appendix reference | Symbol / construct | Package mapping | Notes |
|--------------------|---------------------------|---------------------|-------|
| Appendix mixture form | Two-component mixture with parameters $\delta, \varepsilon, \nu$ | [`gen_effects()`](../reference/gen_effects.html) with `true_dist = "Mixture"` | Bit-identical with appendix code under Engine A1 (T1a) |
| Appendix site-size + SE form | $n_j$ Gamma + censor + round; $\widehat{se}_j^{\,2} = \kappa / n_j$ | [`gen_site_sizes()`](../reference/gen_site_sizes.html) Engine A1 | T1a fixture |
| Appendix observation form | $\hat\tau_j \sim \mathcal{N}(\tau_j, \widehat{se}_j^{\,2})$ + row shuffle | [`gen_observations()`](../reference/gen_observations.html) Engine A1 path | Order of RNG calls is itself an invariant (T12) |
| Appendix default | $p = 0.5$, $R^2 = 0$, $s_Y^2 = 1$ → $\kappa = 4$ | [`multisitedgp_design()`](../reference/multisitedgp_design.html) defaults | T2/T3 fixture |
| Appendix G shapes | Gaussian, two-component mixture, asymmetric Laplace | [`gen_effects()`](../reference/gen_effects.html) `true_dist` ∈ {Gaussian, Mixture, ALD} | Three of the package's eight shapes |
| Appendix site-size marginal | $n_j \sim \mathrm{TruncGamma}(\bar n_j, \mathrm{CV} \cdot \bar n_j; n_{j,\min} = 5)$ | [`gen_site_sizes()`](../reference/gen_site_sizes.html) Engine A2 | T10 fixture |

### 6.3 Package-only extensions

| Feature | Symbol / construct | Package mapping | Notes |
|--------------|----------------------------|----------------------|-------|
| Site-level covariates | $\tau_j = \tau + \mathbf{x}_j^{\!\top} \boldsymbol\beta + \sigma_\tau z_j$ | [`gen_effects()`](../reference/gen_effects.html) with `formula`, `beta`, `data` | Reduces to JEBS Eq. (2) when `formula = NULL` (T5) |
| Precision dependence (rank) | Target Spearman $\rho_S$ between $\tau_j$ (or $z_j$) and $\widehat{se}_j^{\,2}$ | [`align_rank_corr()`](../reference/align_rank_corr.html) | Identity permutation when `dependence = "none"` (T7) |
| Precision dependence (copula) | Target Pearson $\rho_P$ via Gaussian copula | [`align_copula_corr()`](../reference/align_copula_corr.html) | M4 vignette covers selection criteria |
| Precision dependence (hybrid) | Copula initialization + rank refinement | [`align_hybrid_corr()`](../reference/align_hybrid_corr.html) | M4 vignette covers when each path is appropriate |
| Direct-precision path | Inverse design from $(I, R)$ to $\widehat{se}_j^{\,2}$ marginal | [`gen_se_direct()`](../reference/gen_se_direct.html) | M3 vignette covers feasibility region |
| Five new G shapes | Student-t, skew-normal [@azzalini1985class], point-mass-plus-slab, user, DPM stub | [`gen_effects()`](../reference/gen_effects.html) `true_dist` ∈ {StudentT, SkewN, PointMassSlab, User, DPM} | Five of the package's eight shapes; `Gaussian` default preserves JEBS Eq. (2) |

## 7. Twelve testable invariants

The specification in Sections 1–4 is only as precise as the
implementation it pins down. Twelve invariants, each a one-line
contract on observable output, form the testable backbone of this
correspondence — every one of them lives in
`tests/testthat/test-jebs-parity.R` and surrounding test files in the
package source. The numbering follows the blueprint Ch. 4 §7 listing,
with T1 split into T1a (Engine A1, bit-identical with the JEBS
appendix) and T1b (Engine A2, distributional parity) per the
blueprint's R1.4 fix.

### T1a — Engine A1 bit-identical parity (mixture)

Engine A1 (legacy, `legacy_jebs_censor_round`) under the JEBS
appendix's exact configuration — `true_dist = "Mixture"` with
$(\delta, \varepsilon, \nu) = (5, 0.3, 2)$, $J = 100$, $\bar n_j = 80$,
$\mathrm{CV} = 0.5$, $n_{j,\min} = 4$, `formula = NULL`,
`dependence = "none"` — produces a simulation tibble whose
xxhash64 digest matches the digest of the JEBS appendix pipeline
output across the seed set $\{42, 1, 2024, 12345\}$. *Bit-identical*
means floating-point byte equality, not distributional similarity.

### T1b — Engine A2 distributional parity (Gaussian)

Engine A2 (modern, `trunc_gamma_moment`) with `true_dist = "Gaussian"`
and the package defaults ($J = 50$, $\sigma_\tau = 0.25$, $\bar n_j =
80$, $\mathrm{CV} = 0.5$, $n_{j,\min} = 5$, $p = 0.5$, $R^2 = 0$,
$s_Y^2 = 1$) produces $\tau_j$ samples whose empirical mean is within
$2\sigma_\tau / \sqrt{J}$ of $\tau$, whose empirical variance is
within $2\sigma_\tau^{\,2} / \sqrt{J}$ of $\sigma_\tau^{\,2}$, and
whose KS test against $\mathcal{N}(\tau, \sigma_\tau^{\,2})$ passes
at the 0.05 level in at least 95 percent of $M = 1000$ Monte Carlo
replications. *Distributional* parity, not bit-identical.

### T2 — kappa is exactly 4 at defaults

At $p = 0.5$, $R^2 = 0$, $s_Y^2 = 1$, the realized $\kappa$ equals 4
exactly (no rounding). This is the bridge from the closed form
$\kappa = (1/p + 1/(1-p))(1 - R^2) s_Y^2$ to the package output the
user sees. Verified directly:

```{r T2-check}
compute_kappa(p = 0.5, R2 = 0, var_outcome = 1)
identical(compute_kappa(p = 0.5, R2 = 0, var_outcome = 1), 4)
```

### T3 — kappa-rate identity holds for every site

Under the *site-size-driven* path, $\widehat{se}_j^{\,2} \cdot n_j =
\kappa$ for every $j$ to floating-point precision. Equivalently, the
within-design variance of $n_j \cdot \widehat{se}_j^{\,2}$ across $j$
is zero. The check on `dat_M1` from Section 4.2 is the operational
form of this contract.

### T4 — Stage-2 independence under `dependence = "none"`

When `dependence = "none"`, the realized Spearman and Pearson
correlations between $\tau_j$ and $\widehat{se}_j^{\,2}$ are
statistically indistinguishable from zero. The acceptance band is the
empirical 95th percentile of $|\hat\rho_S|$ from 1000 random
independent permutations (a *permutation envelope*) — not the Fisher-z
asymptotic critical value, which is a Pearson approximation that is
tight but inexact for Spearman. The blueprint corrected an earlier
draft on this point; the package follows the permutation form.

### T5 — covariate-off equivalence

When `formula = NULL` (or when the user passes `beta = 0`), the
covariate-augmented code path produces output identical to the
unconditional path to floating-point precision ($\epsilon = 10^{-12}$).
This is what makes the package's covariate handling a strict extension
rather than a reparameterization.

### T6 — residual equals marginal SD when no covariates

When `formula = NULL`, the `print.multisitedgp_data()` method reports
`sigma_tau_resid` and `sigma_tau_marg` as numerically identical, and
both equal the input `sigma_tau` to floating-point precision. With
covariates, the marginal SD picks up
$\mathrm{Var}(\mathbf{x}_j^{\!\top}\boldsymbol\beta)$ and the two
quantities diverge — see [A4](a4-covariates-precision-dependence.html).

### T7 — precision dependence reduces to identity

When `dependence = "none"`, the alignment functions
[`align_rank_corr()`](../reference/align_rank_corr.html),
[`align_copula_corr()`](../reference/align_copula_corr.html), and
[`align_hybrid_corr()`](../reference/align_hybrid_corr.html) all
reduce to the identity permutation: input vector equals output
vector, exactly. This is the operational form of the "off by default"
contract on Layer 3.

### T8 — informativeness formula

[`informativeness()`](../reference/informativeness.html) implements
JEBS Equation (4) verbatim:

$$
I \;=\; \frac{\sigma_\tau^{\,2}}{\sigma_\tau^{\,2} + \exp\bigl(\overline{\ln \widehat{se}_j^{\,2}}\bigr)},
$$

where the bar denotes the arithmetic mean over $j = 1, \ldots, J$.
Section 6.1 entry for Eq. (4) is the same statement.

### T9 — default G shape is Gaussian

When `true_dist` is omitted, the package defaults to `"Gaussian"`. At
$J \in \{50, 200\}$ across $M = 1000$ Monte Carlo replications, the
Shapiro-Wilk test against $\mathcal{N}(\tau, \sigma_\tau^{\,2})$
rejects at the 0.01 level a count consistent with the binomial
acceptance interval — concretely, the count of $p \le 0.01$ events
falls in $[\mathrm{qbinom}(0.025, M, 0.01),\, \mathrm{qbinom}(0.975, M, 0.01)] = [4, 17]$
for $M = 1000$. The acceptance band acknowledges that under a true
normal, the Shapiro-Wilk $p$-value is uniform, so the count of small
$p$-values is itself a binomial random variable.

### T10 — site-size truncation enforces $n_{j,\min}$

At $n_{j,\min} = 5$, no $n_j < 5$ ever appears in the simulation
output, and the realized mean $\bar n_j$ is within $\pm 0.5$ percent
of the requested mean. The truncation is hard, not soft, in both
engines.

### T11 — informativeness range matches JEBS over the design grid

Across the JEBS simulation grid — $J \in \{25, 50, 100, 200, 300\}$,
$\bar n_j \in \{10, 20, 40, 80, 160\}$, $\mathrm{CV} \in \{0, 0.25,
0.50, 0.75\}$, $\sigma_\tau \in \{0.05, 0.10, 0.15, 0.20, 0.25\}$
giving $5 \times 5 \times 4 \times 5 = 500$ design cells — the
realized informativeness values cover the JEBS-reported range
$I \in [0.01, 0.71]$ (Lee et al. 2025, p. 13). The full grid is the
contract; a default-CV slice is not sufficient.

### T12 — Engine A1 RNG sequence is locked

Under Engine A1, the random-number-generator call sequence is exactly:
`rgamma(J)` for site sizes, `rnorm(J)` for the first noise draw, then
the row-shuffle internal `sample()` call from
`dplyr::slice_sample(prop = 1)`, then `rnorm(J)` for any subsequent
noise. No additional RNG calls. Reordering, omitting the
`slice_sample` shuffle, or re-seeding mid-pipeline all violate T12 and
break T1a parity simultaneously. Engine A2's RNG sequence is separate
and is the subject of T20 in the blueprint (out of M1 scope).

### How the invariants relate to the four layers

A short cross-reference, for orientation:

| Invariant | Layer | What it pins |
|-----------|-------|--------------|
| T1, T1a, T1b | Layer 4 (and Engine A1/A2 pipelines end-to-end) | Stage 1 unbiasedness; appendix bit-equivalence; modern distributional parity |
| T2, T3 | Layer 2 | $\kappa = 4$ at defaults; deterministic kappa-rate identity |
| T4, T7 | Layer 3 | Independence under `dependence = "none"`; identity permutation contract |
| T5, T6 | Layer 1 | Covariate-off equivalence; residual-equals-marginal SD |
| T8, T11 | Cross-layer (diagnostics) | Informativeness formula; informativeness range |
| T9 | Layer 1 | Default G shape is Gaussian |
| T10 | Layer 2 | Site-size truncation |
| T12 | Engine A1 | RNG call order locked to JEBS appendix |

### Forward pointers — T13–T20 in companion vignettes

T1–T12 above are the invariants that fit naturally inside the
two-stage DGP specification this vignette covers. The
blueprint's full T-numbered list extends further; the remaining
invariants live with the layer they constrain:

- **T13–T17 — per-shape unit-variance standardization** — checks
  that ALD, Mixture, Student-$t$, skew-normal, and point-mass-slab
  generators all return $\mathrm{Var}(z_j) = 1$ at the package's
  closed-form rescalers. Operationalized in
  [M2: G-distribution catalog and standardization](m2-g-distribution-catalog.html)
  §4 (sanity checks) and the package's $J = 10^6$ test suite.
- **T18–T19 — covariate decomposition + Spearman tie handling** —
  identifiability of the residual / marginal split when covariates
  enter Layer 1, and the rank-correlation behavior at tied
  precision values. Operationalized in
  [A4: Covariates and precision dependence](a4-covariates-precision-dependence.html)
  for the residual-vs-marginal demonstration and in
  [M4: Precision dependence theory](m4-precision-dependence-theory.html)
  §3.2 for the rank ties.
- **T20 — Engine A2 reproducibility** — the modern-engine RNG-call
  sequence (the analogue of T12 for Engine A2) and its hash
  determinism across reruns. Operationalized in
  [M7: Reproducibility and provenance](m7-reproducibility-provenance.html)
  §3 (canonical-hash determinism) and §5 (manifest pattern).

## 8. Limitations and assumptions

The two-stage specification carries seven assumptions that are
explicit in blueprint Ch. 4 §8 and worth surfacing here:

1. *Independence baseline.* By default, $\tau_j \perp
   \widehat{se}_j^{\,2}$. Activating `dependence` breaks this on
   purpose (Section 4.3, [M4](m4-precision-dependence-theory.html)).
2. *Linear covariate effect.* The covariate term enters as
   $\mathbf{x}_j^{\!\top}\boldsymbol\beta$. Nonlinear $\mathbf{x}_j
   \to \tau_j$ relationships are reachable only via a user-supplied
   $G$ callback ([M5](m5-custom-g-distributions.html)).
3. *Finite second moments of $G$.* $\sigma_\tau^{\,2} < \infty$ is
   required. Heavy-tailed Student-t with $\nu \le 2$ is rejected at
   construction.
4. *Within-site Normality of the per-site estimate.* CLT-justified for
   $n_j$ moderately large; for very small sites ($n_j < 10$) the
   approximation is tighter but degrades.
5. *Effect-size scale.* $s_Y^2 = 1$ is the default; changing it
   rescales $\kappa$ multiplicatively.
6. *Single treatment arm with complete randomization within site.*
   Cluster-randomized, blocked, and factorial designs are outside the
   current package surface.
7. *No measurement error in $\widehat{se}_j^{\,2}$.* The package
   treats reported standard errors as exact, matching the standard
   meta-analytic plug-in convention [@walters2024empirical].

## 9. Where to next

- [M2: G-distribution catalog and standardization](m2-g-distribution-catalog.html)
  — the eight choices for $G$, their parameterizations, and how
  standardization is enforced (T2 is operationalized there).
- [M3: Margin and SE models](m3-margin-se-models.html) — Engine A1
  vs. Engine A2, and the *direct-precision* path.
- [M4: Precision dependence theory](m4-precision-dependence-theory.html)
  — the three injection methods and when each is appropriate.
- [M7: Reproducibility and provenance](m7-reproducibility-provenance.html)
  — canonical hashes, RNG-state capture, and how to write a
  defensible methods appendix.
- For a worked applied example that touches every layer,
  [A1: Getting started](a1-getting-started.html); for the diagnostic
  rubric that turns the invariants in Section 7 into PASS / WARN /
  FAIL flags on a single simulation,
  [A3: Diagnostics in practice](a3-diagnostics-in-practice.html).

## References

The foundational data-generating process and the JEBS preset
implement the design in @lee2025improving. Empirical-Bayes context
for site-effect estimation, including the assumption that reported
sampling standard errors are taken as known, is collected in
@walters2024empirical. The asymmetric Laplace shape used in the $G$
catalog is from @yu2005three, and the skew-normal shape is from
@azzalini1985class. The applied calibration benchmarks underlying the
education presets are from @weiss2017howmuch.

The Stage-1 Gaussian-sampling justification (CLT-based on
within-site randomization with finite fourth moments) is the
standard meta-analytic baseline; the package's blueprint Ch. 4 §2.1
gives the formal motivation. Specific canonical references in the
meta-analytic methodology literature are left to the analyst's
choice rather than prescribed here.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
