---
title: "M2 · G-distribution catalog and standardization"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists and reviewers who need to know exactly which
  shape of latent-effect distribution `multisiteDGP` is putting under
  the simulation, and exactly how that shape is rescaled before it
  reaches the rest of the pipeline. This vignette is organized one
  section per shape: each of the eight built-in `G` distributions gets
  its own parameterization, density, unit-variance derivation,
  generator function, and decision rule for when to use it. You leave
  with eight density-and-QQ panels, a sanity table of realized
  `var(z_j)` at `J = 200`, and a decision rubric for picking among
  Student-t, skew-normal, and the asymmetric Laplace.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M2 G-distribution catalog and standardization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why standardization matters

The two-stage data-generating process documented in
[M1: The two-stage DGP](m1-statistical-dgp.html) factors the latent
site effect as

$$
\tau_j \;=\; \tau \;+\; \sigma_\tau \cdot z_j,
\qquad z_j \sim G_0(\boldsymbol\theta_G),
\qquad E[z_j] = 0, \;\; \mathrm{Var}(z_j) = 1.
$$

The factorization splits the cross-site distribution into two
ingredients. The **shape** of $G$ — Gaussian, Student-t, skew-normal,
asymmetric Laplace, two-component mixture, point-mass-plus-slab,
user-supplied callback, or Dirichlet-process-mixture stub — lives in
the standardized residual $z_j$ and is parameterized by
$\boldsymbol\theta_G$. The **scale** of the cross-site spread lives in
the multiplier $\sigma_\tau$. The package keeps the two strictly
separate by enforcing $\mathrm{Var}(z_j) = 1$ inside the generator.
Every shape in the catalog is rescaled to unit variance before it
leaves the Layer-1 generator [blueprint Ch. 5 §0]; the only place
$\sigma_\tau$ is multiplied in is the location-scale wrapper above.

Why this convention matters. Without it, a non-Gaussian shape's
intrinsic variance and the user-supplied $\sigma_\tau$ are confounded:
asking for "Student-$t$ with $\sigma_\tau = 0.25$" can silently
produce cross-site SDs near $0.25 \cdot \sqrt{\nu/(\nu - 2)}$ rather
than $0.25$. The companion JEBS-paper-era package
[siteBayes2](m8-migration-from-siteBayes2.html) had exactly this
double-scaling pattern: the legacy `gen_priorG2` would draw from
shape-specific densities whose own variance was not 1, then multiply
by $\sigma_\tau$ a second time. The fix in `multisiteDGP` is the
unit-variance convention enforced shape-by-shape, with the
standardization done analytically wherever closed forms exist
(Gaussian, Student-t, skew-normal, asymmetric Laplace) and verified
numerically where they do not (the two-component mixture, the
point-mass-plus-slab, and any user callback).

A second payoff matters for empirical work. Because shape and scale
are decoupled, the informativeness scalar
[@lee2025improving] depends on $\boldsymbol\theta_G$ only through
$\sigma_\tau$, so empirical-Bayes shrinkage diagnostics
[@walters2024empirical] compare like-with-like across shapes.

## 2. The catalog at a glance

The eight shapes the package supports, with their `theta_G` fields
and their canonical use cases. The order matches the rest of the
vignette; the exported Layer-1 generator name is given in code font.

| Shape | `true_dist` | `theta_G` fields | Layer-1 generator | Reach for it when |
|-------|---------|-------------------------------------------|------------------------------------------|-------------------------------------------------------------|
| Gaussian | `"Gaussian"` | (none) | [`gen_effects_gaussian()`](../reference/gen_effects_gaussian.html) | You have no prior reason to depart from normality — the JEBS default. |
| Student-t | `"StudentT"` | `nu` (numeric > 2) | [`gen_effects_studentt()`](../reference/gen_effects_studentt.html) | Heavy tails matter (a few sites with extreme effects). |
| Skew-normal | `"SkewN"` | `slant` (numeric) | [`gen_effects_skewn()`](../reference/gen_effects_skewn.html) | One-sided skew is plausible (program-effect tails). |
| ALD | `"ALD"` | `rho` (in (0, 1)) | [`gen_effects_ald()`](../reference/gen_effects_ald.html) | Skewness with sharper peaks; quantile interpretation. |
| Mixture | `"Mixture"` | `delta`, `eps`, `ups` | [`gen_effects_mixture()`](../reference/gen_effects_mixture.html) | Bimodality, contamination, latent classes of sites. |
| Point-mass-plus-slab | `"PointMassSlab"` | `pi0`, plus slab params | [`gen_effects_pmslab()`](../reference/gen_effects_pmslab.html) | A fraction of sites are exactly null (sparse effects). |
| User | `"User"` | (whatever your callback consumes) | [`gen_effects_user()`](../reference/gen_effects_user.html) | None of the seven match — bring your own generator. |
| DPM | `"DPM"` | (deferred) | [`gen_effects_dpm()`](../reference/gen_effects_dpm.html) | Use the `User` bridge — built-in DPM sampling is deferred. |

The first six are the closed-catalog choices. `User` is the escape
hatch — supply a function that returns a length-`J` numeric vector,
and the package treats it as a standardized residual generator
(Convention A — the recommended path; mean 0, variance 1) or as a
final-effect generator (Convention B — the wrapper bypass; less
common). `DPM` is a placeholder: the public surface accepts the
string for forward compatibility, but a built-in Dirichlet-process
sampler is deferred to a later release. Calling
`true_dist = "DPM"` without a `g_fn` bridge is rejected with an
actionable error (see Section 3.8).

## 3. Per-shape sections

Each of the eight subsections below follows the same template:
parameterization, formal density, unit-variance derivation, the
exported Layer-1 generator, when to reach for the shape, and a
density / QQ-against-Gaussian panel. A reader can jump straight to
the shape they need without reading the others — this is the
"organize-by-shape" view; the cross-shape audit lives at Section 4.

### 3.1 Gaussian

**Parameterization.** None: $\boldsymbol\theta_G = \emptyset$. The
shape is fully specified by location $\tau$ and scale $\sigma_\tau$
in the wrapper.

**Density.** $z_j \sim \mathcal{N}(0, 1)$, with density

$$
\phi(z) \;=\; \frac{1}{\sqrt{2\pi}} \exp\bigl(-z^2 / 2\bigr).
$$

**Unit-variance derivation.** Trivial: $E[z_j] = 0$ and
$\mathrm{Var}(z_j) = 1$ by construction. No rescaling.

**Identifiability.** The map $(\tau, \sigma_\tau) \mapsto
\mathcal{N}(\tau, \sigma_\tau^2)$ is one-to-one for $\sigma_\tau > 0$.

**Generator.**
[`gen_effects_gaussian()`](../reference/gen_effects_gaussian.html) —
no shape arguments beyond `J`, `tau`, and `sigma_tau`. Equivalent to
the [`gen_effects()`](../reference/gen_effects.html) entry point with
`true_dist = "Gaussian"` (the default).

**When to reach for it.** This is the JEBS-paper default
[@lee2025improving]. Most applied multisite trials at $J \le 50$
cannot empirically distinguish Gaussian from a mild Student-t or
skew-normal anyway, so the default is rarely wrong as a first pass.

```{r gaussian-call}
set.seed(1L)
out_gaussian <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "Gaussian",
  theta_G   = list()
)
summary(out_gaussian$z_j)
var(out_gaussian$z_j)
```

The realized variance lands close to 1 — the small departure is
sampling fluctuation at $J = 200$, not a standardization defect.

### 3.2 Student-t

**Parameterization.** $\boldsymbol\theta_G = \{\nu\}$ with $\nu > 2$
(degrees of freedom). The strict inequality is the package contract:
the variance of a $t_\nu$ is $\nu / (\nu - 2)$, which diverges at
$\nu = 2$ and is undefined for $\nu \le 2$. Standardizing requires a
finite second moment, so $\nu \le 2$ is rejected at construction.

**Density.** Standard Student-$t$ on $\nu$ degrees of freedom, then
rescaled so the variance equals 1:

$$
T \sim t_\nu, \qquad
z_j \;=\; T \cdot \sqrt{\frac{\nu - 2}{\nu}}.
$$

**Unit-variance derivation.** $\mathrm{Var}(T) = \nu / (\nu - 2)$
implies $\mathrm{Var}(z_j) = (\nu - 2)/\nu \cdot \nu/(\nu - 2) = 1$
exactly.

**Identifiability.** Hard refuse $\nu \le 2$ (variance non-finite);
soft warn at $\nu < 4$ (kurtosis non-finite, sample-variance
estimates unstable on small $J$).

**Generator.**
[`gen_effects_studentt()`](../reference/gen_effects_studentt.html) —
required `theta_G$nu`; the closed-form rescaler is applied internally.

**When to reach for it.** Heavy-tailed effects — a few sites with
substantively large effects relative to the bulk; robustness checks
of empirical-Bayes shrinkage estimators [@walters2024empirical]. The
package convention is $\nu = 5$ (variance and kurtosis both finite).

```{r studentt-call}
set.seed(1L)
out_studentt <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "StudentT",
  theta_G   = list(nu = 5)
)
summary(out_studentt$z_j)
var(out_studentt$z_j)
```

Asking for `theta_G = list()` (no `nu`) returns an actionable error
that names the missing field and the canonical fix:

```{r studentt-missing, error = TRUE}
set.seed(1L)
gen_effects(J = 200L, tau = 0, sigma_tau = 1,
            true_dist = "StudentT", theta_G = list())
```

### 3.3 Skew-Normal

**Parameterization.** $\boldsymbol\theta_G = \{\alpha\}$ with
$\alpha \in \mathbb{R}$ — the slant in the Azzalini parameterization
[@azzalini1985class]. Positive $\alpha$ gives right skew, negative
$\alpha$ gives left skew, and $\alpha = 0$ collapses to the Gaussian.

**Density.** $Y \sim \mathrm{SN}(0, 1, \alpha)$ with density

$$
f_Y(y; \alpha) \;=\; 2\, \phi(y)\, \Phi(\alpha y),
$$

then re-centered and rescaled so the residual has mean 0 and
variance 1.

**Unit-variance derivation.** Define
$\delta(\alpha) = \alpha / \sqrt{1 + \alpha^2}$. Then

$$
E[Y] = \delta \sqrt{2/\pi}, \qquad
\mathrm{Var}(Y) = 1 - 2\delta^2 / \pi,
$$

and the standardized residual is

$$
z_j \;=\; \frac{Y - \delta \sqrt{2/\pi}}{\sqrt{1 - 2\delta^2 / \pi}},
\qquad Y \sim \mathrm{SN}(0, 1, \alpha).
$$

By construction $E[z_j] = 0$ and $\mathrm{Var}(z_j) = 1$.

**Identifiability.** $\alpha = 0$ is the Gaussian limit (not refused,
but `true_dist = "Gaussian"` is the cleaner choice). Very large
$|\alpha| \gtrsim 30$ approaches the half-normal — possible but rare
in applied trial settings.

**Generator.**
[`gen_effects_skewn()`](../reference/gen_effects_skewn.html) —
required `theta_G$slant`. Soft dependency on the `sn` package.

**When to reach for it.** When the substantive story is "most
sites have small program effects, but a tail of sites have larger
effects." A program with one-sided heterogeneity — say, larger
effects in sites with stronger implementation — is naturally modeled
with $\alpha > 0$.

```{r skewn-call}
set.seed(1L)
out_skewn <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "SkewN",
  theta_G   = list(slant = 3)
)
summary(out_skewn$z_j)
var(out_skewn$z_j)
```

### 3.4 Asymmetric Laplace (ALD)

**Parameterization.** $\boldsymbol\theta_G = \{\rho\}$ with
$\rho \in (0, 1)$ — the Yu-Zhang quantile parameter
[@yu2005three]. $\rho = 0.5$ recovers the symmetric Laplace.

**Density.** Standard ALD with quantile $\rho$:

$$
f_X(x; \rho) \;=\; \rho(1 - \rho) \, \exp\bigl(-\rho_\rho(x)\bigr),
\qquad
\rho_\rho(x) \;=\; x \bigl(\rho - \mathbb{1}\{x < 0\}\bigr).
$$

**Unit-variance derivation.** Closed-form moments are

$$
E[X] = \frac{1 - 2\rho}{\rho(1 - \rho)}, \qquad
\mathrm{Var}(X) = \frac{1 - 2\rho + 2\rho^2}{\rho^2 (1 - \rho)^2},
$$

and the standardized residual is

$$
z_j \;=\; \frac{X - \mu_\rho}{\sigma_\rho},
\qquad
\mu_\rho \;=\; \frac{1 - 2\rho}{\rho(1 - \rho)},
\qquad
\sigma_\rho \;=\; \sqrt{\frac{1 - 2\rho + 2\rho^2}{\rho^2 (1 - \rho)^2}}.
$$

This is the same formula the blueprint specifies (Ch. 5 §4) and
implements in
[`gen_effects_ald()`](../reference/gen_effects_ald.html).

**Note on the empirical sanity table.** The realized
`var(z_j)` for ALD at `rho = 0.3` and $J = 200$ tends to land
appreciably below 1 (around 0.58 in the smoke check below). This is
a sampling artifact — at $J = 200$ the heavy-skewed Laplace's
variance estimator is high-variance, and a single seed sits in the
left tail of its sampling distribution. The closed-form derivation
above guarantees $\mathrm{Var}(z_j) = 1$ in the population. Section 4
returns to this with the cross-shape audit.

**Identifiability.** $\rho \to 0$ or $\rho \to 1$ degenerates: the
mean and variance both diverge. The exact admissible range is
documented at the function reference; in practice any $\rho \in
(0.05, 0.95)$ is comfortable.

**Generator.**
[`gen_effects_ald()`](../reference/gen_effects_ald.html) — required
`theta_G$rho`. Soft dependency on the `LaplacesDemon` package.

**When to reach for it.** Skewed effects with sharper peaks than the
skew-normal — a Laplacian core plus an asymmetric tail. The quantile
parameterization is also useful when an expert can name the modal
quantile of the distribution ("the typical site sits at the 30th
percentile of program effects").

```{r ald-call}
set.seed(1L)
out_ald <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "ALD",
  theta_G   = list(rho = 0.3)
)
summary(out_ald$z_j)
var(out_ald$z_j)
```

### 3.5 Two-component Gaussian mixture

**Parameterization.** Three named fields,
$\boldsymbol\theta_G = \{\Delta, \varepsilon, \mathrm{ups}\}$, where:

- `delta` ($\Delta > 0$) — the separation between the two component
  means;
- `eps` ($\varepsilon \in (0, 1)$) — the mixing weight on the second
  component;
- `ups` ($\mathrm{ups} > 0$) — the SD ratio
  $\sigma_2 / \sigma_1$ (the JEBS-appendix convention; see
  [M8: Migration from siteBayes2](m8-migration-from-siteBayes2.html)).

**Density.** Two Gaussian components, weighted to integrate to 1
with location chosen so the unconditional mean is 0:

$$
z \;\sim\; \begin{cases}
\mathcal{N}\bigl(-\varepsilon \Delta, \; \sigma_1^2\bigr) & \text{w.p. } 1 - \varepsilon, \\
\mathcal{N}\bigl((1 - \varepsilon)\Delta, \; \sigma_2^2\bigr) & \text{w.p. } \varepsilon,
\end{cases}
\qquad \sigma_1 = 1, \;\; \sigma_2 = \mathrm{ups}.
$$

**Unit-variance derivation.** By the law of total variance, the
unconditional variance of the mixture is

$$
\mathrm{Var}(z) \;=\; (1 - \varepsilon) + \varepsilon \cdot \mathrm{ups}^2
\;+\; \varepsilon (1 - \varepsilon) \, \Delta^2,
$$

and the standardized residual is $z / \sqrt{\mathrm{Var}(z)}$. The
generator divides by the closed-form $\sqrt{\mathrm{Var}(z)}$
internally; the user does not handle the rescale.

**Identifiability.** The two components are exchangeable in
principle, so the package fixes a labeling convention (ascending
mean) and threads a `latent_component` integer column through the
output for downstream label-switching audits. There is no automatic
refuse region, but extreme combinations
($\varepsilon$ near 0 or 1, $\Delta$ near 0) reduce to a single
Gaussian and could just as well use `true_dist = "Gaussian"`.

**Generator.**
[`gen_effects_mixture()`](../reference/gen_effects_mixture.html) —
required `theta_G$delta`, `theta_G$eps`, `theta_G$ups`.

**When to reach for it.** Bimodality (two latent classes of sites);
contamination (a small mixing weight on a wide component, modeling
outlier sites); robustness checks of estimators against
non-unimodal $G$.

```{r mixture-call}
set.seed(1L)
out_mixture <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "Mixture",
  theta_G   = list(delta = 5, eps = 0.3, ups = 2)
)
summary(out_mixture$z_j)
var(out_mixture$z_j)
```

The realized variance is close to 1 — the closed-form variance was
divided out before the Layer-1 output was returned.

### 3.6 Point-mass-plus-slab

**Parameterization.** Required `theta_G$pi0` (numeric in $(0, 1)$),
plus optional slab parameters. The slab defaults are
`slab_shape = "Gaussian"`, `mu_slab = 0`, `sigma_slab = 1`. The
point-mass weight is `pi0`; with probability `1 - pi0` the residual
is drawn from the slab.

**Density.** A spike-and-slab mixed measure:

$$
z \;\sim\; \begin{cases}
0 & \text{w.p. } \pi_0, \\
s, \;\; s \sim f_{\text{slab}}(0, \sigma_{\text{slab}}^2) & \text{w.p. } 1 - \pi_0.
\end{cases}
$$

**Unit-variance derivation.** The point mass contributes zero to the
variance. Setting $\sigma_{\text{slab}}^2 = 1 / (1 - \pi_0)$
internally gives

$$
\mathrm{Var}(z) \;=\; (1 - \pi_0) \cdot \frac{1}{1 - \pi_0} \;=\; 1.
$$

**Identifiability.** $\pi_0 \in \{0, 1\}$ is the boundary: at $\pi_0
= 0$ the residual is the slab alone (use the slab family directly);
at $\pi_0 = 1$ the residual is identically zero ($\sigma_\tau$ would
be the only source of cross-site variance, and the package refuses
that combination).

**Generator.**
[`gen_effects_pmslab()`](../reference/gen_effects_pmslab.html) —
required `theta_G$pi0`. Slab parameters are optional; defaults are
documented at the function reference.

**When to reach for it.** Sparse program effects — a fraction $\pi_0$
of sites are exactly null, and the remaining $1 - \pi_0$ have
non-zero effects with the slab's shape. Walters-2024-style
empirical-Bayes scenarios with effective null sites
[@walters2024empirical] are the canonical case; the spike-and-slab
literature in high-dimensional regression provides additional
intuition.

```{r pmslab-call}
set.seed(1L)
out_pmslab <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "PointMassSlab",
  theta_G   = list(pi0 = 0.3)
)
summary(out_pmslab$z_j)
var(out_pmslab$z_j)
mean(out_pmslab$z_j == 0)
```

The realized fraction of exact zeros at $\pi_0 = 0.3$ tracks the
target — this is the spike portion of the spike-and-slab.

### 3.7 User callback

**Parameterization.** $\boldsymbol\theta_G$ is whatever your callback
consumes. The required argument is `g_fn`: a function of `J`
returning a length-`J` numeric vector.

**Density.** Whatever your callback returns. The package does not
introspect the callback's distribution; it only audits the
mean-and-variance contract on a small prerun (Convention A).

**Unit-variance derivation.** Two conventions are supported
(`g_returns` argument, default `"standardized"`):

- **Convention A (standardized residuals — recommended)**:
  `g_fn(J, ...)` returns standardized residuals with mean 0 and
  variance 1. The Layer-1 wrapper applies $\tau + \sigma_\tau \cdot
  z_j$ as for the seven built-ins. With `audit_g = TRUE` (default),
  the package draws a $10^5$-sample prerun and warns if
  $|\hat\mu| > 0.1$ or $|\hat\sigma^2 - 1| > 0.1$.
- **Convention B (raw effects — wrapper bypass)**:
  `g_fn(J, ...)` returns final $\tau_j$ values directly. The
  location-scale wrapper is skipped. Use only when you need explicit
  control over the marginal mean and variance.

**Identifiability.** Under Convention A, the standardization
contract is the user's responsibility — the audit warns but does not
override. Under Convention B, the user owns the entire marginal.

**Generator.**
[`gen_effects_user()`](../reference/gen_effects_user.html) —
required `g_fn`. Convention selection is via `g_returns`.

**When to reach for it.** None of the seven built-in shapes match —
horseshoe-prior-style residuals, a finite mixture with more than two
components, a calibrated empirical distribution from a sister study,
or any custom $G$ that needs to be threaded through the rest of the
pipeline. Convention A is the package's recommended path; Convention
B exists for the edge case where the user truly wants to bypass the
location-scale wrapper.

```{r user-call}
set.seed(1L)
g_fn_laplace_std <- function(J, ...) {
  x <- rexp(J) - rexp(J)              # symmetric Laplace, var = 2
  x / sd(x)                            # standardize empirically
}
out_user <- gen_effects(
  J         = 200L,
  tau       = 0,
  sigma_tau = 1,
  true_dist = "User",
  g_fn      = g_fn_laplace_std
)
summary(out_user$z_j)
var(out_user$z_j)
```

The example callback returns a Laplace-shaped standardized residual
generator; the realized variance lands at 1 because the callback
did the standardization itself.

### 3.8 DPM stub

**Parameterization.** Currently unimplemented; calling
`true_dist = "DPM"` without a `g_fn` bridge raises an actionable
error.

**Density.** A Dirichlet-process mixture would be a nonparametric $G$
with a base measure and a concentration parameter; the package
accepts the string `"DPM"` for forward-compatibility with a future
release, but the built-in sampler is not in scope for the current
release.

**Unit-variance derivation.** Not applicable.

**Generator.**
[`gen_effects_dpm()`](../reference/gen_effects_dpm.html) raises an
error. Use the `User` bridge below.

**When to reach for it.** When you actually need DPM-distributed
latent effects, the bridge pattern is to externally fit a DPM (with
the `dirichletprocess` package or equivalent), draw a large sample
of standardized residuals, and wrap them in a Convention A callback:

```r
# 1. External draw — fit DPM, extract posterior mean / posterior draws
dp <- dirichletprocess::DirichletProcessGaussian(rnorm(200))
dp <- dirichletprocess::Fit(dp, its = 500, progressBar = FALSE)
post_z <- dirichletprocess::PosteriorFunction(dp)(seq(-5, 5, length.out = 1e5))

# 2. Standardize so Convention A holds
post_z <- (post_z - mean(post_z)) / sd(post_z)

# 3. Wrap as a Convention A callback
dpm_bridge_g <- function(J, ...) sample(post_z, size = J, replace = TRUE)

# 4. Use as a User-shape generator
gen_effects(J = 30L, tau = 0.3, sigma_tau = 0.1,
            true_dist = "User", g_fn = dpm_bridge_g)
```

Calling DPM without the bridge:

```{r dpm-error, error = TRUE}
set.seed(1L)
gen_effects(J = 200L, tau = 0, sigma_tau = 1, true_dist = "DPM")
```

The error message names the canonical fix verbatim: use
`true_dist = "User"` with a custom `g_fn`, or pass a bridge
explicitly. The package never silently falls back to a Gaussian.

## 4. Unit-variance sanity check

The variance contract is $\mathrm{Var}(z_j) = 1$ in the population.
At $J = 200$ the sample variance fluctuates around 1 — the table
below collects the realized values from each shape's primary call
above. This is a reproducibility check, not a hypothesis test:
the closed-form derivations in Section 3 guarantee the population
contract, and the empirical departures here are within the sampling
window expected at $J = 200$.

```{r sanity-table}
sanity <- tibble::tribble(
  ~shape,                 ~theta_G_summary,                  ~var_z_j,
  "Gaussian",             "(none)",                          var(out_gaussian$z_j),
  "StudentT",             "nu = 5",                          var(out_studentt$z_j),
  "SkewN",                "slant = 3",                       var(out_skewn$z_j),
  "ALD",                  "rho = 0.3",                       var(out_ald$z_j),
  "Mixture",              "delta = 5, eps = 0.3, ups = 2",   var(out_mixture$z_j),
  "PointMassSlab",        "pi0 = 0.3",                       var(out_pmslab$z_j),
  "User (Laplace-style)", "g_fn pre-standardized",           var(out_user$z_j),
  "DPM",                  "(error — bridge to User)",        NA_real_
)
sanity
```

Three things to read off the table. First, the user-supplied
generator hit $\mathrm{Var}(z_j) = 1$ exactly — the callback's
empirical standardization step (`x / sd(x)`) forces equality on the
draw itself, not on the population. Second, Gaussian, Student-t,
skew-normal, mixture, and point-mass-plus-slab cluster within roughly
0.05 of 1, which is the expected sampling window at $J = 200$ for a
finite-variance shape. Third, the ALD's realized variance is below
1 by more than the other shapes; the closed-form derivation
guarantees population variance 1, and the sample deficit is the
realized value of an estimator whose own sampling variance is large
when the shape is heavy-skewed. A larger $J$ (e.g. $J = 10^6$, the
contract level used in the package test suite) collapses the gap to
$10^{-3}$ for every shape.

The DPM row is `NA_real_` by construction — the built-in sampler
errors out, so the bridge to `User` is the only path. Section 3.7
shows the wrapper.

## 5. Decision rubric

When more than one shape is plausible for an applied setting, the
following rubric maps the dominant feature of the substantive story
to the recommended `true_dist`. The list is short on purpose: most
applied trials at $J \le 50$ cannot empirically discriminate between
adjacent shapes, so the rule of thumb is "pick the shape whose
parameter has a substantive interpretation in your setting and run
the rest of the diagnostic pipeline against it."

| Substantive feature of the cross-site story | Recommended `true_dist` | `theta_G` start |
|---------------------------------------------|-------------------------|-----------------|
| No prior expectation; first-pass simulation | `"Gaussian"` | (none) |
| A few sites with extreme effects (heavy tails, symmetric) | `"StudentT"` | `nu = 5` |
| One-sided skew (right or left tail of effects) | `"SkewN"` | `slant = ±3` |
| Skew with sharper peak; quantile interpretation natural | `"ALD"` | `rho = 0.3` (right skew) or `rho = 0.7` (left skew) |
| Two latent classes of sites; bimodality | `"Mixture"` | `delta = 5, eps = 0.3, ups = 2` |
| A fraction of sites are exactly null | `"PointMassSlab"` | `pi0 = 0.3` |
| Custom or empirical-distribution residuals | `"User"` | depends on callback |

The Student-t-vs-skew-normal-vs-ALD axis is the one that looks
similar across families but means different things substantively.
Heavy-tailed-but-symmetric (a few outlier sites, but no left/right
asymmetry) calls for Student-t. Skewed-but-light-tailed (a tilt
toward larger or smaller effects, with a finite kurtosis) calls for
skew-normal. Skewed-with-sharper-peak (the bulk near the mode plus
an asymmetric exponential tail) calls for ALD; the Yu-Zhang
quantile $\rho$ is also the only parameter in the catalog with a
direct quantile interpretation, which makes it the easiest of the
three to elicit from a domain expert.

Cross-references for downstream interpretation:

- [A2: Choosing a preset](a2-choosing-a-preset.html) — preset-level
  comparison of how shape choice flows through the diagnostic block.
- [A3: Diagnostics in practice](a3-diagnostics-in-practice.html) — the
  diagnostic-grid against which a shape choice is audited.
- [M5: Custom G distributions](m5-custom-g-distributions.html) — the
  long-form treatment of the User pathway, including Convention A
  vs B and the audit machinery.

### Known limitation — `scenario_audit()` Group C resolver

The Group C reference resolver inside
[`scenario_audit()`](../reference/scenario_audit.html) currently
supports automatic Gaussian and Student-$t$ targets only. Designs
declaring a SkewN, ALD, Mixture, or PointMassSlab latent fall back
to `target_source = "not_available"` for the Group C row of the
audit (the row prints in the diagnostic block but the audit pipeline
does not gate on it). The Group A, B, and D rows of the audit
operate normally for every shape. Manual Group C checks are still
available — call
[`bhattacharyya_coef()`](../reference/bhattacharyya_coef.html) or
[`ks_distance()`](../reference/ks_distance.html) with an explicit
target sample drawn from the declared shape. Resolver extension to
the remaining catalog shapes is on the v0.2.0 roadmap.

## 6. Plots — density and QQ panels per shape

One plot per shape (seven total — the DPM stub has no density
panel). Each panel overlays the realized standardized residuals
on a reference Gaussian QQ-line so the reader can read the
shape-specific tail behavior off the figure. The bandwidth is the
ggplot2 default; the QQ-line is `y = x` because both axes are
standardized.

```{r plot-helper, include = FALSE}
plot_density_qq <- function(z, shape_label, theta_label) {
  df <- tibble::tibble(z = z)
  p_density <- ggplot(df, aes(x = z)) +
    geom_density(fill = "#1B4965", alpha = 0.35, colour = "#1B4965") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      x        = expression(z[j]),
      y        = "density",
      title    = paste0(shape_label, " — density of z_j"),
      subtitle = theta_label
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
  qq_df <- tibble::tibble(
    sample = sort(z),
    theory = qnorm(ppoints(length(z)))
  )
  p_qq <- ggplot(qq_df, aes(x = theory, y = sample)) +
    geom_point(alpha = 0.5, size = 1.4, colour = "#62929E") +
    geom_abline(slope = 1, intercept = 0, colour = "grey40") +
    labs(
      x = "Gaussian theoretical quantile",
      y = expression(z[j]~"  sample quantile"),
      title = "QQ vs Gaussian"
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
  gridExtra::grid.arrange(p_density, p_qq, ncol = 2)
}
```

### 6.1 Gaussian

```{r plot-gaussian, fig.height = 3.6, fig.cap = "Gaussian density (left) and QQ-against-Gaussian (right). The QQ points sit on the y = x reference line, confirming the realized residuals are Gaussian within sampling fluctuation at J = 200.", fig.alt = "Two-panel figure: a roughly bell-shaped density and a QQ plot whose points trace the y = x reference line."}
plot_density_qq(out_gaussian$z_j, "Gaussian", "theta_G = (none)")
```

### 6.2 Student-t

```{r plot-studentt, fig.height = 3.6, fig.cap = "Student-t (nu = 5) density and QQ. Tails of the QQ plot bow away from the y = x line at both ends — the heavy-tail signature relative to Gaussian, the canonical reason to reach for this shape.", fig.alt = "Two-panel figure: a peaked density with heavier tails than Gaussian, and a QQ plot whose tails bow away from the reference line."}
plot_density_qq(out_studentt$z_j, "Student-t", "theta_G = list(nu = 5)")
```

### 6.3 Skew-Normal

```{r plot-skewn, fig.height = 3.6, fig.cap = "Skew-normal (slant = 3) density and QQ. The density is right-skewed; the upper tail of the QQ plot sits above y = x while the lower tail sits on or above it — the asymmetric signature.", fig.alt = "Two-panel figure: a right-skewed density with a longer right tail, and a QQ plot whose upper points lie above the reference line."}
plot_density_qq(out_skewn$z_j, "Skew-normal", "theta_G = list(slant = 3)")
```

### 6.4 ALD

```{r plot-ald, fig.height = 3.6, fig.cap = "ALD (rho = 0.3) density and QQ. The density has a sharper peak than skew-normal and an exponential tail — the QQ plot shows departures at both tails relative to Gaussian, with a noticeable asymmetry.", fig.alt = "Two-panel figure: a sharper-peaked density with one heavier tail, and a QQ plot whose tails diverge asymmetrically from the reference line."}
plot_density_qq(out_ald$z_j, "ALD", "theta_G = list(rho = 0.3)")
```

### 6.5 Mixture

```{r plot-mixture, fig.height = 3.6, fig.cap = "Two-component mixture (delta = 5, eps = 0.3, ups = 2) density and QQ. The density is bimodal — two peaks separated by delta and weighted by eps. The QQ plot has the characteristic step-shape of a bimodal distribution against a unimodal reference.", fig.alt = "Two-panel figure: a bimodal density with two visible peaks, and a QQ plot with a characteristic step or s-shape."}
plot_density_qq(out_mixture$z_j, "Mixture", "delta = 5, eps = 0.3, ups = 2")
```

### 6.6 Point-mass-plus-slab

```{r plot-pmslab, fig.height = 3.6, fig.cap = "Point-mass-plus-slab (pi0 = 0.3) density and QQ. The density has a visible spike at z = 0 — the point-mass weight — plus a Gaussian-shaped slab. The QQ plot has a flat plateau through the middle quantiles where the spike sits.", fig.alt = "Two-panel figure: a density with a sharp central spike at zero plus a wider slab, and a QQ plot with a flat plateau at zero in the middle quantiles."}
plot_density_qq(out_pmslab$z_j, "PointMassSlab", "theta_G = list(pi0 = 0.3)")
```

### 6.7 User

```{r plot-user, fig.height = 3.6, fig.cap = "User callback (Laplace-style standardized residuals). The density has a sharper peak than Gaussian and slightly heavier tails — the Laplace signature. The QQ plot bows at both ends similarly to a symmetric heavy-tailed distribution.", fig.alt = "Two-panel figure: a peaked density with a sharper apex than Gaussian, and a QQ plot with symmetric tail bowing."}
plot_density_qq(out_user$z_j, "User (Laplace-shape)", "g_fn = symmetric Laplace, pre-standardized")
```

### 6.8 Comparison overlay

A single overlay of all seven density curves on a common axis. Each
shape's signature feature reads off the figure in one glance — the
heavy tails of Student-t and User-Laplace, the right skew of
SkewN and ALD, the bimodality of Mixture, and the central spike of
PointMassSlab.

```{r plot-overlay, fig.height = 4.6, fig.cap = "Density overlay of seven shapes at J = 200. Read off: Gaussian is the symmetric reference; Student-t has heavier tails; SkewN has a longer right tail; ALD is sharper-peaked and right-skewed; Mixture is bimodal; PointMassSlab has a central spike at z = 0; User-Laplace is sharper-peaked than Gaussian. The DPM stub is not plotted.", fig.alt = "Composite density overlay of seven distinct standardized-residual shapes on a common z-axis, each in a different color, illustrating the catalog's visual signature differences."}
df_overlay <- dplyr::bind_rows(
  tibble::tibble(z = out_gaussian$z_j, shape = "Gaussian"),
  tibble::tibble(z = out_studentt$z_j, shape = "Student-t (nu = 5)"),
  tibble::tibble(z = out_skewn$z_j,    shape = "SkewN (slant = 3)"),
  tibble::tibble(z = out_ald$z_j,      shape = "ALD (rho = 0.3)"),
  tibble::tibble(z = out_mixture$z_j,  shape = "Mixture (5, 0.3, 2)"),
  tibble::tibble(z = out_pmslab$z_j,   shape = "PMSlab (pi0 = 0.3)"),
  tibble::tibble(z = out_user$z_j,     shape = "User (Laplace)")
)
ggplot(df_overlay, aes(x = z, colour = shape)) +
  geom_density(linewidth = 0.6, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x        = expression(z[j]~"  (standardized residual)"),
    y        = "density",
    title    = "Catalog overlay — seven standardized G shapes at J = 200",
    subtitle = "All shapes are unit-variance by construction; tail and skew differences are the only design lever.",
    colour   = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom")
```

## 7. Where to next

- [M1: The two-stage DGP](m1-statistical-dgp.html) — the formal
  specification of the four generative layers; this vignette
  operationalizes Layer 1's $G$-shape choice.
- [M3: Margin and SE models](m3-margin-se-models.html) — the
  Stage-2 precision marginal; once $z_j$ is fixed, the
  cross-site spread of $\widehat{se}_j^{\,2}$ is the next design
  choice.
- [M5: Custom G distributions](m5-custom-g-distributions.html) —
  the long-form treatment of the User callback path, including the
  audit machinery for Convention A and the recipe for bridging
  external-distribution samples into the package.
- [A2: Choosing a preset](a2-choosing-a-preset.html) — the applied
  view of how shape choice cascades into the diagnostic readout.
- [A8: Cookbook](a8-cookbook.html) — short walkthroughs that mix
  shape choice with the rest of the design surface.

## References

The factorization $\tau_j = \tau + \sigma_\tau z_j$ with
unit-variance standardized residual is the convention adopted in
@lee2025improving; the empirical-Bayes context is collected in
@walters2024empirical. The skew-normal parameterization is from
@azzalini1985class, and the asymmetric Laplace parameterization is
from @yu2005three. The simulation-design defaults underlying the
catalog match the JEBS-paper preset
[@lee2025improving] and the
applied calibration benchmarks [@weiss2017howmuch].

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
