---
title: "M3 · Margin and SE models — site-size and direct-precision paths"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists choosing between a multisite-trial design (where
  per-site sample sizes determine sampling precision) and a
  meta-analytic design (where reported standard errors are the design
  surface), this vignette derives the two precision margins
  `multisiteDGP` exposes, contrasts the two front doors `sim_multisite()`
  and `sim_meta()`, demonstrates that the package enforces
  paradigm-strict dispatch by erroring on cross-door arguments, and
  shows the realized SE distribution under each path. You leave with
  the mapping rule for picking a path, two diagnostic plots
  (kappa-over-n overlay and (I, R)-implied target overlay), and the
  identifiability statement that lets the two paths be reasoned about
  independently.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M3 Margin and SE models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why two paradigms

Multisite trials and meta-analyses describe the same downstream
quantity — a per-unit estimate $\hat\tau_j$ paired with a reported
sampling standard error $\widehat{se}_j$ — but they hand the analyst a
different design surface. A multisite-trial design arrives
as a list of per-site sample sizes $n_j$: the trialist negotiated a
target enrolment with each site, the protocol fixed an allocation
share $p$ between treatment and control, and the per-site sampling
variance follows from $n_j$ via the closed-form
$\widehat{se}_j^{\,2} = \kappa / n_j$ derived in
[M1: The two-stage DGP](m1-statistical-dgp.html). A meta-analytic
design, by contrast, arrives as a list of reported
$(\hat\tau_j, \widehat{se}_j)$ pairs: studies in the corpus reported
their effect sizes and standard errors directly, and the underlying
per-study sample sizes are heterogeneous, sometimes unrecoverable, and
in any case not the natural design knob a methodologist turns when
asking "what happens to my estimator when the precision spread widens?"

`multisiteDGP` exposes both surfaces as separate front doors. The
*site-size-driven* path (Paradigm A in the blueprint) takes per-site
sample sizes — distributed via a truncated Gamma — and derives
sampling variances. The *direct-precision* path (Paradigm B in the
blueprint) takes informativeness and heterogeneity targets directly
and constructs the SE marginal to match. Both paths route the same
Layer 1 latent-effect generator, the same Layer 3 precision-dependence
machinery, and the same Layer 4 observation draw; the only difference
is Layer 2 — the SE-margin generator. Keeping the two paths separated
at the front door — rather than rolling them into a single dispatcher
— is what makes simulation provenance auditable: a reader can tell at
a glance which design surface the simulation lives on, and the
package errors on cross-door arguments before any compute happens.

## 2. Paradigm A — site-size-driven

### 2.1 The formal model

Under the *site-size-driven* path, the per-site sampling variance is
the deterministic ratio of a configuration-derived constant $\kappa$
and the per-site sample size $n_j$:

$$
\widehat{se}_j^{\,2} \;=\; \frac{\kappa}{n_j},
\qquad
\kappa \;=\; \left(\frac{1}{p} + \frac{1}{1 - p}\right)(1 - R^2)\, s_Y^2.
$$

The closed form for $\kappa$ comes from the
balanced-randomization Neyman variance: $p$ is the treatment-allocation
share, $R^2$ is the share of outcome variance explained by site-level
covariates (zero in the absence of covariates), and $s_Y^2$ is the
residual outcome variance on the response scale. At the package
defaults ($p = 0.5$, $R^2 = 0$, $s_Y^2 = 1$), $\kappa = 4$ exactly. The
per-site sample sizes are drawn from a truncated Gamma marginal with
user-supplied conditional mean $\bar n_j$, conditional CV
$\mathrm{CV}$, and lower truncation $n_{j,\min}$:

$$
n_j \;\stackrel{\text{i.i.d.}}{\sim}\; \mathrm{TruncGamma}\bigl(\bar n_j,\, \mathrm{CV};\, n_{j,\min}\bigr).
$$

A two-line check confirms $\kappa$ at the defaults and the closed form
under non-zero $R^2$:

```{r kappa-defaults}
compute_kappa(p = 0.5, R2 = 0,   var_outcome = 1)
compute_kappa(p = 0.5, R2 = 0.4, var_outcome = 1)
```

The first call returns 4 (the JEBS default); the second returns
$4 \cdot (1 - 0.4) = 2.4$ — covariate adjustment shrinks $\kappa$
multiplicatively. Section 5 returns to this with a kappa-rate overlay
on a real simulation.

### 2.2 Two engines for the site-size margin

The truncated-Gamma marginal is implemented by two engines that hit
two different invariants. Engine A1 (`A1_legacy`,
`legacy_jebs_censor_round`) reproduces the JEBS paper's published
numerical results bit-for-bit: the engine matches the JEBS appendix's
exact RNG call sequence — `rgamma(J)` for site sizes, lower-censoring
at $n_{j,\min}$ via `pmax`, integer rounding, then a row-shuffle inside
`dplyr::slice_sample(prop = 1)` — and the canonical hash of its output
under the published preset matches @lee2025improving's appendix
output byte-for-byte. Engine A2 (`A2_modern`, `trunc_gamma_moment`)
solves the truncated-Gamma moment equations numerically so that the
*conditional* mean and CV given $n_j \ge n_{j,\min}$ exactly match the
user-supplied $\bar n_j$ and $\mathrm{CV}$. Engine A1's
censor-and-round procedure shifts those conditional moments away from
the user inputs; Engine A2 keeps them on target. Both engines deliver
the same closed-form $\widehat{se}_j^{\,2} = \kappa/n_j$ identity at
the *standardized residual* scale — the package's testable invariant
T1b — and both feed the same downstream Layer 3 / Layer 4 machinery.

The choice is mechanical:

- Pick `engine = "A1_legacy"` when the explicit goal is to reproduce
  @lee2025improving's published numerical results bit-for-bit (testable
  invariant T1a). The presets [`preset_jebs_paper()`](../reference/preset_jebs_paper.html)
  and [`preset_jebs_strict()`](../reference/preset_jebs_strict.html) wire
  this engine in by default.
- Pick `engine = "A2_modern"` (the package default) for any new
  simulation. The conditional-moment match makes the user's
  $(\bar n_j, \mathrm{CV})$ inputs faithful to the realized site-size
  distribution, which is the contract a methodologist running power
  studies, calibration sweeps, or sensitivity analyses needs.

### 2.3 The front door

The unified entry point for the *site-size-driven* path is
[`sim_multisite()`](../reference/sim_multisite.html). Either pass a
fully-constructed `multisitedgp_design` with `paradigm = "site_size"`,
or hand the wrapper flat arguments and the package builds the design
inline:

```{r paradigm-A-flat}
dat_A <- sim_multisite(preset_education_modest(), seed = 1L)
dat_A
```

The print method's diagnostic block reports the realized
informativeness `I` near 0.30, the realized heterogeneity ratio `R`
above 10, and the kappa-rate identity $\widehat{se}_j^{\,2} \cdot n_j =
\kappa = 4$ holding by construction:

```{r paradigm-A-kappa-rate}
range(dat_A$se2_j * dat_A$n_j)
unique(round(dat_A$se2_j * dat_A$n_j, 8))
```

The `range()` check returns `(4, 4)` and the `unique()` check returns
the single value 4 — the kappa-rate identity holds for every site, the
testable invariant T3 in [M1 §7](m1-statistical-dgp.html). The
realized informativeness reads back via either the diagnostics
attribute or the standalone helper:

```{r paradigm-A-I}
attr(dat_A, "diagnostics")$I_hat
compute_I(dat_A$se2_j, sigma_tau = 0.20)
```

Both calls return the same value (0.303) — the diagnostic block stores
the result of a [`compute_I()`](../reference/compute_I.html) call on
the realized sampling variances, and the standalone helper recomputes
it from the same vector. The realized $I$ falls below the design
intent slightly because the kappa-over-$n_j$ marginal carries Jensen
gap from the $n_j$ heterogeneity (CV = 0.5).

### 2.4 When to pick this path

The *site-size-driven* path fits the design surface when (a) the study
specifies per-site sample sizes — so $\bar n_j$, $\mathrm{CV}$, and
$n_{j,\min}$ are quantities you can defend on substantive grounds —
and (b) the precision heterogeneity that emerges from the site-size
distribution is itself part of the simulation question. Power
calculations for actual multisite trials, calibration sweeps against
realized sample-size distributions in the
@weiss2017howmuch education-trial benchmarks, and any
JEBS-paper reproduction
[@lee2025improving] all fit here. The path's stochasticity — the
realized $(I, R)$ are *consequences* of the $(\bar n_j, \mathrm{CV})$
draws, not exact inputs — is the feature, not the bug: it makes the
simulation faithful to the trial specification rather than to a
post-hoc precision target.

The Layer 2 generator is exposed individually as
[`gen_site_sizes()`](../reference/gen_site_sizes.html) for users
composing the four layers manually.

## 3. Paradigm B — direct-precision

### 3.1 The formal model

Under the *direct-precision* path, the user supplies the two precision
diagnostics directly: the mean informativeness scalar $I$ and the
heterogeneity ratio $R$. The Layer 2 generator back-calculates a
deterministic SE marginal that hits both targets exactly. The
informativeness identity from
[@lee2025improving]'s Equation (4) implies that the geometric mean of
$\widehat{se}_j^{\,2}$ is determined by $\sigma_\tau$ and $I$:

$$
\mathrm{GM}\bigl(\widehat{se}_j^{\,2}\bigr) \;=\; \sigma_\tau^{\,2} \cdot \frac{1 - I}{I}.
$$

The heterogeneity ratio fixes the spread around the geometric mean.
The package's grid construction places the $J$ values log-evenly over
the interval $[\mathrm{GM} \cdot R^{-1/2}, \mathrm{GM} \cdot R^{1/2}]$:

$$
\widehat{se}_j^{\,2} \;=\; \mathrm{GM} \cdot R^{(j-1)/(J-1) - 1/2},
\qquad j = 1, \ldots, J.
$$

The construction is the Yu-Zhang-style log-even spacing
(@yu2005three's quantile parameterization in spirit, applied here to
the SE marginal rather than the latent-effect shape). The endpoint
arithmetic guarantees that the multiset's geometric mean is
$\mathrm{GM}$ exactly and its max-over-min ratio is $R$ exactly. The
construction is deterministic: no randomness enters the SE marginal
itself, only the optional `shuffle = TRUE` permutation of the grid
values across sites. The site-size column $n_j$ is `NA_integer_` —
this path discards the per-site sample size entirely.

### 3.2 The front door

The unified entry point for the *direct-precision* path is
[`sim_meta()`](../reference/sim_meta.html). The minimal call passes
flat $(J, I, R, \sigma_\tau)$ arguments and lets the wrapper build the
direct-paradigm design inline:

```{r paradigm-B-flat}
dat_B <- sim_meta(J = 50L, I = 0.30, R = 2, sigma_tau = 0.20, seed = 1L)
dat_B
```

The print method's diagnostic block confirms the (I, R) targets are
hit *exactly*: `I: target = 0.300, realized = 0.300, PASS` and
`R: target = 2.000, realized = 2.000, PASS`. The `n_j` column is `NA`
across all 50 rows — the *direct-precision* path has no site-size
margin to populate.

The exactness is verifiable via the diagnostics attribute, which
under this path adds `target_I`, `target_R`, `I_error`, `R_error`,
`I_exact`, and `R_exact` slots:

```{r paradigm-B-exact}
diag_B <- attr(dat_B, "diagnostics")
c(I_hat = diag_B$I_hat, target_I = diag_B$target_I,
  I_error = diag_B$I_error, I_exact = diag_B$I_exact)
c(R_hat = diag_B$R_hat, target_R = diag_B$target_R,
  R_error = diag_B$R_error, R_exact = diag_B$R_exact)
```

Both error slots are zero, both exactness flags are `TRUE`. The
underlying Layer 2 generator is exposed individually as
[`gen_se_direct()`](../reference/gen_se_direct.html) for manual
composition.

### 3.3 When to pick this path

The *direct-precision* path fits the design surface when (a) the study
specifies (or the methodologist controls) the precision diagnostics
themselves rather than a sample-size distribution — meta-analyses
where the corpus reports $(\hat\tau_j, \widehat{se}_j)$ without
recoverable $n_j$, Fay-Herriot small-area estimation problems, or
methods-comparison studies that need to fix informativeness exactly
across cells of a sweep — and (b) the methodologist wants the realized
$(I, R)$ to match the targets bit-for-bit, not stochastically. The
deterministic-grid construction is the feature: when comparing a
shrinkage estimator across $I \in \{0.10, 0.30, 0.50, 0.80\}$, the
*direct-precision* path lets the methodologist hold the precision-side
design constant and isolate estimator behavior, free from the
sample-size noise the *site-size-driven* path introduces by design.

Two presets wire the *direct-precision* path in by default:
[`preset_meta_modest()`](../reference/preset_meta_modest.html) for
modest meta-analytic informativeness and
[`preset_small_area_estimation()`](../reference/preset_small_area_estimation.html)
for the wider precision spread typical of small-area surveys. The
canonical applied walkthrough is
[A7: Case study — meta-analysis](a7-case-study-meta-analysis.html).

## 4. The two paradigms as a contract table

The four-column contract below is the operational summary: which
formal model each path fixes at Layer 2, when to reach for it, and
which front door to call. Cross-door arguments are rejected by both
wrappers (Section 6 demonstrates).

| Paradigm | Formal model (Layer 2 SE margin) | When to pick | Front door / preset family |
|----|----|----|----|
| Site-size-driven (Paradigm A) | $\widehat{se}_j^{\,2} = \kappa / n_j$ with $\kappa = (1/p + 1/(1-p))(1 - R^2) s_Y^2$ and $n_j \sim \mathrm{TruncGamma}(\bar n_j, \mathrm{CV}; n_{j,\min})$ | Per-site sample sizes are part of the design surface; precision heterogeneity is a *consequence* of the site-size distribution. | [`sim_multisite()`](../reference/sim_multisite.html); [`preset_jebs_paper()`](../reference/preset_jebs_paper.html), [`preset_jebs_strict()`](../reference/preset_jebs_strict.html), [`preset_education_modest()`](../reference/preset_education_modest.html), [`preset_education_substantial()`](../reference/preset_education_substantial.html), [`preset_education_small()`](../reference/preset_education_small.html), [`preset_walters_2024()`](../reference/preset_walters_2024.html), [`preset_twin_towers()`](../reference/preset_twin_towers.html) |
| Direct-precision (Paradigm B) | $\widehat{se}_j^{\,2} = \mathrm{GM} \cdot R^{(j-1)/(J-1) - 1/2}$ with $\mathrm{GM} = \sigma_\tau^{\,2}(1 - I)/I$, $n_j = \mathtt{NA\_integer\_}$ | Precision diagnostics $(I, R)$ are part of the design surface; per-site sample sizes are absent or incidental. | [`sim_meta()`](../reference/sim_meta.html); [`preset_meta_modest()`](../reference/preset_meta_modest.html), [`preset_small_area_estimation()`](../reference/preset_small_area_estimation.html) |

The two paths are *isomorphic at the level of the realized
$\widehat{se}_j^{\,2}$ vector* — once Layer 2 has populated `se2_j`,
Layer 3 and Layer 4 see the same input and run identically. This is
what makes the contrast a front-door distinction rather than a
downstream one: the design surface differs, but the simulation
pipeline downstream of Layer 2 is shared.

## 5. The realized SE distribution under each paradigm

The two plots below are diagnostic, not decorative: each one shows
the realized SE marginal under one path and overlays the formal-model
target the path was supposed to hit. Read each plot for whether the
realized distribution lands on the target overlay.

### 5.1 Plot 1 — Paradigm A realized SE distribution

Under the *site-size-driven* path with `preset_education_modest()`,
the SE marginal inherits the right-skew of the truncated-Gamma site-
size marginal: small-$n_j$ sites correspond to large-$\widehat{se}_j$
sites. The overlay shows the deterministic curve $\kappa/n_j$ at
$\kappa = 4$ across the realized $n_j$ range — every realized point
sits exactly on the curve, not approximately, because the kappa-rate
identity is deterministic at the package's testable invariant T3.

```{r plot-A-build, results = "hide"}
set.seed(1L)
dat_A <- sim_multisite(preset_education_modest(), seed = 1L)
kappa_A <- compute_kappa(p = 0.5, R2 = 0, var_outcome = 1)
df_A <- tibble::tibble(
  n_j   = dat_A$n_j,
  se2_j = dat_A$se2_j
)
n_grid <- seq(min(df_A$n_j), max(df_A$n_j), length.out = 200)
df_curve <- tibble::tibble(
  n_j      = n_grid,
  se2_curve = kappa_A / n_grid
)
```

```{r plot-A, fig.height = 5.6, fig.cap = "Paradigm A realized se squared values from preset_education_modest at J = 50 (filled dots) sit exactly on the kappa-over-n curve (solid line, kappa = 4). The histogram on the upper margin shows the right-skewed se squared distribution that emerges from the truncated-Gamma site-size marginal. Confirms the kappa-rate identity (T3) and the right-skew shape consequence of n_j heterogeneity at CV = 0.5.", fig.alt = "Two-panel composite figure showing a histogram of se squared values on top and a scatter of se squared against n_j with a kappa-over-n curve overlaid below. Realized points sit exactly on the curve."}
p_top <- ggplot(df_A, aes(x = se2_j)) +
  geom_histogram(bins = 18, fill = "#1B4965", colour = "white", alpha = 0.85) +
  labs(x = NULL, y = "count") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_bottom <- ggplot(df_A, aes(x = n_j, y = se2_j)) +
  geom_line(data = df_curve, aes(x = n_j, y = se2_curve),
            colour = "grey40", linewidth = 0.6) +
  geom_point(alpha = 0.85, size = 2, colour = "#1B4965") +
  labs(
    x = expression(n[j]~"  (per-site sample size)"),
    y = expression(widehat(se)[j]^{2}~"  (sampling variance)")
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

# Compose without depending on patchwork (soft dep).
gridExtra::grid.arrange(
  p_top,
  p_bottom,
  ncol = 1, heights = c(1, 4)
)
```

The single curve in the lower panel is the closed-form
$\widehat{se}_j^{\,2} = \kappa/n_j$ at $\kappa = 4$. Every realized
point sits on the curve to floating-point precision (not
approximately) because the kappa-rate identity is deterministic at
Layer 2 — there is no Layer-2 noise on top of the closed form. The
upper-panel histogram is the marginal of $\widehat{se}_j^{\,2}$
realized at $J = 50$: the right tail at large
$\widehat{se}_j^{\,2}$ comes from the small-$n_j$ tail of the
truncated-Gamma site-size marginal, and the bulk near
$\widehat{se}_j^{\,2} \approx 0.07$ comes from the centre of mass at
$n_j \approx 50$. The kappa-over-$n$ curve and the realized
distribution carry exactly the same information — the right-skew is
the sole consequence of the truncated-Gamma right-skew.

### 5.2 Plot 2 — Paradigm B realized SE distribution

Under the *direct-precision* path with $J = 50$, $I = 0.30$, $R = 2$,
$\sigma_\tau = 0.20$, the SE marginal is the deterministic log-even
grid. The overlay shows the implied geometric mean
$\mathrm{GM} = \sigma_\tau^{\,2}(1 - I)/I = 0.0933$ as a vertical line
and the implied min/max boundaries $\mathrm{GM} \cdot R^{\pm 1/2}$ as
two further vertical lines. Every realized $\widehat{se}_j^{\,2}$
falls inside the $[\mathrm{GM}/\sqrt{R}, \mathrm{GM}\sqrt{R}]$
interval, the multiset's geometric mean lands on the central line,
and the multiset's max-over-min ratio equals $R$ exactly.

```{r plot-B-build, results = "hide"}
set.seed(1L)
dat_B <- sim_meta(J = 50L, I = 0.30, R = 2, sigma_tau = 0.20, seed = 1L)
sigma_tau_B <- 0.20
I_B <- 0.30
R_B <- 2
GM_target <- sigma_tau_B^2 * (1 - I_B) / I_B
df_B <- tibble::tibble(se2_j = dat_B$se2_j)
GM_realized <- exp(mean(log(df_B$se2_j)))
range_B <- range(df_B$se2_j)
```

```{r plot-B-summary}
GM_target
GM_realized
range_B
range_B[2] / range_B[1]
```

The realized geometric mean equals the target $0.0933$ to floating-
point precision; the realized range $[0.0660, 0.1320]$ lands on
$[\mathrm{GM}/\sqrt{2}, \mathrm{GM}\sqrt{2}]$ exactly; the
max-over-min ratio is 2, the requested $R$.

```{r plot-B, fig.cap = "Paradigm B realized se squared values from sim_meta(J = 50, I = 0.30, R = 2, sigma_tau = 0.20) (histogram). Vertical solid line: implied geometric mean GM = sigma_tau squared times (1 - I) over I = 0.0933. Dashed lines: GM divided by square root of R and GM times square root of R, the implied min and max. Confirms the deterministic grid hits the (I, R)-implied target exactly: every realized point lies inside the bracket, GM is on the solid line, and max over min equals 2.", fig.alt = "Histogram of se squared values from a Paradigm B simulation. A vertical line marks the geometric mean target; two dashed vertical lines mark the implied min and max. Histogram bars span exactly between the dashed lines."}
ggplot(df_B, aes(x = se2_j)) +
  geom_histogram(bins = 18, fill = "#62929E", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = GM_target, colour = "#1B4965",
             linewidth = 0.7) +
  geom_vline(xintercept = GM_target / sqrt(R_B), linetype = "dashed",
             colour = "grey40") +
  geom_vline(xintercept = GM_target * sqrt(R_B), linetype = "dashed",
             colour = "grey40") +
  annotate("text", x = GM_target, y = Inf, vjust = 1.4,
           label = "GM target", colour = "#1B4965", size = 3.2) +
  annotate("text", x = GM_target / sqrt(R_B), y = Inf, vjust = 1.4,
           label = "GM / sqrt(R)", colour = "grey30", size = 3.0) +
  annotate("text", x = GM_target * sqrt(R_B), y = Inf, vjust = 1.4,
           label = "GM * sqrt(R)", colour = "grey30", size = 3.0) +
  labs(
    x        = expression(widehat(se)[j]^{2}~"  (sampling variance)"),
    y        = "count",
    title    = "Paradigm B — direct-precision SE marginal at J = 50",
    subtitle = "I = 0.30, R = 2, sigma_tau = 0.20  ->  GM = 0.0933, range = [0.0660, 0.1320]"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

Three things to read off the figure. First, the histogram is bounded
exactly by the two dashed vertical lines — the $J = 50$ realized
$\widehat{se}_j^{\,2}$ values fall inside
$[\mathrm{GM}/\sqrt{R}, \mathrm{GM}\sqrt{R}]$ by construction, with
the endpoints attained at $j = 1$ and $j = J$. Second, the histogram's
shape is approximately uniform on the log scale (the design choice was
*log-even* spacing) — the visible mild non-uniformity in linear bins
reflects the log-to-linear stretch, not a design failure. Third, the
solid central line at $\mathrm{GM} = 0.0933$ sits exactly at the
geometric centre of the multiset.

The contrast between the two plots is exactly the contrast between
the two paradigms: the *site-size-driven* path produces an SE
distribution whose shape is a *consequence* of the site-size choice
(right-skewed, with a long tail from small-$n_j$ sites); the
*direct-precision* path produces an SE distribution whose shape is
*specified* (a log-even grid bounded by the (I, R) target).

## 6. Identifiability and switching

The two paradigms share the same Layer 2 *output* signature —
`se2_j`, `se_j`, and an `n_j` column populated under Paradigm A and
`NA_integer_` under Paradigm B — but their *input* signatures are
disjoint. Paradigm A consumes $(\bar n_j, \mathrm{CV}, n_{j,\min}, p,
R^2, s_Y^2, \mathrm{engine})$; Paradigm B consumes $(I, R, \mathrm{shuffle},
\mathrm{se\_fn}, \mathrm{se\_args})$. The package keeps the two
input vocabularies separated at the front door: handing
`sim_meta()` a Paradigm A argument or `sim_multisite()` a Paradigm B
argument is rejected with a coherence error before any compute
happens.

```{r wrong-door-meta, error = TRUE}
sim_meta(J = 50L, nj_mean = 80, cv = 0.5, sigma_tau = 0.20, seed = 1L)
```

The error message names the wrong-door arguments (`nj_mean`, `cv`)
and points to the correct front door (`sim_multisite()`). The same
coherence check runs on the sister wrapper:

```{r wrong-door-multisite, error = TRUE}
sim_multisite(J = 50L, I = 0.30, R = 2, sigma_tau = 0.20, seed = 1L)
```

Again, the wrong-door arguments (`I`, `R`) are named and the
caller is redirected to `sim_meta()`. The construction-time check is
deliberate: cross-paradigm dispatch is strict by design, so the
provenance attribute on a returned object always reflects exactly
one paradigm and never an implicit mix. A simulation labelled
`paradigm = "site_size"` in the diagnostics block was generated by
the *site-size-driven* Layer 2 generator, not by Paradigm B with
post-hoc relabelling.

The same paradigm-strict contract holds at the
[`multisitedgp_design()`](../reference/multisitedgp_design.html)
level. Constructing a design with `paradigm = "site_size"` and then
handing it to `sim_meta()` — or constructing a design with
`paradigm = "direct"` and handing it to `sim_multisite()` — is also
rejected. The wrapper checks `design$paradigm` against its own
locked paradigm at entry. Below, an explicit Paradigm A design is
forced through `sim_meta()`:

```{r wrong-door-design, error = TRUE}
des_A <- multisitedgp_design(
  paradigm = "site_size",
  J        = 50L,
  sigma_tau = 0.20,
  nj_mean  = 50,
  cv       = 0.50,
  nj_min   = 10L
)
sim_meta(des_A, seed = 1L)
```

The error message names the design's paradigm (`"site_size"`) and
points to the correct sister wrapper. The package never silently
re-labels a design; it always errors and lets the caller decide.

Why this matters. A simulation pipeline that conflates the two
paradigms — for example, by dispatching on a flag and silently
swapping the Layer 2 generator — would let provenance drift on
re-run: the same code path could yield different `attr(x,
"diagnostics")$paradigm` values depending on argument order. The
paradigm-strict front door rules that drift out at construction time.
This is the operational form of the M1 §4 four-layer factorization:
Layer 2 is the only differentiating layer between the two paths, so
keeping the two front doors disjoint is the cleanest place to draw
the line.

The single point where the two paths *can* be reasoned about jointly
is the realized informativeness scalar $I$. Under Paradigm A,
$I$ is a *consequence* of $(\bar n_j, \mathrm{CV})$ via the Jensen
gap on $\kappa/n_j$; under Paradigm B, $I$ is an *input*. The
helper [`compute_I()`](../reference/compute_I.html) reads the
realized $I$ from any `se2_j` vector regardless of which path
generated it:

```{r compute-I-cross-paradigm}
compute_I(dat_A$se2_j, sigma_tau = 0.20)
compute_I(dat_B$se2_j, sigma_tau = 0.20)
```

The Paradigm A simulation lands at $I \approx 0.303$ and the Paradigm
B simulation at exactly $0.300$ — the diagnostic helper itself is
paradigm-blind, but the simulations it reads are paradigm-strict at
the front door.

## 7. Where to next

- [A1: Getting started](a1-getting-started.html) — applied
  walkthrough of the *site-size-driven* path from preset to
  diagnostics.
- [A6: Case study — multisite trial](a6-case-study-multisite.html) —
  full Paradigm A workflow including precision dependence and
  shrinkage diagnostics.
- [A7: Case study — meta-analysis](a7-case-study-meta-analysis.html) —
  full Paradigm B workflow, including handoff to `metafor` via
  [`as_metafor()`](../reference/as_metafor.html).
- [M1: The two-stage DGP](m1-statistical-dgp.html) — the formal
  factorization; this vignette zoomed in on Layer 2 and how the two
  Layer-2 generators differ.
- [M4: Precision dependence theory](m4-precision-dependence-theory.html)
  — the rank, copula, and hybrid alignment methods that sit on top of
  Layer 2 in both paradigms.

## References

The site-size-driven path's $\kappa$ closed form and Engine A1's
bit-identical reproducibility contract come from
@lee2025improving. The empirical-Bayes context for treating reported
sampling standard errors as known — relevant to both paths but
especially to the direct-precision path's plug-in of
$(\hat\tau_j, \widehat{se}_j)$ pairs from a meta-analytic corpus —
is collected in @walters2024empirical. The education-trial
calibration heritage that motivates the
[`preset_education_modest()`](../reference/preset_education_modest.html)
and sister Paradigm A presets is from @weiss2017howmuch.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
