---
title: "M4 · Precision dependence — three injection methods"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists building simulations whose generative process must
  encode a controlled coupling between site effects and reported
  precision, this vignette walks through the three precision-dependence
  injection methods `multisiteDGP` exposes — `align_rank_corr()`,
  `align_copula_corr()`, and `align_hybrid_corr()` — one section per
  method. Each section gives the mathematical form, what the method
  preserves exactly, what it distorts, and when to choose it. You leave
  with three side-by-side scatterplots at a common target
  $\rho_S = 0.40$, the rationale for the package's hybrid default, and
  a decision rubric for picking among the three.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M4 Precision dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why independence fails in real multisite trials

The generative skeleton on which `multisiteDGP` is built treats the
latent site effect $\tau_j$ and the reported sampling variance
$\widehat{se}_j^{\,2}$ as independent. That independence is the
*design's null position*: under perfect random allocation, no selection,
and uniform implementation fidelity, knowing how precisely a site was
estimated tells you nothing about that site's true effect. The
[M1: The two-stage DGP](m1-statistical-dgp.html) vignette derives the
joint distribution under that null; the [M3: Margin and SE
models](m3-margin-se-models.html) vignette derives the two precision
margins separately. For a wide swath of pedagogically clean simulation
work, that independence is fine.

Reality, however, departs. The blueprint's chapter on precision
dependence (Ch. 7, _Precision Dependence_) collects three mechanisms by
which empirical multisite trials systematically violate the null. Small
studies often report systematically larger or smaller effects than
large ones — the canonical funnel-asymmetry pattern that motivates
publication-bias diagnostics in meta-analysis [@walters2024empirical].
Selection on observable site characteristics (catchment breadth,
treatment fidelity, baseline outcome variance) couples site-level
sample size with site-level effect heterogeneity. And adaptive
implementation — sites scaling up or down enrolment based on interim
signals — can induce coupling between $\tau_j$ and $\widehat{se}_j^{\,2}$
even when the original allocation was random. When a methodologist
wants to ask "is my estimator robust to the kind of dependence I see
in the wild?", the simulation must be able to inject the dependence on
purpose, with a controlled target [@lee2025improving].

That is what Layer 3 does. Three exported alignment functions —
`align_rank_corr()`, `align_copula_corr()`, and `align_hybrid_corr()` —
permute the precision margin against the latent effect to produce a
target rank correlation between $z_j$ (the standardized residual
effect) and $\widehat{se}_j^{\,2}$. The three differ in *how* they
permute and consequently in *what* they preserve exactly, what they
distort, and what they cost in compute. The rest of this vignette
walks them through one at a time.

```{r common-design}
# A common reference design used throughout this vignette: J = 80
# sites, the package-canonical Gaussian Layer 1, a moderate site-size
# margin (mean 60, CV 0.4), no covariates. We will plug different
# Layer 3 dependence injectors into this fixed input skeleton.
common_args <- list(
  J         = 80L,
  sigma_tau = 0.20,
  nj_mean   = 60,
  cv        = 0.4,
  seed      = 1L
)
target <- 0.40

# Independence baseline — for every primary call that follows, we want
# the reader to see what the un-injected joint distribution looks like.
des_none <- do.call(multisitedgp_design,
                    c(common_args, dependence = "none"))
dat_none <- sim_multisite(des_none)
dat_none
```

The independence baseline is what every dependence-injection method
permutes *out of*. The realized residual Spearman in `dat_none` will
sit close to zero up to Monte Carlo noise — the print method's
`rho_S` line confirms this. From here, three sections.

## 2. The rank-permutation method (`align_rank_corr`)

### 2.1 Mathematical form

The rank method takes the input pair $(z_j, \widehat{se}_j^{\,2})$
and proposes random pair-swaps on the precision column. A swap is
accepted if the post-swap residual Spearman correlation between $z_j$
and $\widehat{se}_j^{\,2}$ is closer to the user's target `rank_corr`
than before; otherwise rejected. The procedure terminates when the
realized Spearman lands within `tol` of the target or when `max_iter`
swaps have been proposed.

Formally, write $\sigma$ for the realized site-to-grid permutation of
the pre-alignment precision multiset. The rank aligner returns the $\sigma$
that minimizes

$$
\bigl|\rho_S\bigl(z_j,\, \widehat{se}_{\sigma(j)}^{\,2}\bigr) - \rho_S^\star\bigr|
$$

over the swap-reachable subspace, with $\rho_S^\star$ the user's target.
Because $\sigma$ is by construction a permutation of the pre-alignment
precision values, the empirical marginal of $\widehat{se}_j^{\,2}$
is bit-identical to what came in from Layer 2.

### 2.2 What the method preserves and distorts

**Preserves exactly:** the empirical marginal of $\widehat{se}_j^{\,2}$
(as a multiset) and trivially the empirical marginal of $z_j$ (which
is never moved). No site has its precision value rounded, smoothed, or
resampled — only relabeled.

**Distorts:** the *shape* of the joint distribution. The rank aligner
cares only about the ordering of $\widehat{se}_j^{\,2}$ relative to
$z_j$; it has no copula structure, no notion of joint smoothness, and
no tail-dependence guarantee. For most applied targets in
$|\rho_S^\star| \le 0.7$ this is irrelevant — the realized scatter
looks like a Gaussian-copula scatter to the eye — but at the boundary
$|\rho_S^\star| \to 1$ the rank aligner can produce visibly
"step-shaped" plots because the swap-reachable subspace becomes
thinner.

**Cost in compute:** one acceptance test per proposed swap. For
$J = 80$ and a target of $\rho_S = 0.40$ the algorithm typically
converges in a handful of swaps from a near-zero baseline.

```{r rank-method}
# align_rank_corr through the multisitedgp_design front door.
des_rank <- multisitedgp_design(
  J         = 80L,
  sigma_tau = 0.20,
  nj_mean   = 60,
  cv        = 0.4,
  dependence = "rank",
  rank_corr  = target,
  seed       = 1L
)
dat_rank <- sim_multisite(des_rank)
dat_rank
```

The print method's `rho_S` line shows `target=0.400, realized=0.401,
PASS`. The hill-climb landed on the target within tolerance after a
handful of swaps. The detailed dependence diagnostics tell the same
story:

```{r rank-diag}
attr(dat_rank, "diagnostics")$dependence_diagnostics
```

`converged = TRUE`, `iterations = 4`, `residual = 0.00132`. For an
$8\times 10^4$-cell swap budget the algorithm spent four. The
realized Spearman is exact within the user's tolerance.

```{r rank-marginal-check}
# Marginal preservation: the rank method only reorders precision values,
# so the sorted multisets must coincide bit-for-bit with the baseline.
all.equal(sort(dat_rank$se2_j), sort(dat_none$se2_j))
```

The sorted precision multiset coincides with the baseline. No
precision value was created, smoothed, or rounded.

### 2.3 When to choose `align_rank_corr`

Choose the rank method when:

- You want **exact** preservation of the precision multiset (the values
  are calibrated targets — $\widehat{se}_j^{\,2}$ from a real trial,
  for example — and you cannot tolerate even Monte Carlo perturbation).
- The target $|\rho_S^\star|$ is moderate (under $0.7$ or so) and the
  realized correlation must hit the target tightly.
- You want the fastest no-iteration smoke test in scenarios where
  copula structure is not part of the analysis.

Avoid the rank method when:

- The target sits near $\pm 1$. The swap-reachable subspace narrows
  and the algorithm may stall before converging.
- You need a smooth copula-shaped joint scatter for a downstream
  density-based diagnostic.

## 3. The Gaussian-copula method (`align_copula_corr`)

### 3.1 Mathematical form

The copula method draws a correlated latent score $L_j$ for each site
from a bivariate Gaussian with target Pearson correlation
`pearson_corr` and the rank-normalized $z_j$ on the other axis. It
then sorts the empirical $\widehat{se}_j^{\,2}$ multiset and assigns
the values to sites in the order $L_j$ implies. Concretely, with
$U_j = \Phi^{-1}\!\bigl(\text{rank}(z_j)/(J + 1)\bigr)$ and a draw

$$
L_j = \rho^\star\, U_j + \sqrt{1 - (\rho^\star)^2}\, V_j,
\qquad V_j \stackrel{\text{iid}}{\sim} \mathcal{N}(0, 1),
$$

the aligner pairs the $j$th-order-statistic of $\widehat{se}_j^{\,2}$
with the site whose $L_j$ has the $j$th rank. No iteration; one pass.

The package's `align_copula_corr()` calls the target a *latent
Gaussian-copula Pearson correlation*. The realized residual Spearman
is approximately the user's `pearson_corr` (the two are within Monte
Carlo noise at large $J$ for well-behaved marginals) but not exactly.

### 3.2 What the method preserves and distorts

**Preserves exactly:** the empirical marginal of $\widehat{se}_j^{\,2}$
as a multiset (the assignment is by rank of the latent score, so the
sorted values are unchanged). And the marginal of $z_j$, never moved.

**Distorts:** the realized residual Spearman correlation. The aligner
targets the *latent* Pearson correlation between $L_j$ and the
rank-normalized $z_j$; the implied residual Spearman is approximately
$(6/\pi)\arcsin(\rho^\star/2)$ but not exactly, because finite-$J$
ranks are discrete. At $J = 80$ and $\rho^\star = 0.40$, the realized
residual Spearman lands modestly above target — the residual diagnostic
will show roughly a tenth of a correlation point of slack.

**Cost in compute:** one $J$-vector latent draw and one sort. The
copula aligner is the fastest of the three by a wide margin.

```{r copula-method}
# align_copula_corr — note that the target argument is `pearson_corr`,
# not `rank_corr`. The copula targets a latent Gaussian Pearson
# correlation, and the realized residual Spearman follows from the
# arcsine mapping, with finite-J slack.
des_copula <- multisitedgp_design(
  J            = 80L,
  sigma_tau    = 0.20,
  nj_mean      = 60,
  cv           = 0.4,
  dependence   = "copula",
  pearson_corr = target,
  seed         = 1L
)
dat_copula <- sim_multisite(des_copula)
dat_copula
```

The print method's `rho_S` line shows `target=0.000, realized=0.438,
FAIL`. Two things to read: (a) the design's *registered* rank-target
slot is empty (the user passed `pearson_corr`, not `rank_corr`), so the
diagnostic infrastructure has nothing to compare the realized Spearman
against — that is what `target=0.000` means in this row; and (b) the
realized Spearman is $0.438$, modestly above the latent Pearson target
of $0.40$. The detailed diagnostics give the full story:

```{r copula-diag}
attr(dat_copula, "diagnostics")$dependence_diagnostics
```

The aligner reports `target = 0.40` (interpreted as latent Gaussian
Pearson), `achieved = 0.503` on the latent Pearson scale, and
`achieved_spearman = 0.438` on the residual Spearman scale. The
`residual = 0.103` is the slack on the latent Pearson; the
finite-$J$ Spearman slack is smaller. `mode =
"empirical_rank_pairing"` documents the assignment rule;
`preservation = "empirical_multiset"` confirms that the precision
margin is unchanged.

```{r copula-marginal-check}
all.equal(sort(dat_copula$se2_j), sort(dat_none$se2_j))
```

Bit-identical sorted multisets, again. The copula method preserves
the precision values as exactly as the rank method does — the
empirical-rank pairing variant of the copula aligner uses ordering to
assign, never resampling.

### 3.3 When to choose `align_copula_corr`

Choose the copula method when:

- The simulation lives on the *direct-precision* path (Paradigm B in
  the blueprint) and the natural target is a latent
  Gaussian-copula correlation rather than a realized rank correlation.
- You need the fastest method (no iteration) and can tolerate
  finite-$J$ slack on the realized residual Spearman.
- Downstream diagnostics consume the Gaussian-copula joint shape
  itself (density-overlay plots, copula-density estimators).

Avoid the copula method when:

- The realized residual Spearman must hit the target tightly. Use
  hybrid (the next section) instead.
- You want to specify the target as a Spearman directly. The copula
  takes a latent Pearson; the back-mapping
  $\rho^\star = 2\sin(\pi \rho_S / 6)$ gets you part of the way, but
  the finite-$J$ slack is on the realized side.

## 4. The hybrid method (`align_hybrid_corr`)

### 4.1 Mathematical form

The hybrid method runs the copula aligner first as an initialization
pass, then runs the rank aligner on the result as a polishing pass.
The copula stage lands the realized scatter in the right neighborhood
of the joint distribution — Gaussian-copula-shaped, smooth, marginals
preserved. The rank stage spends a handful of swaps cleaning up the
finite-$J$ slack on the realized residual Spearman, driving it to
within `tol` of `rank_corr`.

$$
\sigma = \sigma_{\text{rank}}\bigl(\sigma_{\text{copula}}(\text{input})\bigr).
$$

The composition keeps the copula stage's Gaussian-copula shape (modulo
a small polishing perturbation) and inherits the rank stage's tight
target accuracy. The `init` and `polish` arguments expose the
two-stage structure: `init = "copula"` (default) gives the copula
initialization; `polish = "hill_climb"` (default) gives the rank
polish. Either stage can be replaced or skipped.

### 4.2 What the method preserves and distorts

**Preserves exactly:** the empirical marginal of $\widehat{se}_j^{\,2}$.
Both stages assign by rank, so the sorted precision multiset is
unchanged.

**Preserves approximately:** the Gaussian-copula joint shape. The
polishing pass perturbs the copula scatter slightly — far less than a
pure rank aligner from a cold start, because most swaps are not needed.

**Distorts:** essentially nothing of practical concern at typical
applied targets. At $J = 80$ and $\rho_S^\star = 0.40$, the realized
residual Spearman lands within `tol = 0.02` of the target, the
precision multiset is bit-identical, and the joint scatter retains the
copula stage's smooth shape.

**Cost in compute:** the copula stage's $J$-vector latent draw plus
the rank stage's polishing swaps. For typical applied targets the
polishing budget is negligible. At boundary targets
($|\rho_S^\star| \to 1$) the polishing burden grows but stays under
the pure-rank cost, because the copula initialization placed the
scatter in the right neighborhood.

```{r hybrid-method}
# align_hybrid_corr is the package default. The dependence_spec slot
# of the design carries `dependence = "hybrid"` automatically when the
# user passes `rank_corr` without specifying a method, but here we
# pass it explicitly for clarity.
des_hybrid <- multisitedgp_design(
  J          = 80L,
  sigma_tau  = 0.20,
  nj_mean    = 60,
  cv         = 0.4,
  dependence = "hybrid",
  rank_corr  = target,
  seed       = 1L
)
dat_hybrid <- sim_multisite(des_hybrid)
dat_hybrid
```

The print method's `rho_S` line shows `target=0.400, realized=0.400,
PASS`. The hybrid landed exactly on target. The detailed diagnostics
expose the two-stage internal structure:

```{r hybrid-diag}
attr(dat_hybrid, "diagnostics")$dependence_diagnostics
```

The list is structured as `init` (the copula stage's diagnostics) and
`polish` (the rank stage's). The init stage reports the copula
target it received from the front door (the rank target was mapped to
its latent-Pearson cousin by the design constructor) and the realized
Spearman it landed on after one pass. The polish stage took
`iterations = 1` to drive the realized Spearman to `0.4000305`,
landing with `residual = 3.05e-05`. Notice how few swaps polishing
needed — copula initialization did the heavy lifting.

```{r hybrid-marginal-check}
all.equal(sort(dat_hybrid$se2_j), sort(dat_none$se2_j))
```

Bit-identical sorted multisets. The hybrid preserves the precision
margin as exactly as either pure method does.

### 4.3 When to choose `align_hybrid_corr`

Choose the hybrid method when:

- You want both tight realized-Spearman accuracy *and* a
  Gaussian-copula-shaped joint scatter.
- The simulation will be cited or audited (the hybrid is the package
  default; reaching for it is the "no surprise" choice).
- The target sits in the typical applied range
  ($|\rho_S^\star| \le 0.7$) where the copula initialization places
  the scatter close to target and the polishing budget is small.

Avoid the hybrid method when:

- You need raw speed and can tolerate the copula-only slack — use
  `align_copula_corr()` directly.
- The simulation specifically demands no Monte Carlo noise on the
  joint scatter (e.g., a deterministic-replay smoke test) — use
  `align_rank_corr()` from a cold start, which uses no `rnorm()`
  draws.

## 5. Side-by-side scatter comparison

The three methods all target the same realized residual correlation —
roughly $\rho_S = 0.40$ — but the *shape* of the joint scatter is
where they differ. Building three plots on the same axes shows what
each method emphasizes.

```{r build-scatter-data}
# Stack the three datasets with a method label for ggplot's facet grid.
to_scatter <- function(dat, label) {
  data.frame(
    method = label,
    z_j    = dat$z_j,
    se2_j  = dat$se2_j
  )
}

scatter_df <- rbind(
  to_scatter(dat_rank,   "rank"),
  to_scatter(dat_copula, "copula"),
  to_scatter(dat_hybrid, "hybrid")
)
scatter_df$method <- factor(
  scatter_df$method,
  levels = c("rank", "copula", "hybrid")
)
head(scatter_df)
```

```{r plot-rank, fig.cap = "Rank-method scatter at target $\\rho_S = 0.40$. Higher $\\widehat{se}_j^{\\,2}$ pairs with higher $z_j$ across the full range, and the precision values along the y-axis match the baseline multiset bit-for-bit, confirming the realized residual Spearman lands within tolerance of 0.40 by reordering — not resampling.", fig.alt = "Scatterplot of standardized residual z_j on the x-axis and sampling variance se2_j on the y-axis under the rank-permutation method at target Spearman 0.40; points spread across a positive monotone band."}
ggplot(subset(scatter_df, method == "rank"),
       aes(x = z_j, y = se2_j)) +
  geom_point(alpha = 0.7, color = "steelblue", size = 2) +
  geom_smooth(method = "loess", se = FALSE,
              color = "darkblue", linewidth = 0.7) +
  labs(
    title = "Rank-permutation method (rank_corr = 0.40)",
    x = expression(z[j]~"(standardized residual effect)"),
    y = expression(widehat(se)[j]^2~"(sampling variance)")
  ) +
  theme_minimal(base_size = 11)
```

```{r plot-copula, fig.cap = "Copula-method scatter at the same target. The Gaussian-copula structure produces a smoother, more elliptical cloud than the rank scatter; the realized residual Spearman lands modestly above target ($\\widehat\\rho_S \\approx 0.44$), confirming the finite-$J$ slack flagged in the dependence diagnostics.", fig.alt = "Scatterplot of standardized residual z_j on the x-axis and sampling variance se2_j on the y-axis under the Gaussian-copula method at target Spearman 0.40; the cloud is smoother and more elliptical than the rank-method version, with realized correlation slightly above target."}
ggplot(subset(scatter_df, method == "copula"),
       aes(x = z_j, y = se2_j)) +
  geom_point(alpha = 0.7, color = "darkorange", size = 2) +
  geom_smooth(method = "loess", se = FALSE,
              color = "darkred", linewidth = 0.7) +
  labs(
    title = "Gaussian-copula method (pearson_corr = 0.40)",
    x = expression(z[j]~"(standardized residual effect)"),
    y = expression(widehat(se)[j]^2~"(sampling variance)")
  ) +
  theme_minimal(base_size = 11)
```

```{r plot-hybrid, fig.cap = "Hybrid-method scatter at the same target. The cloud retains the copula method's smooth elliptical shape but the realized residual Spearman lands within 0.0001 of 0.40 after a single polishing swap, confirming the recommended default delivers both joint-shape smoothness and target accuracy.", fig.alt = "Scatterplot of standardized residual z_j on the x-axis and sampling variance se2_j on the y-axis under the hybrid (copula-init plus rank-polish) method at target Spearman 0.40; the cloud retains the copula's elliptical shape while the realized correlation matches target within 0.0001."}
ggplot(subset(scatter_df, method == "hybrid"),
       aes(x = z_j, y = se2_j)) +
  geom_point(alpha = 0.7, color = "darkgreen", size = 2) +
  geom_smooth(method = "loess", se = FALSE,
              color = "darkgreen", linewidth = 0.7) +
  labs(
    title = "Hybrid method (rank_corr = 0.40)",
    x = expression(z[j]~"(standardized residual effect)"),
    y = expression(widehat(se)[j]^2~"(sampling variance)")
  ) +
  theme_minimal(base_size = 11)
```

The three plots tell a consistent story. The rank method's scatter
shows the same underlying $(z_j, \widehat{se}_j^{\,2})$ pairs, just
permuted. The copula method's scatter looks smoother because the
assignment was driven by a continuous latent-score draw rather than
by individual swap moves. The hybrid method's scatter inherits the
copula stage's smoothness and the rank stage's target accuracy, which
is why it is the package default.

To make the *exact* match of marginals across methods concrete,
sorting the precision column from each dataset gives the same
sequence by construction:

```{r marginal-grid}
# Same multiset, different assignment.
identical(
  list(sort(dat_rank$se2_j), sort(dat_copula$se2_j),
       sort(dat_hybrid$se2_j), sort(dat_none$se2_j)),
  rep(list(sort(dat_none$se2_j)), 4)
)
```

Four sorted vectors, all identical. The methods differ only in *which*
site receives *which* precision value.

## 6. Recommended default and decision rubric

The package's recommended default is **hybrid**. For production
designs, the hybrid combines:

- The copula initialization's smooth Gaussian-copula joint shape (the
  scatter looks like what most readers expect of a "correlated $z$ and
  $\widehat{se}^{\,2}$" sample).
- The rank polishing pass's tight target accuracy (the realized
  residual Spearman lands within `tol = 0.02` of target).
- Bit-exact preservation of the pre-alignment precision multiset.

The cost is modest: one $J$-vector latent Gaussian draw plus a small
number of polishing swaps. For typical applied targets (the bulk of
applied multisite-trial work, and certainly the JEBS-reproduction band
$|\rho_S| \le 0.5$ studied in @lee2025improving), the polishing
budget is negligible — at $J = 80$ and target $0.40$, polishing took
a single iteration.

The other two methods are not deprecated; they have specific niches:

- **`align_rank_corr()`** is the right pick for fast smoke tests where
  the joint shape is irrelevant and only the realized Spearman matters,
  or for deterministic-replay scenarios where no `rnorm()` draws are
  desired.
- **`align_copula_corr()`** is the right pick for direct-precision
  designs in meta-analysis settings where the natural target is a
  latent Pearson correlation and the finite-$J$ slack on the realized
  Spearman is acceptable.

A compact decision rule:

| Need                                    | Method                  | Rationale                                                     |
|-----------------------------------------|-------------------------|---------------------------------------------------------------|
| Production / cited simulation           | `align_hybrid_corr()`   | Default; smooth shape + tight target accuracy                 |
| Fast smoke test, joint shape irrelevant | `align_rank_corr()`     | Fewest swaps; no `rnorm()` draws; exact multiset              |
| Direct-precision designs                | `align_copula_corr()`   | Latent-Pearson target is natural; one-pass speed              |
| Bit-deterministic replay                | `align_rank_corr()`     | Hill-climb proposes via `sample.int()`; copula uses `rnorm()` |
| Boundary targets ($|\rho_S| \to 1$)     | `align_hybrid_corr()`   | Copula init narrows the search; pure rank may stall           |

For users who pass arguments to `multisitedgp_design()` without
specifying `dependence` explicitly, the constructor selects the
appropriate method based on which target argument is supplied. Passing
`rank_corr` without `dependence` triggers an error so the user must
opt in to a method by name.

## 7. A note on what is not modelled

All three methods produce joint scatters with **zero asymptotic upper
tail dependence**:

$$
\lambda_U \;=\; \lim_{u \to 1^-}
\Pr\!\bigl(F_{\widehat{se}^{\,2}}(\widehat{se}_j^{\,2}) > u
        \;\bigm|\;
        F_{\tau}(\tau_j) > u\bigr)
\;=\; 0.
$$

Real-world mechanisms that one might suspect of inducing tail-coupling
(rare-population sites with simultaneously large effects and large
sampling variance) are not represented by any of the three methods.
A Student-$t$ copula extension is on the medium-term roadmap (the
blueprint Ch. 7 §4 _Tail Dependence Caveat_ flags this as a future
item).
For JEBS-band simulations the body-of-distribution dependence captured
by the three current methods is sufficient — the empirical-Bayes
estimators studied in @lee2025improving are most sensitive to the
body, not the tails.

## 8. Where to next

- [A4: Covariates and precision dependence](a4-covariates-precision-dependence.html)
  shows the three methods composing with the covariate machinery —
  what changes when $z_j$ is the covariate-residualized standardized
  effect rather than the raw effect.
- [A6: Case study — multisite trial](a6-case-study-multisite.html)
  walks through a full applied workflow that uses
  `align_hybrid_corr()` as the default Layer 3 injector and shows the
  downstream diagnostics that read off the dependence target.
- [M3: Margin and SE models](m3-margin-se-models.html) derives the two
  precision margins on which Layer 3 acts — site-size-driven and
  direct-precision — and shows the cross-paradigm guards.
- [M7: Reproducibility and provenance](m7-reproducibility-provenance.html)
  covers the seed-and-hash discipline that makes hybrid
  simulations bit-replayable across users and machines despite the
  copula stage's `rnorm()` draws.

The reference pages give the full argument lists and contracts:
[`align_rank_corr()`](../reference/align_rank_corr.html),
[`align_copula_corr()`](../reference/align_copula_corr.html),
[`align_hybrid_corr()`](../reference/align_hybrid_corr.html), and the
front-door wrapper [`multisitedgp_design()`](../reference/multisitedgp_design.html).

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
