---
title: "M5 · Custom G distributions"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For methodologists who need a latent site-effect distribution
  outside the eight-shape catalog — a horseshoe residual, a
  deconvolution draw from a sister study, an externally fitted
  Dirichlet process mixture. The user-callback path threads a custom
  shape through `multisiteDGP` under one of two conventions: A returns
  standardized residuals (the package owns the location-scale wrapper),
  B returns response-scale effects (the callback owns the marginal).
  You leave with a worked example per convention, the audit recipe
  that catches contract violations, and the bridge pattern for routing
  a DPM draw into the package.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M5 Custom G distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The problem — when the catalog runs out

The eight built-in $G$ distributions documented in
[M2: G-distribution catalog and standardization](m2-g-distribution-catalog.html)
cover most applied shapes: Gaussian, Student-$t$, skew-normal,
asymmetric Laplace, two-component mixture, point-mass-plus-slab, the
user-callback escape hatch, and the deferred DPM stub. The case for
this vignette is the residual one: you want a shape the catalog does
not have. Common triggers are a horseshoe-style residual with both a
sharp peak and heavy tails, a finite mixture with three or more
components, an empirical residual distribution deconvolved from a
sister trial, or a nonparametric draw from a Dirichlet process mixture
fit elsewhere.

The package routes all of these through the same entry point —
`true_dist = "User"` plus a `g_fn` callback — under two conventions for
what the callback returns. Convention A is the recommended path: the
callback returns standardized residuals $z_j$ with mean 0 and variance
1, and the package applies $\tau + \sigma_\tau \cdot z_j$ to produce
$\tau_j$. Convention B is the escape hatch: the callback returns
$\tau_j$ directly, and the package skips the wrapper. The split exists
because some shapes — point-mass slabs in particular — are awkward to
standardize, and a callback that builds the slab on the response scale
is cleaner than forcing a unit-variance reframe. The factorization
follows @lee2025improving.

## 2. Convention A — the standardized residual callback

Convention A is the default. The callback signature is
`function(J, ...)` and the return contract is two-line: a finite
numeric vector of length `J`, with empirical mean near 0 and variance
near 1. The package applies the location-scale wrapper

$$
\tau_j \;=\; \tau \;+\; \sigma_\tau \cdot z_j,
$$

so the user's $\sigma_\tau$ scales the shape directly — without the
double-scaling pitfall the JEBS-paper-era package
[siteBayes2](m8-migration-from-siteBayes2.html) had. Layer 1 returns a
tibble with `site_index`, `z_j`, and `tau_j` columns; downstream code
reads $z_j$ off the result the same way as for any built-in shape.

The example callback below mixes two beta densities to produce an
asymmetric residual the catalog's skew-normal cannot reproduce: half
the sites draw from a $\mathrm{Beta}(2, 5)$ on $(0, 1)$ (right-skewed,
mode near 0.2), the other half draw the negative of a
$\mathrm{Beta}(5, 2)$. The centering and SD-scaling step at the end is
the user's responsibility under Convention A.

```{r conv-a-callback}
beta_diff_g <- function(J, ...) {
  flip <- stats::rbinom(J, 1, 0.5)
  raw  <- ifelse(
    flip == 1,
    stats::rbeta(J, 2, 5),
    -stats::rbeta(J, 5, 2)
  )
  (raw - mean(raw)) / stats::sd(raw)
}
```

The Layer-1 generator
[`gen_effects_user()`](../reference/gen_effects_user.html) accepts the
callback and returns the standard tibble. With `tau = 0.30` and
`sigma_tau = 0.20`, the wrapper math is the same as for the catalog
shapes; the realized marginal moments match the design inputs exactly
within Layer 1.

```{r conv-a-genuser}
set.seed(1L)
eff_a <- gen_effects_user(
  J         = 200L,
  tau       = 0.30,
  sigma_tau = 0.20,
  g_fn      = beta_diff_g,
  g_returns = "standardized",
  audit_g   = TRUE
)
print(eff_a, n = 4)
data.frame(
  mean_z_j   = mean(eff_a$z_j),
  var_z_j    = stats::var(eff_a$z_j),
  mean_tau_j = mean(eff_a$tau_j),
  sd_tau_j   = stats::sd(eff_a$tau_j)
)
```

Three things to read off. First, the `z_j` column has mean 0 and
variance 1 exactly — the callback's empirical standardization
(`(x - mean(x)) / sd(x)`) forces equality on the realized sample.
Second, `tau_j = tau + sigma_tau * z_j` holds row-by-row, so the
response-scale mean is exactly $\tau = 0.30$ and the SD is exactly
$\sigma_\tau = 0.20$. Third, the audit pass (Section 4) is silent —
what you want on a callback that meets the contract.

The same callback drops into a full design.
[`multisitedgp_design()`](../reference/multisitedgp_design.html) accepts
`true_dist = "User"` plus `g_fn`, and
[`sim_multisite()`](../reference/sim_multisite.html) produces the
multisite tibble with site sizes, sampling variances, and diagnostics.

```{r conv-a-design}
design_a <- multisitedgp_design(
  J         = 60L,
  true_dist = "User",
  g_fn      = beta_diff_g,
  sigma_tau = 0.20
)
dat_a <- sim_multisite(design_a, seed = 1L)
print(dat_a, n = 4)
data.frame(
  realized_sigma_tau = stats::sd(dat_a$tau_j),
  informativeness    = informativeness(dat_a),
  feasibility        = feasibility_index(dat_a, warn = FALSE)
)
```

The diagnostic block prints
`sigma_tau: target=0.200, realized=0.200` — the user-callback path
preserves the same target-vs-realized discipline as every catalog
shape, because Convention A's contract is what makes that comparison
meaningful. The custom shape is fully threaded through the downstream
diagnostic surface.

## 3. Convention B — the raw final effects callback

Convention B is the escape hatch. The callback returns response-scale
$\tau_j$ directly, and the package skips the location-scale wrapper.
Reach for it when standardization is awkward — the canonical case is a
point-mass-plus-slab where the user wants to keep the slab on the
response scale, rather than algebraically re-deriving the slab
parameters under the unit-variance reframe.

Under Convention B the package back-derives `z_j` for provenance using
the inverse wrapper

$$
z_j \;=\; \frac{\tau_j - \tau - X_j \boldsymbol\beta}{\sigma_\tau},
$$

so downstream Layer 2 / 3 / 4 code still has a `z_j` column. The
back-derived $z_j$ is no longer guaranteed unit-variance — the
empirical variance reflects whatever the callback produced — and the
audit pass is a no-op under `g_returns = "raw"`.

The example callback below builds a sparse-effects shape: 40% of sites
are exactly null ($\tau_j = 0$), the rest draw from a Gaussian slab
with response-scale SD 0.30. This is a point-mass-plus-slab in spirit,
but `slab_sd` is the response-scale SD the user thinks in.

```{r conv-b-callback}
sparse_g <- function(J, pi0 = 0.4, slab_sd = 0.30, ...) {
  null_mask <- stats::rbinom(J, 1, pi0)
  ifelse(
    null_mask == 1,
    0,
    stats::rnorm(J, mean = 0, sd = slab_sd)
  )
}
```

The Layer-1 call passes `g_returns = "raw"` and `audit_g = FALSE`.
Slab parameters travel through `g_args`. `tau` and `sigma_tau` are
still required — Convention B uses them only for back-derived `z_j`,
but the back-derivation still needs them.

```{r conv-b-genuser}
set.seed(1L)
eff_b <- gen_effects_user(
  J         = 200L,
  tau       = 0.30,
  sigma_tau = 0.20,
  g_fn      = sparse_g,
  g_returns = "raw",
  g_args    = list(pi0 = 0.4, slab_sd = 0.30),
  audit_g   = FALSE
)
print(eff_b, n = 4)
data.frame(
  fraction_exact_zero = mean(eff_b$tau_j == 0),
  mean_tau_j          = mean(eff_b$tau_j),
  sd_tau_j            = stats::sd(eff_b$tau_j),
  var_z_j_back        = stats::var(eff_b$z_j)
)
```

Three differences from Convention A. First, about 40% of the `tau_j`
entries are exactly 0 (realized 0.415 — the spike portion of the
spike-and-slab). Second, the response-scale SD is whatever the
callback produced (around 0.234), not the design's `sigma_tau = 0.20`;
Convention B does not enforce a target-vs-realized match on the SD.
Third, back-derived `z_j` does not have unit variance — the value
above 1 reflects the un-standardized slab plus spike, divided by
`sigma_tau`. The trade-off: the callback owns the marginal, and the
package's standardization guarantee no longer applies.

Reach for Convention B when the response-scale interpretation matters
more than the unit-variance one — sparse slabs, finite mixtures with
response-scale component means, calibrated empirical distributions
where the moments are the deliverable. The catalog-side analog is
[`gen_effects_pmslab()`](../reference/gen_effects_pmslab.html), which
applies an internal $1 / (1 - \pi_0)$ rescaling to keep the residual
unit-variance — clean for the diagnostic block, less natural when the
user wants the slab on the response scale.

## 4. `audit_g` — what it actually checks

With `audit_g = TRUE` (default), the package draws an extra
$10^5$-site sample from the callback in a side RNG stream that does
not disturb the caller's seed, computes the empirical mean and
variance, and warns if either deviates from the contract by more than
0.1 in absolute terms. The threshold is loose by design — at $10^5$
samples sampling fluctuation around the mean is $\sim 0.003$, so a
0.1 deviation is several hundred SEs and only real violations fire.
The audit does not abort, does not modify the draw, and is a no-op
under Convention B.

A violating callback triggers a structured warning with three lines:
a summary, the realized audit moments, and a fix hint. The example
below uses a callback that returns $\mathcal{N}(0.5, 1.5^2)$ draws —
wrong on both moments.

```{r audit-warns, warning = TRUE}
bad_g <- function(J, ...) stats::rnorm(J, mean = 0.5, sd = 1.5)

set.seed(1L)
out_bad <- gen_effects_user(
  J       = 100L,
  g_fn    = bad_g,
  audit_g = TRUE
)
```

The warning reports the violation summary, the realized audit moments
(here roughly 0.50 and 2.27), and the fix hint
`Return mean-zero, unit-variance draws or use g_returns = "raw"`. The
function still returns the tibble — the audit reports, the run
proceeds.

To inspect the audit moments programmatically without relying on the
warning channel, the manual version is straightforward: run the
callback at $10^5$, compute the moments, confirm finiteness and the
length contract. The example below is equivalent to what
`audit_g = TRUE` does internally.

```{r audit-manual}
manual_audit_g <- function(g_fn, J = 1e5L, ...) {
  z <- g_fn(J, ...)
  data.frame(
    n_returned   = length(z),
    length_match = length(z) == J,
    all_finite   = all(is.finite(z)),
    audit_mean   = mean(z),
    audit_var    = stats::var(z),
    contract_ok  = abs(mean(z)) < 0.1 && abs(stats::var(z) - 1) < 0.1
  )
}
set.seed(1L)
manual_audit_g(beta_diff_g)
manual_audit_g(bad_g)
```

The first row reports `contract_ok = TRUE` for `beta_diff_g`: mean
and variance both inside the 0.1 tolerance. The second row reports
`contract_ok = FALSE` for `bad_g`: variance over 2, mean near 0.5,
both far outside tolerance. The internal audit call uses
`withr::with_preserve_seed()`, so `audit_g = TRUE` and `audit_g = FALSE`
produce the same `z_j` draw — the package determinism smoke check
(hash `c52e75f276d82836` for
`sim_multisite(preset_jebs_paper(), seed = 1L)`) is independent of
whether the audit runs.

## 5. The DPM bridge

The catalog's eighth shape, `true_dist = "DPM"`, is a deferred
placeholder. Calling it without a `g_fn` raises an actionable error.

```{r dpm-error, error = TRUE}
gen_effects(J = 50L, true_dist = "DPM")
```

The error message names the canonical fix on the third line: use
`true_dist = "User"` with a custom `g_fn`, or pass an explicit DPM
bridge through `gen_effects_dpm()`. The package never silently falls
back to Gaussian.

When you actually need DPM-distributed latent effects, the bridge
pattern is short. Fit a DPM offline (with `dirichletprocess`, or a
JAGS / Stan posterior), draw a sample of standardized residuals, wrap
them in a Convention A callback. The fit is the expense — minutes to
hours depending on $J$, base measure, and concentration — and is not
amortized inside `multisiteDGP`. The illustration below uses a
deterministic three-atom mixture in place of a real DPM fit. For a
real fit, replace the closure body with a `sample()` over the DPM's
posterior atoms weighted by cluster weights, then standardize.

```{r dpm-fakefit}
fake_dpm_atoms   <- c(-1.5, 0, 1.2)
fake_dpm_weights <- c(0.3, 0.4, 0.3)
fake_dpm_atom_sd <- 0.30

fake_dpm_g <- function(J, atoms = fake_dpm_atoms,
                       weights = fake_dpm_weights,
                       atom_sd = fake_dpm_atom_sd, ...) {
  idx <- sample(seq_along(atoms), J, replace = TRUE, prob = weights)
  raw <- stats::rnorm(J, mean = atoms[idx], sd = atom_sd)
  (raw - mean(raw)) / stats::sd(raw)
}
```

[`gen_effects_dpm()`](../reference/gen_effects_dpm.html) accepts the
bridge through `g_fn` and routes internally to `gen_effects_user()`.
Convention A / B selection, the audit, and back-derivation are
identical — the function exists to give the bridge a named entry
point that matches the `true_dist` string.

```{r dpm-bridge}
set.seed(1L)
bridged <- gen_effects_dpm(
  J         = 200L,
  tau       = 0.30,
  sigma_tau = 0.20,
  g_fn      = fake_dpm_g,
  audit_g   = TRUE
)
print(bridged, n = 4)
data.frame(
  mean_z_j   = mean(bridged$z_j),
  var_z_j    = stats::var(bridged$z_j),
  mean_tau_j = mean(bridged$tau_j),
  sd_tau_j   = stats::sd(bridged$tau_j)
)
```

The bridged residual hits the unit-variance contract exactly, the
response-scale moments match the design inputs, and the audit is
silent. From the pipeline's perspective this reads as a catalog
shape — diagnostic block, precision marginal, and adapters all consume
the bridged tibble the same way as a Gaussian draw.

The bridge has two costs beyond compute. First, it does not
introspect the DPM's structure: the diagnostic block reads moments
and order statistics off the residual draw, not posterior cluster
weights. If your inference target is the DPM's cluster structure
rather than the realized $G$ it implies, fit the DPM in its native
package and use `multisiteDGP` only for the simulation-design wrapper.
Second, the empirical standardization in the closure forces the
realized draw to unit-variance exactly but does not guarantee the
population variance of the underlying DPM is 1; for the
simulation-design surface this distinction is mostly cosmetic.

## 6. Two contrasting plots

The convention contrast is easiest to read off the residual densities.
Convention A's standardized residual has a clean unit-variance
interpretation that compares directly with the catalog shapes;
Convention B's response-scale draw lives on the user's
parameterization, with a spike-and-slab signature the catalog's
standardized parameterization would have to algebraically reframe.

```{r plot-conv-a, fig.height = 4, fig.cap = "Convention A residual density (blue) over a Gaussian reference (grey dashed). The custom shape is bimodal — the two beta components produce peaks near z = -1 and z = +1 — with thinner tails than Gaussian. Key read: the residual is unit-variance by construction, so the package's wrapper applied tau + sigma_tau * z_j to produce tau_j the same way it does for any catalog shape.", fig.alt = "Density plot showing a bimodal blue density on a standardized-residual axis, overlaid against a single-peak grey Gaussian reference."}
df_a <- tibble::tibble(z = eff_a$z_j)
df_ref <- tibble::tibble(
  z = seq(-4, 4, length.out = 400),
  d = stats::dnorm(seq(-4, 4, length.out = 400))
)
ggplot(df_a, aes(x = z)) +
  geom_density(fill = "#1B4965", colour = "#1B4965", alpha = 0.40) +
  geom_line(data = df_ref, aes(x = z, y = d),
            linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  labs(
    x        = expression(z[j]~"  (standardized residual)"),
    y        = "density",
    title    = "Convention A — standardized residual from beta_diff_g",
    subtitle = "User callback returns z_j with mean 0 and variance 1; package owns the wrapper."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r plot-conv-b, fig.height = 4, fig.cap = "Convention B response-scale tau_j histogram from sparse_g. The exact-zero spike captures about 40 percent of sites (the pi0 weight); the slab portion is Gaussian with SD 0.30 around zero. Dashed lines mark tau and zero. Key read: the callback owns the marginal — the wrapper is bypassed, the spike sits exactly at zero rather than at tau, and the cross-site SD is whatever the slab plus spike produce, not the design's sigma_tau.", fig.alt = "Histogram of response-scale tau_j values with a tall narrow bar at zero (the spike) and a broader bell-shaped slab portion."}
df_b <- tibble::tibble(tau_j = eff_b$tau_j)
ggplot(df_b, aes(x = tau_j)) +
  geom_histogram(bins = 35, fill = "#62929E",
                 colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0,    linetype = "dashed", colour = "grey20") +
  geom_vline(xintercept = 0.30, linetype = "dashed", colour = "#1B4965") +
  labs(
    x        = expression(tau[j]~"  (response-scale effect)"),
    y        = "count",
    title    = "Convention B — raw response-scale draw from sparse_g",
    subtitle = "User callback returns tau_j directly; spike at zero, slab around tau = 0.30."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

The two plots carry the convention contrast. Convention A lives on
the standardized-residual axis, where every shape — catalog or
callback — is comparable by construction. Convention B lives on the
response scale, where the spike at zero is the substantive feature
and the marginal is owned by the callback.

## 7. Where to next

If, while writing your callback, the shape starts looking like a
catalog entry, prefer the catalog. The eight built-in shapes have
closed-form unit-variance derivations (where possible) and are
exercised at $J = 10^6$ in the test suite — realized variance lands
within $10^{-3}$ of 1. A custom callback only matches that guarantee
if the standardization step does, and the catalog's closed-form path
is one less moving part to audit. Substantive triggers for the
catalog side:

- Symmetric heavy tails — Student-$t$
  ([`gen_effects_studentt()`](../reference/gen_effects_studentt.html));
  the closed-form rescaler is exact.
- One-sided skew — skew-normal
  ([`gen_effects_skewn()`](../reference/gen_effects_skewn.html)) with
  the Azzalini slant [@azzalini1985class].
- Skew with a sharper peak — asymmetric Laplace
  ([`gen_effects_ald()`](../reference/gen_effects_ald.html)) with the
  Yu-Zhang quantile $\rho$ [@yu2005three].
- Two latent classes — two-component mixture
  ([`gen_effects_mixture()`](../reference/gen_effects_mixture.html)).
- A fraction of sites exactly null — point-mass-plus-slab
  ([`gen_effects_pmslab()`](../reference/gen_effects_pmslab.html)),
  whose $1 / (1 - \pi_0)$ rescaling keeps the residual unit-variance
  and the diagnostic block clean.

The user-callback path is the right tool when none of these match.
Three motifs commonly trigger it: horseshoe-style residuals (sharp
peak with very heavy tails — neither Student-$t$ nor ALD captures
both), finite mixtures with three or more components, and empirical
distributions extracted from a sister study by deconvolution or DPM
posterior draws [@walters2024empirical; @lee2025improving]. For the
first two, Convention A is the path; for the third, the DPM bridge
from Section 5. Sibling vignettes for deeper context:
[M1: The two-stage DGP](m1-statistical-dgp.html) (formal
specification of the four generative layers — the callback operates
inside Layer 1),
[M2: G-distribution catalog and standardization](m2-g-distribution-catalog.html)
(the eight built-in shapes and the unit-variance discipline the
callback inherits),
[A8: Cookbook](a8-cookbook.html) (short walkthroughs that mix custom
callbacks with the rest of the design surface), and
[A5: Calibrating to real data](a5-calibrating-real-data.html) (when
calibration against an observed precision distribution suggests a
shape outside the catalog).

## References

The factorization $\tau_j = \tau + \sigma_\tau z_j$ with a
unit-variance standardized residual and the empirical-Bayes context
for site-effect estimation are from @lee2025improving and
@walters2024empirical. The catalog parameterizations cross-referenced
in the closing pointer are @azzalini1985class for the skew-normal
slant and @yu2005three for the asymmetric Laplace quantile.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
