---
title: "M6 · Adapters and downstream packages"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For analysts who simulate with multisiteDGP and then fit, plot, or
  power-analyze with a downstream package — `metafor`, `baggr`, or
  `multisitepower`. Adapters strip the package's S3 class and rename
  the canonical effect, SE, and site columns to the target ecosystem's
  convention; numerical values are not transformed. You leave with the
  formal contract, the three built-in rename maps, the round-trip
  invariant, the reserved-name guard, and acceptance criteria for
  adding a fourth adapter.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M6 Adapters and downstream packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. The adapter contract

A `multisitedgp_data` object carries seven canonical columns
(`site_index`, `z_j`, `tau_j`, `tau_j_hat`, `se_j`, `se2_j`, `n_j`),
an S3 class, and design / provenance attributes. None of the names
match the conventions analysis packages fit with: `metafor::rma()`
reads `yi`, `vi`; `baggr::baggr()` reads `tau`, `se`; `multisitepower`
reads `site`, `estimate`, `se`, `n`.

The adapter contract is two clauses, both load-bearing. First, an
adapter strips the `multisitedgp_data` class so the result is a plain
`tbl_df` and the downstream package's S3 dispatch is not surprised.
Second, an adapter renames the canonical columns to the target
ecosystem's convention but does **not** alter numerical values: every
renamed column is bit-for-bit identical to its source. The
factorization $y_j = \tau_j + e_j$,
$e_j \sim \mathcal{N}(0, \sigma_j^2)$ [@lee2025improving] is preserved
on disk. Two corollaries follow: the round-trip invariant (Section 3)
holds by construction; reserved-name collisions are unrecoverable —
the adapter errors rather than overwrite a user covariate named `yi`.

## 2. The three built-in adapters

The package ships three adapters, each named for its target. All
three follow the same pattern: a generic; a `.multisitedgp_data`
method that runs the soft-dep guard, the reserved-name check, and the
rename; a `.default` method that errors with the class of the
offending input. Implementations live in `R/adapters.R` lines 80-218.

A 50-site simulation from `preset_education_modest()` is the shared
input for all three sections below.

```{r data}
dat <- sim_multisite(preset_education_modest(), seed = 1L)
print(dat, n = 4)
```

### 2.1  `as_metafor()` — meta-analysis fitting

Use `as_metafor()` when the next step is `metafor::rma()`,
`metafor::escalc()`, or any of the `metafor` plotting / influence
helpers. The rename map is

| canonical column   | metafor column |
|--------------------|----------------|
| `tau_j_hat`        | `yi`           |
| `se2_j`            | `vi`           |
| `se_j`             | `sei`          |

Three-column output: effect, sampling variance, standard error.

```{r as-metafor}
mf <- as_metafor(dat)
print(mf, n = 4)
class(mf)
```

The class line confirms the strip — `mf` is a plain `tbl_df`. Pass
to `metafor::rma()` directly:

```{r metafor-rma, eval = requireNamespace("metafor", quietly = TRUE)}
metafor::rma(yi = yi, vi = vi, data = mf, method = "REML")
```

The empirical-Bayes shrinkage `metafor::rma()` produces takes the
adapted `yi` / `vi` as inputs — the precision-driven posterior is
the analysis target [@walters2024empirical].

### 2.2  `as_baggr()` — Bayesian aggregation with optional truth

Use `as_baggr()` when the next step is `baggr::baggr()` or its
diagnostic plots. The rename map is two columns by default and three
with `include_truth = TRUE`:

| canonical column   | baggr column | when included     |
|--------------------|--------------|-------------------|
| `tau_j_hat`        | `tau`        | always            |
| `se_j`             | `se`         | always            |
| `tau_j`            | `tau_true`   | `include_truth`   |

`include_truth = TRUE` exposes the latent true effects $\tau_j$ as a
separate column. `baggr` itself does not consume `tau_true` — the
column is for diagnostic comparison of `baggr`'s posterior against the
simulation truth, e.g., posterior-mean coverage of true effects. It
is never an input to fitting.

```{r as-baggr}
bg <- as_baggr(dat)
print(bg, n = 4)

bg_truth <- as_baggr(dat, include_truth = TRUE)
print(bg_truth, n = 4)
```

When the simulation is the inferential target — sampling-design
research, calibration of an applied trial — set `include_truth = TRUE`.
When the simulation stands in for an opaque real dataset, the default
`include_truth = FALSE` matches what `baggr` would see in production.

### 2.3  `as_multisitepower()` — empirical-Bayes power

Use `as_multisitepower()` when the next step is the empirical-Bayes
power surface in the `multisitepower` package. The rename map is

| canonical column   | multisitepower column | when included           |
|--------------------|-----------------------|-------------------------|
| `site_index`       | `site`                | always                  |
| `tau_j_hat`        | `estimate`            | always                  |
| `se_j`             | `se`                  | always                  |
| `n_j`              | `n`                   | when any `n_j` non-`NA` |

The `n` column is conditional. Direct-precision designs leave `n_j`
all-`NA` because per-site sample size is not part of the precision
specification; the adapter omits `n` rather than emit all-`NA`.
Site-size paradigm designs (the default in
`preset_education_modest()`) include `n`.

`multisitepower` is a soft `Suggests` dependency and may not be
installed on a given machine. The chunk below is guarded; when the
package is absent, the adapter raises a classed soft-dependency error
that the chunk traps and prints, exercising the guard from Section 5.

```{r as-multisitepower}
if (requireNamespace("multisitepower", quietly = TRUE)) {
  mp <- as_multisitepower(dat)
  print(mp, n = 4)
} else {
  err <- tryCatch(as_multisitepower(dat), error = conditionMessage)
  cat(err, sep = "\n")
}
```

When `multisitepower` is installed, the four-column tibble feeds
straight into the package's power surface.

## 3. Round-trip invariant

The adapter contract's "no numerical change" clause is the round-trip
invariant. Every renamed column is bit-for-bit identical to the
canonical column it came from — `identical()` returns `TRUE`, not just
`all.equal()` within tolerance. The check is what makes the adapter a
relabeling rather than a transformation.

```{r round-trip-identity}
data.frame(
  metafor_yi_eq_tau_j_hat   = identical(mf$yi,        dat$tau_j_hat),
  metafor_vi_eq_se2_j       = identical(mf$vi,        dat$se2_j),
  metafor_sei_eq_se_j       = identical(mf$sei,       dat$se_j),
  baggr_tau_eq_tau_j_hat    = identical(bg$tau,       dat$tau_j_hat),
  baggr_se_eq_se_j          = identical(bg$se,        dat$se_j),
  baggr_tau_true_eq_tau_j   = identical(bg_truth$tau_true, dat$tau_j)
)
```

All six entries return `TRUE` — stronger than `all.equal()` tolerance,
because the adapter is an attribute-strip plus a rename, not a
transform, so floating-point representation is unchanged.

The figure below illustrates the invariant on the `metafor` rename:
each `(yi, tau_j_hat)` pair lands exactly on the unit line, and the
`(vi, se2_j)` and `(sei, se_j)` pairs do the same.

```{r plot-round-trip, fig.height = 4.5, fig.cap = "Round-trip identity for as_metafor on a 50-site simulation. Each panel plots a metafor column against its canonical multisitedgp source; all 50 points lie exactly on the unit line, confirming the adapter is a relabeling. Key read: read off any horizontal deviation from the diagonal as a contract violation — there should be none.", fig.alt = "Three side-by-side scatter plots of metafor columns yi, vi, sei against canonical multisitedgp columns tau_j_hat, se2_j, se_j; in each panel all points fall on the y = x diagonal."}
rt <- tibble::tibble(
  rename     = rep(c("yi vs tau_j_hat", "vi vs se2_j", "sei vs se_j"), each = 50),
  canonical  = c(dat$tau_j_hat, dat$se2_j, dat$se_j),
  adapted    = c(mf$yi,         mf$vi,     mf$sei)
)
ggplot(rt, aes(x = canonical, y = adapted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linewidth = 0.5) +
  geom_point(colour = "#1B4965", alpha = 0.85, size = 1.6) +
  facet_wrap(~ rename, scales = "free", nrow = 1) +
  labs(
    x        = "canonical multisitedgp column",
    y        = "as_metafor() column",
    title    = "Round-trip identity for as_metafor()",
    subtitle = "Every point on the y = x line; the rename preserves values exactly."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

The same picture holds for `as_baggr()` and `as_multisitepower()` —
the rename map differs, but the invariant does not.

## 4. Reserved-name protection

Each adapter reserves the column names it produces. If a user-supplied
design carries a covariate whose name collides with a reserved name,
the adapter aborts before the rename — silently overwriting the
covariate would corrupt the analysis. The reserved sets are
small enough to fit in a table:

| adapter             | reserved input names                              |
|---------------------|---------------------------------------------------|
| `as_metafor()`      | `yi`, `vi`, `sei`                                 |
| `as_baggr()`        | `tau`, `se` (plus `tau_true` if `include_truth`)  |
| `as_multisitepower()` | `site`, `estimate`, `se`, `n`                   |

The check fires only on user-supplied covariate columns, not on the
seven canonical columns the adapter is renaming from. The check
constants live as `.RESERVED_METAFOR`, `.RESERVED_BAGGR`, and
`.RESERVED_MULTISITEPOWER` at the top of `R/adapters.R`.

A demonstration: build a 20-site design that carries a covariate
named `yi`, the metafor effect column. The adapter aborts with a
three-line message — the violation, the offending reserved set, and
the fix.

```{r reserved-collision-metafor}
covariates <- data.frame(yi = as.numeric(scale(seq_len(20L))))
collision <- sim_multisite(
  multisitedgp_design(
    J        = 20L,
    formula  = ~ yi,
    beta     = 0.10,
    data     = covariates
  ),
  seed = 1L
)
err_mf <- tryCatch(as_metafor(collision), error = conditionMessage)
cat(err_mf, sep = "\n")
```

The same machinery fires for `as_baggr()` on a covariate named `tau`:

```{r reserved-collision-baggr}
covariates2 <- data.frame(tau = stats::rnorm(20L))
collision2 <- sim_multisite(
  multisitedgp_design(
    J        = 20L,
    formula  = ~ tau,
    beta     = 0.10,
    data     = covariates2
  ),
  seed = 1L
)
err_bg <- tryCatch(as_baggr(collision2), error = conditionMessage)
cat(err_bg, sep = "\n")
```

The fix the message points to: rename the covariate before adapting,
or use `tibble::as_tibble()` for a plain tibble without renames.

A non-colliding covariate threads through cleanly. Below we add two —
`school_type` (factor) and `urbanicity` (numeric) — and confirm both
land in the adapted tibble alongside the renamed canonical columns.

```{r covariate-passthrough}
cv <- data.frame(
  school_type = factor(rep(c("A", "B"), 25L)),
  urbanicity  = stats::rnorm(50L)
)
dat_cv <- sim_multisite(
  multisitedgp_design(
    J        = 50L,
    formula  = ~ school_type + urbanicity,
    beta     = c(0.05, 0.02),
    data     = cv
  ),
  seed = 1L
)
mf_cv <- as_metafor(dat_cv)
print(mf_cv, n = 4)
```

The output has five columns: the three-column metafor rename, then
the two pass-through covariates in original order. Internally,
`setdiff(names(x), .canonical_data_columns())` selects the carry-through
set (`R/adapters.R` line 167) — any non-canonical column qualifies.

## 5. Soft-dependency guard

`metafor`, `baggr`, and `multisitepower` are listed in `Suggests`,
not `Imports` — none is required to use `multisiteDGP` itself, only
to use the matching adapter. Each adapter calls
`requireNamespace("target", quietly = TRUE)` through the internal
`.require_soft_dependency()` helper and aborts with a three-line
error when the target package is missing:

| line  | content                                                   |
|-------|-----------------------------------------------------------|
| 1     | `Package "<pkg>" is required for "<adapter>()".`          |
| 2     | `<pkg> is a multisiteDGP Suggests dependency …`           |
| 3     | `Use install.packages("<pkg>") and try again.`            |

The exact message body for `multisitepower` (commonly absent on a
fresh CRAN check environment) is the chunk output of Section 2.3. The
guard fires *before* the reserved-name check.

The `.default` method of each adapter handles the wrong-class case —
calling `as_metafor()` on a list, a data frame, or any object that is
not a `multisitedgp_data` aborts with a parallel three-line message.

```{r default-dispatch}
err_default <- tryCatch(
  as_metafor(list(yi = 1, vi = 1)),
  error = conditionMessage
)
cat(err_default, sep = "\n")
```

The first line names the adapter and the violated input class, the
second prints the actual class of the offending object, and the third
gives the canonical fix — call `sim_multisite()` or `sim_meta()` first.

## 6. Acceptance criteria for a new adapter

The package ships three adapters because three downstream ecosystems
are the common applied case. A fourth adapter — say `as_metaSEM()` or
`as_robumeta()` — is straightforward to add, but only if it satisfies
the six checks below. The list is the maintainer's checklist; it
doubles as the test fixture the new adapter's PR should ship with.

**(C1) Input contract.** The adapter is an S3 generic with a
`.multisitedgp_data` method and a `.default` method. The
`.multisitedgp_data` method takes a `multisitedgp_data` object; the
`.default` method errors with the offending object's class. No
adapter consumes a raw `data.frame` or a list — the input is always
the simulator's output.

**(C2) Output rename map.** The adapter's docstring includes the
canonical-to-target rename map as a three-column or four-column
table, and the same map appears as a unit-test fixture asserting
`identical(out$<target_col>, x$<canonical_col>)` for each row of the
map. Conditional columns (e.g., `n` only when `n_j` non-`NA`) are
documented as such; the unit test exercises both the present and
absent branches. The output is a plain `tbl_df` — no
`multisitedgp_data` class survives.

**(C3) Soft-dependency guard.** The target package is added to
`DESCRIPTION` `Suggests`, never `Imports`. The adapter's first
substantive line is a `.require_soft_dependency("<pkg>", "<adapter>")`
call. The guard's error path is exercised in the test suite by a
`mockery::stub()` that returns `FALSE` from `requireNamespace()`, or
by skipping the adapter test when the package is unavailable
(`testthat::skip_if_not_installed()`).

**(C4) Reserved-name protection.** Every column the adapter's rename
map produces appears in a top-level `.RESERVED_<adapter>` constant.
The `.multisitedgp_data` method calls
`.check_adapter_reserved_names(x, reserved = .RESERVED_<adapter>, adapter = "<name>")`
*after* the soft-dep guard and *before* the rename. The unit test
confirms the abort fires with the reserved-set listed in the message.

**(C5) Round-trip invariant.** A unit test confirms `identical()`
between every renamed column and its canonical source on a fixture
simulation. Floating-point tolerance (`all.equal()`) is not adequate
— the contract is bit-for-bit identity, so the test must be the
strict `identical()` form. A `testthat::expect_identical()` call per
column is the conventional shape.

**(C6) Covariate pass-through.** A unit test threads a covariate
through `sim_multisite()` (or `sim_meta()`) and confirms it appears
in the adapted tibble after the renamed canonical columns, in
original order. The pass-through is automatic via
`.bind_adapter_covariates()` if (C1)-(C5) are satisfied; the test
just confirms the helper still binds them.

A fourth adapter shipping all six fixtures is what makes the addition
non-disruptive. The internal helper machinery
(`.require_soft_dependency()`, `.check_adapter_reserved_names()`,
`.bind_adapter_covariates()`) is generic; the rename body is rarely
longer than the 7-line `.as_metafor_rename()` function.

## 7. Where to next

Adapters are the boundary between simulation design and downstream
analysis — a contract that lets the analysis script read like an
analysis script (`metafor::rma(yi = yi, vi = vi)`) instead of like a
data-rewiring script. The contract is small enough to memorize: strip
the class, rename to the target convention, do not touch the values.
The round trip is bit-for-bit; the reserved-name check is the cost of
the no-overwrite guarantee; the soft-dep guard is the cost of keeping
`metafor` / `baggr` / `multisitepower` out of `Imports`.

- [A1 · Getting started](a1-getting-started.html) — first encounter
  with the `as_metafor()` adapter end-to-end.
- [A6 · Case study — multisite trial](a6-case-study-multisite.html)
  and [A7 · Case study — meta-analysis](a7-case-study-meta-analysis.html)
  — full analyses that consume the adapters.
- [M1 · The two-stage DGP](m1-statistical-dgp.html) — the
  canonical-column factorization the rename relabels.
- [M3 · Margin and SE models](m3-margin-se-models.html) — paradigm
  context for the front doors that adapters consume.
- [M7 · Reproducibility and provenance](m7-reproducibility-provenance.html)
  — hash invariance under adapter calls.
- Reference pages:
  [`as_metafor()`](../reference/as_metafor.html),
  [`as_baggr()`](../reference/as_baggr.html),
  [`as_multisitepower()`](../reference/as_multisitepower.html).

## References

The factorization $y_j = \tau_j + e_j$ with
$e_j \sim \mathcal{N}(0, \sigma_j^2)$ that the canonical columns
encode is from @lee2025improving. The empirical-Bayes shrinkage
target that `metafor::rma()` produces under the adapted `(yi, vi)`
inputs is the analysis goal in @walters2024empirical.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
