---
title: "M7 · Reproducibility and provenance"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  A practical guide to multisiteDGP's reproducibility contract. It shows how
  to run a seeded simulation, record producer and verifier provenance, save a
  manifest, and verify the schema-v4 canonical numerical hash without
  confusing numerical equivalence with byte-for-byte identity.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M7 Reproducibility and provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  echo = TRUE,
  results = "markup",
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi = 144,
  fig.align = "center",
  out.width = "100%"
)
library(multisiteDGP)
```

## 1. The contract in one paragraph

For the same validated design and the same non-`NULL` seed, the package
pins the RNG triple to `Mersenne-Twister` / `Inversion` / `Rejection`.
It restores the caller's RNG kind and `.Random.seed` exactly when the call
ends. Supported platforms should therefore reproduce the same
**schema-v4 canonical numerical content** and the same
`canonical_hash()`.

The deliberately narrower phrase *canonical numerical content* matters.
`canonical_hash()` rounds doubles to nine significant digits and ignores
attributes outside its payload. Equal hashes do not prove that the raw R
objects or serialized `.rds` files are byte-identical. Exact artifact
identity belongs in a separate SHA-256.

If `seed = NULL`, the package uses the caller's active RNG kind and state.
That is an intentional externally managed stream, not a design-only replay
contract.

## 2. What is recorded

The reproducibility surface has three parts.

1. `canonical_hash(x)` identifies the canonical payload under a named hash
   schema.
2. `provenance_string(x)` reports the stored producer alongside the current
   verifier.
3. A project manifest connects that evidence to a figure, table, or other
   artifact.

The default data hash covers the simulated columns, column names, factor
labels and ordering, paradigm, design hash, schema, and custom-hook names.
It excludes package version and derived diagnostics. Package version remains
in provenance. Diagnostics are functions of the data and can also introduce
platform-sensitive accumulation noise, but selected values can be opted in:

```{r diagnostic-hash}
demo <- sim_multisite(J = 20L, seed = 101L)
canonical_hash(demo)
canonical_hash(demo, diagnostics_to_include = c("I_hat", "R_hat"))
```

Callback bodies and closure environments are not hashed. The payload records
that a hook such as `g_fn` was present. If different callback code changes
the simulated columns, the data hash changes because the data changed; if it
produces the same canonical columns, the hash can remain the same. Archive the
callback source or its defining package version separately.

## 3. Five-step workflow

### 3.1 Step 1 — construct a design

Use a preset or save an explicit `multisitedgp_design`. The design hash is a
quick check that collaborators have the same declared inputs.

```{r design}
design <- preset_jebs_paper()
design_hash <- canonical_hash(design)
design_hash
```

### 3.2 Step 2 — run with an explicit seed

Pass the seed to the wrapper. An ambient `set.seed()` is not a substitute
for the wrapper argument.

```{r simulate}
dat <- sim_multisite(design, seed = 1L)
dat
```

### 3.3 Step 3 — capture hash and provenance

```{r provenance}
data_hash <- canonical_hash(dat)
producer <- attr(dat, "provenance", exact = TRUE)

data_hash
provenance_string(dat)
```

The line distinguishes fields recorded when the object was produced
(`producer_R`, `producer_platform`, `producer_rng`) from the runtime
performing the current check (`verifier_R`, `verifier_platform`). It also
prints the hash schema. `provenance_string()` warns if the current R version
or platform differs from the stored producer; that warning is context for
diagnosis, not an automatic hash failure.

### 3.4 Step 4 — save a manifest

Attach the evidence to the artifact it supports. A useful minimum is:

```{r manifest}
manifest <- tibble::tibble(
  artifact = "fig3_jebs_paper_seed1",
  package_version = producer$multisitedgp_version,
  preset = "preset_jebs_paper",
  seed = producer$seed,
  rng_kind = paste(producer$rng_kind, collapse = "/"),
  rng_policy = producer$rng_policy,
  paradigm = producer$paradigm,
  hash_schema = producer$hash_schema_version,
  canonical_hash = data_hash,
  design_hash = producer$design_hash,
  producer_R = producer$R_version,
  producer_platform = producer$platform
)
manifest
```

When a file is checked in, add its SHA-256 and the generator script SHA-256.
The canonical hash verifies numerical content; the file SHA verifies the exact
artifact in the repository.

### 3.5 Step 5 — verify a rerun

```{r verify}
dat_rerun <- sim_multisite(design, seed = manifest$seed)

verification <- tibble::tibble(
  schema_matches =
    attr(dat_rerun, "provenance", exact = TRUE)$hash_schema_version ==
    manifest$hash_schema,
  design_matches =
    attr(dat_rerun, "provenance", exact = TRUE)$design_hash ==
    manifest$design_hash,
  data_matches = canonical_hash(dat_rerun) == manifest$canonical_hash
)
verification
```

All three checks should be `TRUE`. Never compare bare hash strings from
different schemas as though they had the same meaning.

## 4. RNG isolation

A seeded wrapper call is independent of the caller's prior RNG configuration
and restores that configuration afterward. The example below runs the same
design after two different caller RNG kinds.

```{r rng-isolation}
hash_under_caller_rng <- function(kind, seed) {
  seed_object <- ".Random.seed"
  caller_kind <- RNGkind()
  caller_has_seed <- exists(seed_object, envir = .GlobalEnv, inherits = FALSE)
  caller_seed <- if (caller_has_seed) .Random.seed else NULL
  on.exit({
    do.call(RNGkind, as.list(caller_kind))
    if (caller_has_seed) {
      assign(seed_object, caller_seed, envir = .GlobalEnv)
    } else if (exists(seed_object, envir = .GlobalEnv, inherits = FALSE)) {
      rm(list = seed_object, envir = .GlobalEnv)
    }
  }, add = TRUE)

  RNGkind(kind, "Inversion", "Rejection")
  set.seed(902L)
  before_kind <- RNGkind()
  before_seed <- .Random.seed
  out <- sim_multisite(design, seed = seed)

  stopifnot(identical(RNGkind(), before_kind))
  stopifnot(identical(.Random.seed, before_seed))
  canonical_hash(out)
}

hashes <- c(
  mersenne = hash_under_caller_rng("Mersenne-Twister", 1L),
  lecuyer = hash_under_caller_rng("L'Ecuyer-CMRG", 1L)
)
hashes
length(unique(hashes)) == 1L
```

For `seed = NULL`, none of these package-pinning claims apply. Record the
caller RNG kind and enough stream state in your own workflow if an unseeded
front door is intentional.

## 5. Numerical equivalence is not raw identity

Schema v4 rounds doubles to nine significant digits. This suppresses
irrelevant sub-precision drift, but decimal rounding has a boundary.

```{r precision-boundary}
x <- data.frame(value = 0.2)
x_subprecision <- x
x_subprecision$value <- x$value + 1e-11

c(
  raw_identical = identical(x, x_subprecision),
  hash_identical =
    identical(canonical_hash(x), canonical_hash(x_subprecision))
)

boundary <- 1.234567895
below <- data.frame(value = boundary - .Machine$double.eps)
above <- data.frame(value = boundary + .Machine$double.eps)

c(
  absolute_difference = abs(below$value - above$value),
  hash_identical = identical(canonical_hash(below), canonical_hash(above))
)
```

Thus equal hashes define one explicit numerical-equivalence class, while
different hashes can also flag two very close numbers on opposite sides of a
rounding boundary. Preserve the original data when more precision matters.

## 6. Diagnosing a mismatch

Check these fields in order:

1. **Schema.** A v3 hash and v4 hash are not comparable. Recompute literals
   and manifests after an intentional schema migration.
2. **Design hash and seed.** A mismatch localizes the problem to inputs.
3. **Package version and source revision.** A code change may move data even
   when the hash schema itself is unchanged.
4. **Producer runtime and RNG policy.** Seeded v4 calls pin the RNG triple;
   unseeded calls remain caller-controlled.
5. **Custom callbacks.** Hook names do not identify callback bodies; retrieve
   the archived source.
6. **Artifact SHA.** A canonical match with a binary SHA mismatch can mean
   serialization or non-payload attributes changed.

Operating-system difference is not a blanket exemption. Supported CI platforms
are required to agree on locked v4 canonical cases. At the same time, the
package does not claim raw byte equality across platforms.

The installed policy is available at:

```{r policy-path}
system.file("REPRODUCIBILITY.md", package = "multisiteDGP")
```

## 7. Schema-v4 migration

v4 replaces contradictory v3 wording with one contract. It:

- pins the RNG triple for seeded wrappers;
- records RNG kind and policy in provenance;
- retains factor labels, levels, and orderedness in the payload;
- applies numeric canonicalization to explicitly included diagnostics;
- separates stored producer fields from current verifier fields.

Because the schema is part of the payload, all v3 hash literals and full-object
fixture provenance must be regenerated. This is expected migration work, not
evidence that the underlying data necessarily changed.

## 8. Acknowledgments and next steps

For a lighter introduction, see
[Getting started](a1-getting-started.html). For grid workflows, see the
[Cookbook](a8-cookbook.html). For the formal DGP, continue with
[The two-stage DGP](m1-statistical-dgp.html).

The package implements the multisite-trial simulation design discussed by
@lee2025improving; related empirical-Bayes context appears in
@walters2024empirical.

## References

The bibliography entries are supplied by the package vignette bibliography.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the University of
Alabama. The opinions expressed are those of the authors and do not represent
views of the Institute or the U.S. Department of Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
