---
title: "M7 · Reproducibility and provenance"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For applied researchers who have already run a simulation and now
  want to make sure the result is reproducible — by them tomorrow, by
  a coauthor next month, by a reviewer next year. This vignette walks
  through the five-step workflow `multisiteDGP` is designed around:
  set the seed, capture the provenance string, capture the canonical
  hash, save a manifest, and verify the hash on rerun. You leave with
  a one-line provenance fingerprint, a 16-character hash, a saved
  manifest tibble, and a clear sense of what the package guarantees
  and where the contract can break.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M7 Reproducibility and provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Why reproducibility matters

You ran a simulation. The numbers in your draft — the realized
between-site standard deviation, the shrinkage benefit, the funnel
plot — all came from a single call to `sim_multisite()` on a single
afternoon. Six months from now, a reviewer asks whether you can
reproduce the results, or a coauthor wants to extend them, or a
graduate student tries to swap in a new preset and discovers the
output looks different from the one in your appendix. At that point
you need to be able to say, with one line of evidence: *the
simulation that produced Figure 3 is byte-identical to this rerun*,
or, equivalently, *the contract is broken — here is exactly which
ingredient changed*. That one-line evidence is what `canonical_hash()`
and `provenance_string()` give you, and what the rest of this
vignette walks through.

## 2. The four-piece kit

`multisiteDGP` ships four pieces of machinery for the job. Two are
exported functions you call directly; one is a policy the package
enforces internally on your behalf; one is a pattern (the "manifest")
you assemble per project.

1. **`canonical_hash(x)`** — the 16-character `xxhash64` content
   fingerprint of a simulation object, a design object, or any
   data-frame-shaped result. It normalizes column order, drops row
   names, restricts to a stable diagnostics allowlist, and replaces
   callback functions with presence sentinels. Two runs with the
   same package version, seed, and design produce the same hash; any
   meaningful drift produces a different hash.
2. **`provenance_string(x)`** — a single pipe-delimited line carrying
   package version, paradigm, seed, canonical hash, design hash, hash
   algorithm, R version, and any active hooks. Drop it in the methods
   appendix or a figure caption; it is the smallest unit a reviewer
   needs to identify the run.
3. **The local-seed-stream policy** — internally the package never
   manufactures seeds from the caller's global RNG state. Whenever
   `sim_multisite()` or `sim_meta()` needs to allocate per-row seeds
   (for example, when running a `design_grid()`), it does so with
   `withr::with_seed()` inside an isolated stream so your ambient
   `set.seed()` is never disturbed. The practical consequence is that
   running unrelated code between two `sim_multisite()` calls with
   the same `seed = 1L` cannot change either result.
4. **The golden-fixture manifest** — a tibble (or row in a registry)
   recording the artifact name, package version, preset, seed,
   canonical hash, design hash, hash algorithm, R version, and
   platform for one specific simulation. Saving this alongside a
   paper draft, or committing it to your project repo, is what turns
   the hash from a computation into a contract.

The first two are the user-facing surface. The third is invisible.
The fourth is yours to assemble. Sections 3 through 5 below walk
through how the four pieces fit together in a five-step workflow.

## 3. Workflow walkthrough — pinning a single artifact

The walkthrough is structured around one running example: you have
just run a simulation with `preset_jebs_paper()` at `seed = 1L`, and
you want to pin it. The example is the same one used as the
package's smoke check, so the canonical hash you will see —
`c52e75f276d82836` — is also documented in the
`vignettes/_vignette-frontmatter-template.txt` file as the
package-level determinism check. If your local rerun produces this
exact hash, the version of the package you have installed is in
sync with the published reference.

### 3.1 Step 1 — set the seed

The seed is the entry point to determinism. The package follows the
convention used by `lee2025improving` and the JEBS preset: pass an
integer to the `seed` argument of every primary call, do not rely on
the ambient global RNG state. The setup chunk also calls
`set.seed(1L)` for any helper code outside the primary calls.

```{r seed}
set.seed(1L)
```

A common pitfall: setting the global seed but forgetting to pass
`seed = 1L` to `sim_multisite()`. The two are not interchangeable.
The argument carries the seed through the local seed stream the
package allocates internally; the global call only protects helpers
that draw from `runif()` directly.

### 3.2 Step 2 — run the simulation and capture `provenance_string()`

Run the simulation. The result is a `multisitedgp_data` object —
a tibble with simulation-specific attributes for design, paradigm,
diagnostics, and provenance.

```{r simulate}
dat <- sim_multisite(preset_jebs_paper(), seed = 1L)
dat
```

Now ask the object for its one-line provenance fingerprint.

```{r provenance}
provenance_string(dat)
```

Read the line left to right: package version, paradigm (the
*site-size-driven* path of `lee2025improving`, versus the
*direct-precision* path used in
[A7](a7-case-study-meta-analysis.html)), seed, content hash of the
realized data, fingerprint of the design alone, hash algorithm,
R version, and active callbacks (`none` because
`preset_jebs_paper()` registers no user-supplied `g_fn`, `se_fn`,
or dependence callback). This is the line you paste into the
methods appendix; it is enough for a reviewer to know what was run.

### 3.3 Step 3 — capture `canonical_hash()` separately

You do not strictly need to capture the hash separately — it appears
inside the provenance string. But the hash on its own is the contract
you will check programmatically, so it is worth pulling it into a
named variable.

```{r hash}
h <- canonical_hash(dat)
h
```

Two things are worth noting about what the hash is and is not. It is
*content-canonical*: column order, row names, and the inclusion of
non-numeric diagnostics do not affect it. The package sorts columns
before hashing and restricts diagnostics to a stable allowlist
(`I_hat`, `R_hat`, `rho_S_residual`, `rho_S_marginal`,
`rho_P_residual`, `rho_P_marginal`, `sigma_tau_resid`,
`sigma_tau_marg`). It is *callback-presence-sensitive*: if a custom
`g_fn` was registered, the hash records that a function was present
without hashing the function body or its enclosing environment.
Replacing one user callback with another that produces the same
output gives a different hash, which is the conservative behavior
for an audit trail.

A second hash — the *design hash* — fingerprints the design object
without running the simulation:

```{r design-hash}
design_hash <- canonical_hash(preset_jebs_paper())
design_hash
```

This is useful when you want to verify that two collaborators are
working from the same design specification before they bother running
the simulation. Equal `design_hash`, equal `seed`, and equal package
version is the three-way agreement that licenses an equality check on
`canonical_hash(dat)`.

### 3.4 Step 4 — save a manifest

The manifest is a one-row tibble (or one row in a per-project
registry) that records everything a future rerun would need. There
is no built-in `save_manifest()` function — the manifest is a
pattern, not an object — so the snippet below shows the canonical
shape. Saving the result as a CSV alongside your paper draft, or
committing it to your project repo, is what closes the loop.

```{r manifest}
manifest <- tibble::tibble(
  artifact       = "fig3_jebs_paper_seed1",
  pkg_version    = as.character(utils::packageVersion("multisiteDGP")),
  preset         = "preset_jebs_paper",
  seed           = 1L,
  paradigm       = attr(dat, "paradigm", exact = TRUE),
  canonical_hash = canonical_hash(dat),
  design_hash    = canonical_hash(preset_jebs_paper()),
  hash_algo      = "xxhash64",
  R_version      = paste(R.version$major, R.version$minor, sep = "."),
  platform       = R.version$platform,
  recorded_at    = format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
)
manifest
```

Pick the `artifact` slot to match the figure or table the
simulation feeds. The `pkg_version` is the most-watched field —
this is the column that, when it changes, is the most common
explanation for a hash mismatch on a rerun a year later. The
`recorded_at` timestamp is for forensic ordering only; it is not
part of the determinism contract.

In production work the manifest is typically saved as
`outputs/manifests/<artifact>.csv` and the bare hash as
`outputs/manifests/<artifact>.hash`, both committed to the project
repo. A simple convention is enough; nothing in `multisiteDGP`
prescribes a specific layout.

### 3.5 Step 5 — verify the hash on rerun

You — or your reviewer, or your future self — now want to verify
that a rerun of `sim_multisite(preset_jebs_paper(), seed = 1L)`
produces the same simulation. The check is a one-liner.

```{r verify}
dat_rerun <- sim_multisite(preset_jebs_paper(), seed = 1L)
identical(canonical_hash(dat_rerun), manifest$canonical_hash)
```

`TRUE` is the contract holding. If the comparison returns `FALSE`,
section 5 below walks through the candidate explanations in order
of likelihood (R version, package version, OS).

## 4. Hash stability under reseed — a visual check

It is worth seeing the hash behavior with your own eyes once. The
plot below runs `sim_multisite(preset_jebs_paper(), seed = s)` for
ten values of `s`, then plots the resulting `canonical_hash` as a
factor. The bar for `seed = 1L` is repeated five times under
deliberately scrambled ambient global RNG state to confirm that the
local seed stream insulates the simulation from caller state — the
five repeats produce the same hash. The other nine bars are nine
different seeds; each gives a different hash.

```{r hash-grid, fig.cap = "Each bar is one canonical hash. Five repeats of seed = 1L (scrambled global state between calls) collapse to a single hash, confirming the local-seed-stream policy. Nine different seeds give nine different hashes, confirming the seed argument is the lever.", fig.alt = "Bar chart of canonical hashes; the first five bars all show the same hash from repeated seed = 1L calls under scrambled global RNG state, while the next nine bars each show a distinct hash from a different seed value."}
hash_grid <- function() {
  out <- vector("list", 14L)
  # Five repeats of seed = 1, with global RNG state deliberately scrambled.
  for (k in seq_len(5L)) {
    set.seed(7919L + k * 13L)
    .junk <- runif(100L)
    d <- sim_multisite(preset_jebs_paper(), seed = 1L)
    out[[k]] <- tibble::tibble(
      run = sprintf("rep%02d (seed=1)", k),
      hash = canonical_hash(d),
      seed = 1L
    )
  }
  # Nine more seeds, each once.
  more_seeds <- c(2L, 3L, 4L, 5L, 7L, 11L, 13L, 21L, 42L)
  for (j in seq_along(more_seeds)) {
    s <- more_seeds[[j]]
    d <- sim_multisite(preset_jebs_paper(), seed = s)
    out[[5L + j]] <- tibble::tibble(
      run = sprintf("seed=%d", s),
      hash = canonical_hash(d),
      seed = s
    )
  }
  dplyr::bind_rows(out)
}

hash_tbl <- hash_grid()

ggplot(hash_tbl, ggplot2::aes(x = run, y = 1, fill = hash)) +
  ggplot2::geom_col(width = 0.85) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_viridis_d(option = "C", guide = "none") +
  ggplot2::labs(
    title = "Canonical hash is stable under reseed and changes with seed",
    subtitle = "preset_jebs_paper() · ambient global state varies across rep01-rep05",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank()
  )
```

Two things to read off the plot. First, the five `rep01`-`rep05`
bars (top of the panel) all have the same fill colour: same hash,
even though the ambient global RNG was scrambled before each call.
This is the local-seed-stream policy at work — the only thing that
matters for `canonical_hash(dat)` is the value passed to the `seed`
argument and the design ingredients. Second, the nine seed-varied
bars (bottom of the panel) each have their own fill colour: a
different seed gives a different hash, as expected.

## 5. The reproducibility contract

The package guarantees a precise statement, which is worth writing
out as a contract. Given:

- the **same package version** of `multisiteDGP` (the `0.1.0` line
  in the provenance string),
- the **same seed** passed as the `seed` argument to a primary call
  (`sim_multisite()`, `sim_meta()`, `gen_effects()`,
  `gen_observations()`),
- the **same design** (the same preset call, or the same
  `multisitedgp_design()` arguments, or the same
  `update_multisitedgp_design()` patch),

the package guarantees:

- the **same `canonical_hash(dat)`** byte-for-byte,
- and therefore the same realized point estimates, the same
  realized standard errors, the same diagnostics in the canonical
  allowlist, and the same `provenance_string(dat)` modulo the R
  version slot.

That is the contract. The hash is the witness. If the contract
holds, two researchers running the same line of code on different
machines see the same simulation, full stop. The smoke check used
throughout the package documentation —
`sim_multisite(preset_jebs_paper(), seed = 1L)` produces hash
`c52e75f276d82836` — is a single instance of the contract. If your
local rerun matches that hash, the package install you have is in
sync with the published baseline.

The contract is intentionally narrow. It is *not* a guarantee that:

- your post-simulation analysis pipeline (a Bayesian model fit, a
  bootstrap, a robustness sweep) is reproducible — that is your
  pipeline's responsibility, and a separate hash on the downstream
  artifact is the right tool there;
- a different platform or a different R minor version produces the
  same hash — see Section 6;
- a custom `g_fn` callback gives the same hash if you change its
  body — the package records callback *presence*, not body
  identity, by design (changing the body is a meaningful change to
  the simulation that the user should re-pin manually).

The narrowness is deliberate. A wider contract would be either
unenforceable (downstream pipelines vary too much) or false
(cross-platform byte-equality is not a property the package can
deliver — see below).

## 6. What can break the contract

There are three classes of contract break, in roughly decreasing
order of frequency.

### 6.1 Changing the package version

This is by far the most common. `multisiteDGP 0.1.x → 0.2.x` is a
minor bump that may add fields to the canonical-hash payload, change
the order in which diagnostics are computed, or alter a generator's
internal sampling scheme. Any of these legitimately changes the
hash. The mitigation is the manifest: when you save
`pkg_version = "0.1.0"` alongside the hash, a future you whose
`packageVersion("multisiteDGP")` returns `0.2.0` knows immediately
why the hashes differ, and can decide whether to install the older
version (via `remotes::install_version()` or `renv`) or to re-pin
the artifact at the new version. Either choice is principled; what
is not principled is a silent mismatch.

### 6.2 Changing the R version

R version drift is rarer in practice but does occur. R 4.4 →
R 4.5 changed the default RNG kind for `sample.int()` in some
edge cases (`RNGkind("Rounding")` legacy behavior), and a major-
version bump (R 4.x → R 5.x) is a near-certain hash break. The
package emits a warning when a saved provenance string was generated
under a different R version than the current session.

The provenance string carries the R version as the `R=4.5` slot
specifically so this is grep-able. If you see two provenance lines
that differ only in the `R=...` slot, you have located the cause.

### 6.3 Changing the OS or the platform

Cross-platform hash equality is held to a weaker standard. Linux
x86_64 / amd64 is the strict baseline used for the package's golden
fixtures in `tests/testthat/`. macOS and Windows are held to
*same-machine* reproducibility plus *distributional parity* against
the Linux baseline — the realized distribution of point estimates
and SEs matches Linux to numerical tolerance, but the exact
`canonical_hash` is allowed to drift because of platform-level
floating-point differences in BLAS / LAPACK and compiler math
libraries. The full installed policy is in
`system.file("REPRODUCIBILITY.md", package = "multisiteDGP")`. If
you and a coauthor on different operating systems produce different
hashes for the same `seed` and design, the next check is whether
the *distributions* of `tau_j_hat` and `se_j` agree to numerical
tolerance, not the 16 hex characters.

### 6.4 What does not break the contract

Some things look like they should change the hash but do not. The
ambient global RNG state at the moment of the call does not (see
Section 4). Reordering tibble columns does not (canonicalization
sorts columns). Adding a non-numeric diagnostic does not (the
allowlist is numeric-only). The current working directory and the
system locale do not. Knowing what is *not* in the payload is part
of trusting the hash.

## 7. The IES Form 1 acknowledgments paragraph

The package is distributed with an acknowledgments paragraph
(reproduced verbatim in the footer of every vignette) that names the
funder, grant number, and standard disclaimer. M7 is the dedicated
home in vignette space for explaining *why* the paragraph appears
where it does.

The paragraph identifies the source of support: the Institute of
Education Sciences, U.S. Department of Education, through Grant
R305D240078 to the University of Alabama, with the standard
disclaimer that the views expressed are the authors' and do not
represent the Institute's. Every vignette reproduces the paragraph
in its footer (uniform appearance, low cognitive load for the
reader). M7 is the place where the *role* of the paragraph is
discussed — funder acknowledgment is a condition of the grant and a
research-integrity norm for federally funded methodological work,
and the precise wording is fixed (Form 1) so it can be machine-
checked across the package's documentation surface in the release
audit that greps for the grant number.

When you cite this package in a paper appendix, the courtesy is
either to reproduce the acknowledgment paragraph verbatim or to
cite the package's `CITATION` entry and the underlying methods paper
[@lee2025improving]. The IES grant terms apply only to outputs the
IES project itself produces.

## 8. Where to next

You now have, for any simulation you run with this package: a
one-line provenance string, a 16-character canonical hash, a
verification recipe, and a manifest pattern. The remaining vignettes
build on this base.

- [Getting started](a1-getting-started.html) — A1 §9 is the
  user-track entry point to provenance, with a lighter treatment of
  the same workflow.
- [Cookbook](a8-cookbook.html) — recipe G walks an end-to-end
  manifest workflow inside a `design_grid()` power study.
- [Migration from siteBayes2](m8-migration-from-siteBayes2.html) —
  re-pinning legacy siteBayes2 artifacts under the same audit trail.
- [`canonical_hash()`](../reference/canonical_hash.html) — argument
  reference, including the `algo` and `columns_to_include` arguments
  not used here.
- [`provenance_string()`](../reference/provenance_string.html) — the
  S3 generic and the three methods (data, design, default).

For the methodological foundations the hash is fingerprinting,
start with [The two-stage DGP](m1-statistical-dgp.html) and proceed
through M2 / M3 / M4 in order. For the empirical-Bayes context the
package's hash-pinned simulations are typically deployed in, see
@walters2024empirical.

## References

The `multisiteDGP` data-generating process and the JEBS preset
implement the design in @lee2025improving. Empirical-Bayes context
for the simulation studies the package is typically deployed in is
collected in @walters2024empirical.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
