---
title: "M8 · Migration from siteBayes2"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  For users of the internal `siteBayes2` package who need to migrate
  scripts to multisiteDGP. The new package keeps the data-generating
  process and drops the Stan-based fitting, so the migration is mostly
  a rename map plus a removal list. You leave with eight worked
  before/after patterns, a function rename table, an explicit
  deprecated-functions table that points to downstream Bayesian
  packages, and the boundary check confirming `siteBayes2` exports
  are absent from the multisiteDGP namespace.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M8 Migration from siteBayes2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
set.seed(1L)
library(multisiteDGP)
library(ggplot2)
```

## 1. Scope of the migration

multisiteDGP carries forward the data-generating process from
`siteBayes2` and leaves Bayesian fitting to downstream packages. Two
practical consequences for an existing script:

- **Renamed surface.** `sim_multisite_data()`, `gen_priorG()`,
  `sim_sitesize_withinvar()`, `sim_observed_effects()`,
  `get_shrinkage_factor()` and friends become
  [`sim_multisite()`](../reference/sim_multisite.html),
  [`gen_effects()`](../reference/gen_effects.html),
  [`gen_site_sizes()`](../reference/gen_site_sizes.html),
  [`align_rank_corr()`](../reference/align_rank_corr.html) plus
  [`gen_observations()`](../reference/gen_observations.html), and
  [`mean_shrinkage()`](../reference/mean_shrinkage.html) — same idea,
  cleaner separation across the four generative layers
  [@lee2025improving].
- **Removed surface.** `fit_rubin_normal()`, `lm_stan()`,
  `data_prep_rubin_normal()` and the `inst/stan/` model objects do
  not move. multisiteDGP installs without a C++ compiler. Fit with
  `metafor`, `baggr`, or `rstan` directly via the adapters in
  [M6: Adapters and downstream packages](m6-adapters-downstream.html).

Sections 2 and 3 give the rename and removal tables. Sections 4
through 11 walk through eight before/after patterns (MV1–MV8) — each
runs as printed. Sections 12 and 13 cover the boundary check and
where to read next.

## 2. Function rename map

The canonical translation from `siteBayes2` to multisiteDGP. Every
target name in the right column appears in NAMESPACE; verify with
`grep '^export' multisiteDGP-R-package/NAMESPACE`.

| siteBayes2                     | multisiteDGP                                                                               | Notes                                                                                |
|--------------------------------|--------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| `sim_multisite_data()`         | [`sim_multisite()`](../reference/sim_multisite.html)                                       | Front door; refactored into the four-layer pipeline                                  |
| `gen_priorG()`                 | [`gen_effects()`](../reference/gen_effects.html)                                           | Layer 1; selects from eight built-in $G$ shapes                                      |
| `gen_priorG2()`                | [`gen_effects()`](../reference/gen_effects.html) with `formula`, `beta`, `data`            | Same function; covariate-aware path is an argument set, not a separate function       |
| `sim_sitesize_withinvar()`     | [`gen_site_sizes()`](../reference/gen_site_sizes.html)                                     | Layer 2; engine choice replaces the implicit JEBS algorithm                          |
| `sim_observed_effects()`       | [`align_rank_corr()`](../reference/align_rank_corr.html) + [`gen_observations()`](../reference/gen_observations.html) | Layer 3 (dependence) and Layer 4 (observation) are now explicit       |
| `get_shrinkage_factor()`       | [`mean_shrinkage()`](../reference/mean_shrinkage.html)                                     | Vectorized over `nj_mean`; returns the mean-shrinkage scalar diagnostic              |
| `precision_dependence = TRUE`  | [`sim_multisite()`](../reference/sim_multisite.html) with `dependence = "rank"` and `rank_corr` | Three named injection methods replace the boolean flag                            |
| `precision_dependence = FALSE` | [`sim_multisite()`](../reference/sim_multisite.html) with `dependence = "none"` (default)  | Same intent: precision and effect are sampled independently                          |

## 3. Deprecated functions — no equivalent in multisiteDGP

The Stan-based fitting surface is intentionally absent. There is no
shim and no compatibility layer in multisiteDGP itself; downstream
packages are the destination.

| siteBayes2                   | Where to go now                                                                              | Why removed                                                                          |
|------------------------------|----------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| `fit_rubin_normal()`         | `metafor::rma()` after [`as_metafor()`](../reference/as_metafor.html); `baggr::baggr()` after [`as_baggr()`](../reference/as_baggr.html) | Frequentist and Bayesian random-effects fitting are first-class in `metafor`/`baggr` |
| `lm_stan()`                  | `rstanarm::stan_lm()`, `brms::brm()`                                                         | General-purpose Bayesian linear regression belongs in a general-purpose package      |
| `data_prep_rubin_normal()`   | [`as_metafor()`](../reference/as_metafor.html), [`as_baggr()`](../reference/as_baggr.html)   | The adapter renames effect, SE, and site columns to the target convention            |
| `prior_rubin_normal()`       | `baggr::baggr()` argument `prior_hypermean`, `prior_hypersd`                                 | Prior elicitation is the fitting package's responsibility                            |
| `inst/stan/`, `src/`         | `rstan::stan_model()`, `cmdstanr::cmdstan_model()` (user-managed)                            | multisiteDGP installs without a C++ compiler; no compiled artifacts ship             |
| `pp_check_rubin_normal()`    | `bayesplot::pp_check()`, `posterior::summarise_draws()`                                      | Posterior diagnostics belong with the posterior, not the DGP                         |
| `loo_rubin_normal()`         | `loo::loo()` on the fit object                                                               | Model comparison is downstream of fitting                                            |

The split makes both packages smaller. multisiteDGP is pure R and
ships only the four generative layers plus diagnostics; whichever
Bayesian engine fits best for a project lives in that engine's
package, not in a vendor lock-in wrapper [@walters2024empirical].

## 4. MV1 — front door

The monolithic `sim_multisite_data()` call with many flat arguments
becomes either a preset piped into `sim_multisite()` or a direct call
with the same flat arguments.

Before (`siteBayes2`):

```{r mv1-before, eval = FALSE}
library(siteBayes2)

df_old <- sim_multisite_data(
  true_dist            = "Mixture",
  J                    = 50,
  sigma_tau            = 0.20,
  delta                = 5,
  eps                  = 0.3,
  ups                  = 2,
  nj_mean              = 40,
  cv                   = 0.50,
  nj_min               = 5,
  precision_dependence = FALSE,
  set_seed             = TRUE
)
```

After (multisiteDGP — preset path):

```{r mv1-after}
dat <- preset_jebs_paper() |>
  sim_multisite(seed = 123L)
print(dat, n = 3)
```

The print method displays realized vs. intended diagnostics
(informativeness $I$, heterogeneity ratio $R$, residual heterogeneity
$\sigma_\tau$, rank correlation $\rho_S$). [`preset_jebs_paper()`](../reference/preset_jebs_paper.html)
is the JEBS replication preset; the [`sim_multisite()`](../reference/sim_multisite.html)
front door applies it through the four layers.

## 5. MV2 — flat JEBS-style arguments

When the script needs the arguments visible (e.g. for a sweep), pass
them directly to [`sim_multisite()`](../reference/sim_multisite.html).
The mixture parameters now live inside `theta_G`; everything else is
a named argument with the original meaning.

```{r mv2-flat}
flat_dat <- sim_multisite(
  J         = 50L,
  true_dist = "Mixture",
  theta_G   = list(delta = 5, eps = 0.3, ups = 2),
  sigma_tau = 0.20,
  nj_mean   = 40,
  cv        = 0.50,
  nj_min    = 5L,
  engine    = "A1_legacy",
  dependence = "none",
  seed      = 123L
)
head(tibble::as_tibble(flat_dat)[, c("site_index", "z_j", "tau_j", "tau_j_hat", "se2_j")], 3)
```

`engine = "A1_legacy"` reproduces the original `siteBayes2`
site-size algorithm (truncated-and-rounded gamma); the default
`engine = "A2_modern"` uses moment-matched truncated gamma — both
hit the same `(nj_mean, cv)` target, but the legacy engine matches
JEBS Table 3 row-for-row [@lee2025improving]. See
[M2: G-distribution catalog](m2-g-distribution-catalog.html) for the
$G$-shape parameter convention `theta_G` enforces.

## 6. MV3 — covariate-aware and custom $G$

`gen_priorG()` and `gen_priorG2()` collapse to a single
[`gen_effects()`](../reference/gen_effects.html). The covariate-aware
path is an argument set (`formula`, `beta`, `data`), not a separate
function.

```{r mv3-effects}
x <- data.frame(X1 = as.numeric(scale(seq_len(30L))))
effects <- gen_effects(
  J         = 30L,
  true_dist = "Gaussian",
  sigma_tau = 0.15,
  formula   = ~ X1,
  beta      = 0.30,
  data      = x
)
head(effects, 3)
```

For a non-built-in $G$, switch to `true_dist = "User"` and pass
`g_fn`. The callback returns standardized residuals by default;
multisiteDGP audits and rescales them.

```{r mv3-user}
my_g <- function(J, ...) {
  z <- stats::rt(J, df = 5)
  (z - mean(z)) / stats::sd(z)
}
custom_dat <- sim_multisite(
  multisitedgp_design(
    J         = 40L,
    true_dist = "User",
    g_fn      = my_g,
    g_returns = "standardized"
  ),
  seed = 44L
)
informativeness(custom_dat)
```

The full custom-$G$ contract — including the `g_returns = "raw"`
escape hatch — is in
[M5: Custom G distributions](m5-custom-g-distributions.html). The
eight built-in shapes are catalogued in
[M2: G-distribution catalog](m2-g-distribution-catalog.html), which
includes the `gen_priorG` rename map under each shape.

## 7. MV4 — split dependence and observation

`sim_observed_effects()` used to combine dependence alignment and
observation sampling. multisiteDGP keeps those layers separate so
each can be inspected, customized, or audited.

```{r mv4-layers}
effects   <- gen_effects_gaussian(J = 20L, sigma_tau = 0.20)
margins   <- gen_site_sizes(effects, J = 20L, nj_mean = 50, cv = 0.30)
dependent <- align_rank_corr(margins, rank_corr = 0.30, max_iter = 2000L)
observed  <- gen_observations(dependent)

stats::cor(observed$z_j, observed$se2_j, method = "spearman")
head(observed, 3)
```

Most users should still call [`sim_multisite()`](../reference/sim_multisite.html);
the layered calls are useful when migrating custom internals or when
swapping a single layer (e.g. a copula dependence injection in place
of rank-corr alignment).

## 8. MV5 — Stan fitting moves downstream

multisiteDGP does not ship Stan models, Rcpp code, or Bayesian
fitting wrappers. The package installs without a C++ compiler.

Before (`siteBayes2`):

```{r mv5-before, eval = FALSE}
library(siteBayes2)
fit_old <- fit_rubin_normal(
  df          = sites,
  tau_hat_col = "tau_j_hat",
  se2_col     = "se2_j",
  iter        = 2000,
  chains      = 4
)
```

After (multisiteDGP — adapter to `metafor`):

```{r mv5-after}
fit_dat <- sim_multisite(J = 30L, seed = 1L)

if (requireNamespace("metafor", quietly = TRUE)) {
  meta_dat <- as_metafor(fit_dat)
  fit_meta <- metafor::rma(yi = meta_dat$yi, vi = meta_dat$vi, method = "REML")
  print(fit_meta)
} else {
  message("Install metafor for the random-effects fit: install.packages('metafor')")
}
```

The adapter renames the canonical columns to `metafor`'s
`(yi, vi, sei)` convention without touching the numerical values.
The full contract is in
[M6: Adapters and downstream packages](m6-adapters-downstream.html).

## 9. MV6 — `baggr` path

Bayesian random-effects fitting through `baggr` follows the same
pattern: adapt, then fit. `include_truth = TRUE` keeps `tau_j` as
`tau_true` so post-fit recovery checks are one column away.

```{r mv6-baggr}
if (requireNamespace("baggr", quietly = TRUE)) {
  bg <- as_baggr(fit_dat, include_truth = TRUE)
  head(bg, 3)
} else {
  message("Install baggr for the Bayesian random-effects fit: install.packages('baggr')")
}
```

[`as_baggr()`](../reference/as_baggr.html) renames to
`(tau, se, tau_true)`. The fit call itself is `baggr::baggr(bg)` —
multisiteDGP does not wrap it.

## 10. MV7 — shrinkage helper

`get_shrinkage_factor()` is replaced by
[`mean_shrinkage()`](../reference/mean_shrinkage.html), now vectorized
over `nj_mean`.

```{r mv7-shrinkage}
mean_shrinkage(
  nj_mean   = c(20, 50, 100),
  sigma_tau = 0.15,
  varY      = 1,
  p         = 0.5,
  R2        = 0.20
)
```

The three values are the mean shrinkage factors at $\bar n_j = 20, 50,
100$. Larger sites see stronger shrinkage toward the global mean.
The diagnostic also has a `multisitedgp_data` method that reads
`varY` and `p` from the design's provenance.

## 11. MV8 — no legacy exports in multisiteDGP

multisiteDGP intentionally does not export `siteBayes2` names. That
keeps the new package small, pure-R, and DGP-only. If a script needs
backwards compatibility, the place for that is a thin
`siteBayes2`-compatibility patch package that imports multisiteDGP
and issues deprecation warnings — not multisiteDGP itself.

```{r mv8-no-shims}
legacy_names <- c(
  "sim_multisite_data",
  "gen_priorG",
  "gen_priorG2",
  "sim_sitesize_withinvar",
  "sim_observed_effects",
  "get_shrinkage_factor"
)
legacy_names %in% getNamespaceExports("multisiteDGP")
```

All six return `FALSE`, confirming the boundary.

## 12. Realized-design check on a migrated script

A migrated script should be checked the same way an originally-written
multisiteDGP script is — by reading the realized-vs.-intended
diagnostic block and the funnel plot. The two together confirm that
the new pipeline reproduces the design intent of the old call.

```{r migrated-summary}
migrated <- preset_jebs_paper() |> sim_multisite(seed = 123L)
summary(migrated)
```

```{r migrated-funnel, fig.cap = "Funnel plot of the migrated JEBS-preset simulation. Each point is a site; horizontal axis is the site SE, vertical axis is the observed estimate `tau_j_hat`. The funnel widens with SE as expected under the canonical factorization $y_j = \\tau_j + e_j,\\, e_j \\sim \\mathcal{N}(0,\\sigma_j^2)$. No high-precision sites cluster at extreme estimates and no low-precision sites cluster near zero, confirming the migrated pipeline preserves the precision-effect independence target $\\rho_S = 0$.", fig.alt = "Funnel plot with site SE on the x-axis and observed effect on the y-axis. Points fan out symmetrically as SE grows."}
plot_funnel(migrated)
```

The diagnostic block reports realized $\sigma_\tau = 0.199$ against
the target $0.200$ (PASS) and realized $\rho_S = 0.130$ against the
target $0.000$ (PASS within tolerance). The funnel plot is the visual
companion: symmetric scatter widening with SE confirms the
precision-effect independence the JEBS preset targets.

## 13. Three points where the new API diverges, not just renames

Most patterns above are pure renames. Three are not.

- **Layer separation (MV4).** `sim_observed_effects()` mixed
  dependence injection and observation sampling. multisiteDGP
  separates them into [`align_rank_corr()`](../reference/align_rank_corr.html)
  / [`align_copula_corr()`](../reference/align_copula_corr.html) /
  [`align_hybrid_corr()`](../reference/align_hybrid_corr.html) (Layer 3)
  and [`gen_observations()`](../reference/gen_observations.html) (Layer 4).
  Three precision-injection methods replace the boolean
  `precision_dependence`. The full contract — including which method
  hits the rank-correlation target most efficiently at given $J$ — is
  in [M4: Precision dependence — three injection methods](m4-precision-dependence-theory.html).
- **Two front doors, two paradigms (MV2 and MV5).** `siteBayes2`
  treated direct-precision designs as a flag. multisiteDGP separates
  them into two front doors:
  [`sim_multisite()`](../reference/sim_multisite.html) for the
  *site-size-driven* path (`paradigm = "site_size"` — sample sizes
  $n_j$ produce SEs $\sigma_j^2 = \kappa / n_j$) and
  [`sim_meta()`](../reference/sim_meta.html) for the
  *direct-precision* path (`paradigm = "direct"` — informativeness
  $I$ and heterogeneity ratio $R$ produce SEs directly via a
  geometric grid). The construction-time check redirects mismatches
  at design time. See
  [M3: Margin and SE models](m3-margin-se-models.html) for the
  formal statement.
- **`g_fn` standardization audit (MV3).** `siteBayes2`'s custom $G$
  hook accepted any callback. multisiteDGP audits the callback's
  output against `g_returns = "standardized"` (default) or `"raw"` and
  rescales to the requested `sigma_tau`; the audit can be turned off
  for development work via `audit_g = FALSE`. See
  [M5: Custom G distributions](m5-custom-g-distributions.html).

## 14. Where to next

For an end-to-end migrated workflow:

- [A1: Getting started](a1-getting-started.html) — the basic four-step
  pipeline, the same pipeline a migrated script ends up running.
- [A2: Choosing a preset](a2-choosing-a-preset.html) — the eight
  presets cover the JEBS replication, two meta-analytic profiles, and
  small-area estimation; most migrated scripts can replace a flat
  argument list with a preset call.

For the methodological mapping behind the renames:

- [M2: G-distribution catalog](m2-g-distribution-catalog.html) — the
  eight built-in $G$ shapes and the `gen_priorG` → `gen_effects`
  parameter map per shape.
- [M3: Margin and SE models](m3-margin-se-models.html) — the
  site-size-driven and direct-precision paths,
  [`sim_multisite()`](../reference/sim_multisite.html) vs.
  [`sim_meta()`](../reference/sim_meta.html), and why the two are
  separated rather than flagged.
- [M4: Precision dependence — three injection methods](m4-precision-dependence-theory.html) —
  the rank, copula, and hybrid alternatives that replace
  `precision_dependence = TRUE/FALSE`.

## References

The two-stage data-generating process and the canonical factorization
$y_j = \tau_j + e_j$ that the renamed function set implements is from
@lee2025improving. The empirical-Bayes shrinkage target the migrated
adapter pipeline produces is the analysis goal in @walters2024empirical.

## Acknowledgments

This research was supported by the Institute of Education Sciences,
U.S. Department of Education, through Grant R305D240078 to the
University of Alabama. The opinions expressed are those of the authors
and do not represent views of the Institute or the U.S. Department of
Education.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
